---
title: "Kenya Invasive & Prohibited Plant Species"
subtitle: "9 species with GBIF backbone keys and verified legal/database citations"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(kableExtra)
  library(glue)
})

# Load the verified CSV
inv <- read_csv(
  "kenya_invasive_plants_verified.csv",
  col_types = cols(.default = col_character())
)
```

# Overview

This report presents **`r nrow(inv)` invasive and prohibited plant species** in Kenya with GBIF backbone keys and source-verified legal/database citations.

Every species below is backed by at least one publicly accessible government, database, or peer-reviewed source. The `Note` column in the accompanying CSV contains the full source URLs for each species.

## Key verified sources

1. **FAO/KEPHIS Chapter 26** — "Invasive alien species in Kenya: status and management" (Kedera & Kuria, KEPHIS officials). <https://www.fao.org/4/y5968e/y5968e10.htm> — *Table 1 lists Kenya's invasive plant species with year of arrival and impact severity.*
2. **IUCN Global Invasive Species Database (GISD)** — <https://www.iucngisd.org>
3. **CABI Invasive Species Compendium** — <https://www.cabi.org/isc>
4. **EPPO Global Database** — <https://gd.eppo.int>
5. **Kenya Law** — <https://new.kenyalaw.org>

## Legal framework

| Law | What it does | URL |
|:----|:-------------|:----|
| **EMCA 1999, Section 42(d)** | Prohibits alien and indigenous plant introduction into Kenya's freshwater bodies (see exact wording below) | [Kenya Law](https://new.kenyalaw.org/akn/ke/act/1999/8/eng@2022-12-31) |
| **Suppression of Noxious Weeds Act, Cap 325** | The only Kenyan Act that names specific weed species in its Schedule | [Kenya Law PDF](http://kenyalaw.org/kl/fileadmin/pdfdownloads/Acts/SuppressionofNoxiousWeedsAct_Cap325.pdf) |
| **Plant Protection Act, Cap 324** | Framework law giving KEPHIS power to declare pests; names no species itself | [Kenya Law](https://new.kenyalaw.org/akn/ke/act/1937/24/eng@2022-12-31) |
| **Agriculture Act** (farm forestry rules) | Bans Eucalyptus in wetlands and riparian areas | Cited by [Africa Check](https://africacheck.org/fact-checks/blog/analysis-thirsty-species-science-behind-eucalyptus-tree-ban-kenyas-wetlands) |
| **Water Act** | Bans exotic species with adverse effects on water resources | Cited by [Africa Check](https://africacheck.org/fact-checks/blog/analysis-thirsty-species-science-behind-eucalyptus-tree-ban-kenyas-wetlands) |

## EMCA 1999 Section 42(d) — exact wording

Under Kenya's Environmental Management and Co-ordination Act, 1999, **Section 42(d)** states that no person shall, without the prior written approval of the Director-General:

> *"introduce or plant any part of a plant specimen, whether alien or indigenous, dead or alive, in any river, lake or wetland"*

Source: [EMCA 1999 on Kenya Law](https://new.kenyalaw.org/akn/ke/act/1999/8/eng@2022-12-31)

This is the operative statutory provision for all freshwater aquatic plant invasives in Kenya — *not* Section 55, which covers the coastal zone. Section 42(d) does not name specific species; species-level declarations appear in the Schedule to the **Suppression of Noxious Weeds Act, Cap 325** (which names *Eichhornia crassipes* and *Salvinia auriculata* among other declared noxious weeds).


# Verified Species Table (18-Column Format) {#species-table}

The table below shows all `r nrow(inv)` species with their GBIF backbone keys and verified legal citations. Where EMCA s.42(d) applies, the exact statutory wording is cited.

```{r main-table}
# Build clickable GBIF links
display <- inv |>
  mutate(
    `GBIF Key` = if_else(
      !is.na(GBIFusageKey) & GBIFusageKey != "",
      paste0('<a href="https://www.gbif.org/species/', GBIFusageKey,
             '" target="_blank">', GBIFusageKey, '</a>'),
      "N/A"
    ),
    `Scientific Name` = paste0("<em>", prefName, "</em>"),
    `Common Name(s)` = str_trunc(englishName, 80),
    Classification = classification,
    Family = family,
    `Taxon Level` = taxonLevel,
    Aquatic = if_else(is_aquatic == "yes", paste0("Yes (", aquatic_type_I, ")"), "No"),
    `Legal Status (Kenya)` = legal_status_ke,
    `Restriction Type` = restriction_type,
    `Governing Law` = governing_law
  ) |>
  select(
    `GBIF Key`, `Scientific Name`, `Common Name(s)`, Classification,
    Family, `Taxon Level`, Aquatic,
    `Legal Status (Kenya)`, `Restriction Type`, `Governing Law`
  )

display |>
  kbl(escape = FALSE, caption = "Table 1. Kenya Invasive & Prohibited Plant Species (9 species)") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = TRUE, font_size = 11
  ) |>
  column_spec(1, width = "8em") |>
  column_spec(2, width = "12em") |>
  column_spec(8, width = "14em") |>
  column_spec(10, width = "18em") |>
  scroll_box(height = "600px")
```

**Note on Section 42(d):** For aquatic species (*Pontederia crassipes*, *Salvinia molesta*) and any other species that may be introduced to rivers, lakes, or wetlands, EMCA 1999 s.42(d) applies the exact prohibition quoted above: *"introduce or plant any part of a plant specimen, whether alien or indigenous, dead or alive, in any river, lake or wetland"*.


# Per-Species Source Citations {#citations}

```{r per-species-citations, results='asis'}
for (i in seq_len(nrow(inv))) {
  row <- inv[i, ]
  cat(glue("\n\n## {i}. {row$prefName}\n\n"))
  cat(glue("**Common name(s):** {str_trunc(row$englishName, 120)}\n\n"))
  cat(glue("**GBIF usage key:** [{row$GBIFusageKey}](https://www.gbif.org/species/{row$GBIFusageKey})\n\n"))
  cat(glue("**Family:** {row$family} | **Genus:** {row$genus_I} | **Taxon level:** {row$taxonLevel}\n\n"))
  cat(glue("**Aquatic:** {row$is_aquatic}",
           if_else(!is.na(row$aquatic_type_I), glue(" ({row$aquatic_type_I})"), ""),
           "\n\n"))
  cat(glue("**Classification:** {row$classification}\n\n"))
  cat(glue("**Legal status (Kenya):** {row$legal_status_ke}\n\n"))
  cat(glue("**Restriction type:** {row$restriction_type}\n\n"))
  cat(glue("**Governing law:** {row$governing_law}\n\n"))

  # For aquatic species, highlight the s.42(d) statutory quote
  if (!is.na(row$is_aquatic) && row$is_aquatic == "yes") {
    cat("\n**EMCA 1999 s.42(d) — applicable statutory prohibition:**\n\n")
    cat("> *\"introduce or plant any part of a plant specimen, whether alien or indigenous, dead or alive, in any river, lake or wetland\"*\n\n")
  }

  # Parse Note field for source URLs
  note <- row$Note
  if (!is.na(note) && nchar(note) > 0) {
    cat("**Notes & verified sources:**\n\n")
    # Split on " | " to get individual source references
    parts <- str_split(note, " \\| ")[[1]]
    for (p in parts) {
      # Check if this part contains a URL
      urls <- str_extract_all(p, "https?://[^\\s]+")[[1]]
      if (length(urls) > 0) {
        label <- str_remove(p, "https?://[^\\s]+") |> str_squish()
        for (u in urls) {
          cat(glue("- [{label}]({u})\n\n"))
        }
      } else {
        cat(glue("- {p}\n\n"))
      }
    }
  }

  if (!is.na(row$synonyms) && nchar(row$synonyms) > 0) {
    syn_trunc <- str_trunc(row$synonyms, 200)
    cat(glue("\n**Synonyms:** {syn_trunc}\n\n"))
  }

  cat("\n---\n")
}
```


# Summary Statistics

```{r summary-stats}
inv |>
  count(classification, sort = TRUE) |>
  rename(Classification = classification, `Species Count` = n) |>
  kbl(caption = "Table 2. Species by Legal Classification") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

```{r aquatic-summary}
inv |>
  count(is_aquatic, sort = TRUE) |>
  mutate(is_aquatic = if_else(is_aquatic == "yes", "Aquatic", "Terrestrial")) |>
  rename(`Habitat` = is_aquatic, `Species Count` = n) |>
  kbl(caption = "Table 3. Species by Aquatic Status") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

```{r family-breakdown}
inv |>
  count(family, sort = TRUE) |>
  rename(Family = family, `Species Count` = n) |>
  kbl(caption = "Table 4. Species by Plant Family") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```


# CSV Download

The 18-column CSV file is available at:

`kenya_invasive_plants_verified.csv`

**Schema:** `GBIFusageKey, country, region, jurisdiction, jurisdiction_group, prefName, classification, taxonLevel, family, englishName, synonyms, Note, genus_I, is_aquatic, aquatic_type_I, legal_status_ke, restriction_type, governing_law`


# Methodology

1. **GBIF resolution:** 6 species carried over from `kenya_regulated_plants_gbif.csv` with confirmed usage keys. 3 species (*Tagetes minuta*, *Ipomoea* spp., *Eucalyptus* spp.) resolved via `rgbif::name_backbone()`. Vernacular names and synonyms fetched via `rgbif::name_usage()`.

2. **Source verification (16 April 2026):** Each species checked against GISD, CABI Compendium, EPPO Global Database, FAO/KEPHIS Chapter 26, and Kenya Law. Legal citations reference actual species-naming instruments (Cap 325) or exact statutory provisions (EMCA s.42(d)) rather than framework Acts that name no species.

3. **Transparency:** The `Note` column contains full source URLs for every species. Where legal status could not be confirmed against a publicly available document, this is stated explicitly (e.g., "NOT formally declared noxious weed" for *Lantana camara*).

---

*Report generated `r format(Sys.time(), '%d %B %Y %H:%M %Z')` from `build_invasive_csv.R` output.*
