---
title: "Restricted and Protected Plant Species in the Philippines"
subtitle: "Republic Act 9147, DENR DAOs, and CITES — A Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})

.kqb_pkg <- "/Users/kwizera.jvk/Desktop/Kwiz Computing Comes to Life/Kwiz Quants/libs/kwizbranding"
suppressMessages(devtools::load_all(.kqb_pkg, quiet=TRUE))
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"), file.path(getwd(),"www"),
                            file.path(getwd()), file.path(here::here()))
.kqb_load_logo <- function(fn, ck) {
  for (d in .kqb_logo_candidates) { p <- file.path(d,fn)
    if (file.exists(p)) { assign(ck, paste0("data:image/svg+xml;base64,",
      base64enc::base64encode(readBin(p,"raw",file.size(p)))), envir=.GlobalEnv); return(invisible(TRUE)) } }
}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")

styled_kable <- function(df, caption=NULL, scroll=TRUE, height="450px") {
  k <- df |> kbl(caption=caption, escape=FALSE, align="l") |>
    kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"), full_width=TRUE, font_size=12)
  if (scroll) k <- scroll_box(k, height=height); k
}

# ── Embedded species data ─────────────────────────────────────────────────────
# Sources: RA 9147 (2001); DENR DAO 2019-09 (updated threatened species list);
#          DENR DAO 2004-15 (prior list for reference); CITES CoP19 2022.
species <- tribble(
  ~common_name,                     ~scientific_name,              ~taxonomic_group, ~iucn_status,            ~cites_appendix, ~legal_status_ph,                   ~restriction_type,             ~governing_law,                       ~managing_institution,   ~notes,
  "Philippine teak",                "Tectona philippinensis",      "Magnoliopsida",  "Endangered",            "Not listed",    "Critically Endangered (DAO 2019-09)","Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Philippine endemic; Batangas and Mindoro; <500 mature trees",
  "Waling-waling orchid",           "Vanda sanderiana",            "Magnoliopsida",  "Endangered",            "Appendix II",   "Critically Endangered (DAO 2019-09)","Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Queen of Philippine orchids; Mindanao; heavily collected",
  "Jade vine",                      "Strongylodon macrobotrys",    "Magnoliopsida",  "Vulnerable",            "Not listed",    "Vulnerable (DAO 2019-09)",           "Collection/trade regulated",  "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Luzon endemic; spectacular turquoise flower; deforestation threat",
  "Philippine rafflesia",           "Rafflesia manillana",         "Magnoliopsida",  "Critically Endangered", "Not listed",    "Critically Endangered (DAO 2019-09)","Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Parasitic plant; Luzon; habitat loss primary threat",
  "Rafflesia",                      "Rafflesia speciosa",          "Magnoliopsida",  "Critically Endangered", "Not listed",    "Critically Endangered (DAO 2019-09)","Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Panay endemic; world's largest flowers; critically threatened",
  "Pitogo palm",                    "Cycas edentata",              "Cycadopsida",    "Endangered",            "Appendix I",    "Endangered (DAO 2019-09)",           "Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Philippine cycad; all Cycas spp. CITES listed",
  "Philippine slipper orchid",      "Paphiopedilum adductum",      "Magnoliopsida",  "Critically Endangered", "Appendix I",    "Critically Endangered (DAO 2019-09)","Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Mindanao endemic slipper orchid; all Paphiopedilum Appendix I",
  "Mindanao slipper orchid",        "Paphiopedilum urbanianum",    "Magnoliopsida",  "Endangered",            "Appendix I",    "Endangered (DAO 2019-09)",           "Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Philippine endemic; Mindanao slipper orchid",
  "Dipterocarp (apitong)",          "Dipterocarpus grandiflorus",  "Magnoliopsida",  "Vulnerable",            "Not listed",    "Vulnerable (DAO 2019-09)",           "Regulated harvest",           "RA 9147 / Forestry Code",            "DENR-FMB",              "Dominant lowland forest tree; heavy commercial logging history",
  "Lauan / red meranti",            "Shorea polysperma",           "Magnoliopsida",  "Vulnerable",            "Not listed",    "Vulnerable (DAO 2019-09)",           "Regulated harvest",           "RA 9147 / Forestry Code",            "DENR-FMB",              "Major timber species; rapid decline due to selective logging",
  "Narra (Philippine mahogany)",    "Pterocarpus indicus",         "Magnoliopsida",  "Vulnerable",            "Appendix II",   "Protected (National Tree)",          "Felling prohibited",          "RA 9147 / PD 652",                   "DENR-FMB",              "Philippine national tree; strictly protected from cutting",
  "Giant pitcher plant",            "Nepenthes truncata",          "Magnoliopsida",  "Vulnerable",            "Appendix I",    "Vulnerable (DAO 2019-09)",           "Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Mindanao highland pitcher plant; horticultural trade threat",
  "Attenborough's pitcher plant",   "Nepenthes attenboroughii",    "Magnoliopsida",  "Critically Endangered", "Appendix I",    "Critically Endangered (DAO 2019-09)","Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Palawan Mt Victoria; discovered 2007; immediate poaching post-discovery",
  "Palawan pitcher plant",          "Nepenthes palawanensis",      "Magnoliopsida",  "Vulnerable",            "Appendix I",    "Vulnerable (DAO 2019-09)",           "Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-BMB",              "Palawan endemic; highland carnivorous plant",
  "Almaciga / Manila copal",        "Agathis philippinensis",      "Pinopsida",      "Endangered",            "Not listed",    "Endangered (DAO 2019-09)",           "Collection/trade prohibited", "RA 9147 / DAO 2019-09",              "DENR-FMB",              "Source of Manila copal resin; old-growth forests Luzon-Mindanao",
  "Ifit / Toog",                    "Petersianthus quadrialatus",  "Magnoliopsida",  "Endangered",            "Not listed",    "Endangered (DAO 2019-09)",           "Regulated harvest",           "RA 9147 / Forestry Code",            "DENR-FMB",              "Major timber dipterocarp; over-exploited in lowland forests",
  "Water hyacinth",                 "Eichhornia crassipes",        "Magnoliopsida",  "Least Concern",          "Not listed",    "Priority Invasive Species",          "Control/removal mandated",    "RA 9147 / IACSA programme",          "DENR / BFAR",           "Major freshwater invasive; Lake Lanao and Laguna de Bay",
  "Golden kiwi vine / Asian knotweed","Fallopia japonica",         "Magnoliopsida",  "Least Concern",          "Not listed",    "Invasive Species",                   "Prohibited transport/release","EO 23 / IACSA programme",            "DENR-BMB",              "Listed on IUCN 100 Worst Invasives; expanding in Philippines",
  "Giant ipil-ipil",                "Leucaena leucocephala",       "Magnoliopsida",  "Least Concern",          "Not listed",    "Regulated Invasive Species",         "Restricted planting areas",   "RA 9147 / DA regulations",           "DENR / DA",             "Transformative invasive in secondary forest and agricultural areas",
  "Cogon grass",                    "Imperata cylindrica",         "Magnoliopsida",  "Least Concern",          "Not listed",    "Noxious Invasive Grass",             "Control programmes in uplands","RA 9147 / Upland Dev. Programme",    "DENR / DA / BSWM",      "Covers over 1M ha in Philippines; fire-adapted; replaces native forest"
)

n_species  <- nrow(species)
n_dao      <- sum(str_detect(species$governing_law, "DAO"), na.rm=TRUE)
n_invasive <- sum(str_detect(species$restriction_type, regex("invasive|control|prohibited|removal", ignore_case=TRUE)), na.rm=TRUE)
n_cites_i  <- sum(species$cites_appendix == "Appendix I",  na.rm=TRUE)
n_cites_ii <- sum(species$cites_appendix == "Appendix II", na.rm=TRUE)
n_cr       <- sum(species$iucn_status == "Critically Endangered", na.rm=TRUE)
n_en       <- sum(species$iucn_status == "Endangered", na.rm=TRUE)

proj_root  <- here::here()
gbif_cache <- file.path(proj_root, "output", "tables", "philippines_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  clean_name <- function(nm) nm |> str_remove_all("\\s+spp\\.?$") |> str_remove_all("\\(.*?\\)") |> str_squish()
  names_df <- tibble(name=clean_name(species$scientific_name), kingdom="Plantae")
  backbone_res <- tryCatch(name_backbone_checklist(names_df), error=function(e)
    map_dfr(names_df$name, function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),
      error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  get_vern <- function(key) { if(is.na(key)) return(NA_character_)
    tryCatch({ vn <- name_usage(key=as.numeric(key),data="vernacularNames",limit=100)
      if(is.null(vn$data)||nrow(vn$data)==0) return(NA_character_)
      en <- vn$data |> filter(language %in% c("eng","en")) |> pull(vernacularName) |> unique()
      if(length(en)==0) NA_character_ else paste(en,collapse=", ") }, error=function(e) NA_character_) }
  get_syn <- function(key) { if(is.na(key)) return(NA_character_)
    tryCatch({ syn <- name_usage(key=as.numeric(key),data="synonyms",limit=100)
      if(is.null(syn$data)||nrow(syn$data)==0) return(NA_character_)
      s <- syn$data |> pull(canonicalName) |> unique() |> na.omit()
      if(length(s)==0) NA_character_ else paste(s,collapse=", ") }, error=function(e) NA_character_) }
  bc <- backbone_res |> as_tibble() |>
    mutate(GBIFusageKey=as.character(usageKey), prefName=coalesce(species,canonicalName,verbatim_name),
           gbif_family=family, gbif_genus=genus)
  sm <- species |> mutate(clean_sci=clean_name(scientific_name)) |>
    bind_cols(bc |> select(GBIFusageKey,prefName,gbif_family,gbif_genus,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm |> filter(!is.na(GBIFusageKey)) |> distinct(GBIFusageKey) |> pull(GBIFusageKey)
  vl <- list(); sl <- list()
  for (i in seq_along(keys)) { k <- keys[i]; vl[[k]] <- get_vern(k); sl[[k]] <- get_syn(k); if(i%%5==0) Sys.sleep(0.2) }
  vd <- tibble(GBIFusageKey=names(vl), gbif_english=unlist(vl), gbif_synonyms=unlist(sl[names(vl)]))
  enriched <- sm |> left_join(vd,by="GBIFusageKey") |>
    mutate(englishName=coalesce(gbif_english,common_name), synonyms=gbif_synonyms, family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  enriched |> select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_ph,restriction_type,governing_law) |>
    write_csv(gbif_cache)
}
if (file.exists(gbif_cache)) {
  plants_gbif     <- read_csv(gbif_cache, col_types=cols(.default=col_character()), show_col_types=FALSE)
  n_gbif_resolved <- sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="")
  n_gbif_families <- n_distinct(plants_gbif$family, na.rm=TRUE)
} else { plants_gbif <- NULL; n_gbif_resolved <- 0L; n_gbif_families <- 0L }
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04, size="280px auto"))
uri <- kqb_watermark_uri()
if (nzchar(uri)) cat(paste0('<style>\nbody::after {\n  content: "";\n  position: fixed;\n',
  '  top:0; left:0; width:100%; height:100%;\n  z-index:9000;\n  pointer-events:none;\n',
  '  background-image: url("', uri, '");\n  background-repeat:repeat;\n  background-size:280px auto;\n',
  '  opacity:0.04;\n}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

The Philippines is an **archipelagic megadiverse** country and a global biodiversity hotspot, encompassing approximately **9,250 vascular plant species** with an endemism rate exceeding **60%**. Its forests harbour some of the world's most ecologically unique plants — from the giant *Rafflesia* flowers to an extraordinary diversity of pitcher plants, orchids, and dipterocarps. Despite this biological richness, the Philippines has lost over **90% of its original forest cover** and faces ongoing threats from habitat conversion, commercial logging, and international plant trade.

This report compiles **`r n_species` representative plant species** under restriction in the Philippines. **`r n_dao`** are listed in the DENR Administrative Order 2019-09 Threatened Species List under Republic Act 9147. **`r n_invasive`** are designated invasive or noxious species. **`r n_cites_i`** carry CITES Appendix I status and **`r n_cites_ii`** are on CITES Appendix II. **`r n_cr`** are Critically Endangered and **`r n_en`** are Endangered on the IUCN Red List.

---

# Legal Framework {#legal-framework}

## Republic Act No. 9147 — Wildlife Resources Conservation and Protection Act (2001) {#ra9147}

**Republic Act No. 9147** (enacted 30 July 2001) is the primary legislation governing wildlife including plants in the Philippines. It replaced a patchwork of earlier laws and established:

- **Section 5 — Collection of Wildlife:** No person may collect, hunt, or possess wildlife (including plants) designated as threatened, without a Wildlife Collector's Permit from DENR.
- **Section 7 — Trade of Wildlife:** Trade in threatened species requires a Wildlife Farm Permit and compliance with CITES.
- **Section 27 — Prohibited Acts:** Collecting, hunting, killing, or transporting threatened species without permits. Penalties: up to **6 years imprisonment and ₱1,000,000** fine.
- **Threatened Species List:** Operationalised through DENR Administrative Orders (DAOs) that list species by category (Critically Endangered, Endangered, Vulnerable, Other Threatened Species, Other Wildlife Species).

**Current Threatened Species List:** [DENR DAO 2019-09](https://biodiversity.gov.ph/images/Biodiversity/Legal/DAO2019-09.pdf) — updated list covering plants and animals.

Full text: [Official Gazette — RA 9147](https://www.officialgazette.gov.ph/2001/07/30/republic-act-no-9417/)

<div class="alert alert-warning" role="alert">
**⚠ Implementation Gap** — DAO 2019-09 is the current operative list. However, the IUCN Red List of Philippine flora continues to document new Critically Endangered species faster than DAO amendments can respond. Over 400 Philippine plant species are IUCN-assessed as threatened, but DAO 2019-09 lists a fraction of these. DENR-BMB faces severe resource constraints in conducting new wildlife assessments to update the DAO list.
</div>

## Executive Order No. 23 (2011) — Moratorium on Logging {#eo23}

**Executive Order No. 23** declared a moratorium on the cutting and harvesting of timber in natural and residual forests nationwide, complementing the existing forest cover protections.

## DENR DAO 2019-09 — Threatened Species List {#dao}

The current implementing rules list Philippine threatened species by category (CE, E, VU, OTS, OWS). For plants, approximately 400+ species are included across these categories. The DAO provides the operational basis for collecting and trade permits under RA 9147.

## CITES in the Philippines {#cites}

The Philippines ratified CITES in 1981. DENR-BMB is the Management Authority. All orchids (Appendix II; *Paphiopedilum* spp. Appendix I), all cycads (Appendix I/II), and all *Nepenthes* (Appendix I/II) are listed. *Narra* (*Pterocarpus indicus*) is on CITES Appendix II.

<div class="alert alert-warning" role="alert">
**⚠ Implementation Gap** — The Philippines is a major source country for illegal orchid and *Nepenthes* collection. Despite CITES and DAO protections, international social media platforms (particularly Facebook groups) are documented to facilitate illegal wild-collected pitcher plant sales, with collection occurring from protected areas including Mt Hamiguitan and Palawan.
</div>

## Invasive Alien Species Programme {#ias}

The Philippine Invasive Alien Species programme is coordinated through the **Inter-Agency Committee on Invasive Alien Species** (IACSA). The programme operates under RA 9147 and multiple sectoral laws. A standalone IAS Law has been proposed in Congress but not yet enacted.

---

# Regulatory Institutions {#institutions}

```{r institutions-table}
tribble(
  ~`Abbrev.`, ~`Institution`,                               ~`Ministry/Parent`,         ~`Mandate`,
  "DENR-BMB", "Biodiversity Management Bureau",             "DENR",                     "Primary biodiversity regulator; RA 9147 implementation; CITES Management Authority; threatened species listing",
  "DENR-FMB", "Forest Management Bureau",                   "DENR",                     "Forest cover and timber species; Forestry Code; logging permits; forest rehabilitation",
  "PAWB",     "Protected Areas and Wildlife Bureau (legacy)","DENR (now BMB)",           "Former wildlife bureau functions absorbed into BMB in 2016 reorganisation",
  "DA-BFAR",  "Bureau of Fisheries and Aquatic Resources",   "Dept of Agriculture",      "Aquatic plant and seaweed regulation; fisheries; mangrove management",
  "PCAARRD",  "Phil. Council for Agriculture and Fisheries R&D","DOST",                 "Research on invasive species; plant genetic resources conservation",
  "IACSA",    "Inter-Agency Committee on Invasive Alien Species","DENR / DA / DOST",    "Coordinates IAS management across departments; priority IAS list",
  "PNH",      "Philippine National Herbarium",               "DENR-BMB",                "National botanical reference collection; species identification; plant conservation research"
) |>
  styled_kable(caption="Table 1. Institutions with Plant Regulatory Mandates in the Philippines", height="350px")
```

---

# Comprehensive Plant Species Listing {#species}

## Master Species Table

```{r master-table}
species |>
  arrange(iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`, `Group`=taxonomic_group,
         `IUCN`=iucn_status, `CITES`=cites_appendix,
         `Philippines Status`=legal_status_ph,
         `Managing Inst.`=managing_institution, `Notes`=notes) |>
  styled_kable(caption="Table 2. Restricted/Protected Plant Species in the Philippines", height="600px")
```

## Critically Threatened Species {#critically-threatened}

```{r cr-en-table}
species |>
  filter(iucn_status %in% c("Critically Endangered","Endangered")) |>
  arrange(iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`, `IUCN`=iucn_status,
         `CITES`=cites_appendix, `Philippines Status`=legal_status_ph, `Notes`=notes) |>
  styled_kable(caption=glue("Table 3. Critically Endangered and Endangered Plant Species ({n_cr+n_en} species)"), height="450px")
```

## GBIF Backbone Resolution {#gbif}

```{r gbif-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    mutate(
      gbif_url = if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",
                         glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
      iucn_url = paste0("https://www.iucnredlist.org/search?query=",
                        URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
      `Scientific Name` = if_else(!is.na(gbif_url),
        glue("<a href='{gbif_url}' target='_blank'><em>{prefName}</em></a>"),
        glue("<em>{coalesce(prefName,'')}</em>")),
      `IUCN` = glue("<a href='{iucn_url}' target='_blank'>IUCN &#x2197;</a>")
    ) |>
    transmute(`#`=row_number(), `Scientific Name`, `Common Name`=coalesce(englishName,""),
              `Family`=coalesce(family,""), `GBIF Key`=coalesce(GBIFusageKey,""),
              `Philippines Status`=coalesce(legal_status_ph,""),
              `Synonyms`=str_trunc(coalesce(synonyms,""),80), `IUCN`) |>
    styled_kable(caption=glue("Table 4. Philippines Regulated Plants — GBIF Backbone Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} families)"), height="600px")
} else { cat("*GBIF data will be generated on first render with internet access.*") }
```

---

# Regulatory Gaps {#gaps}

```{r gaps-table}
tribble(
  ~`#`, ~`Gap Area`,                                              ~`Severity`, ~`Status (2025)`,
  "1",  "DAO 2019-09 lists fraction of IUCN-threatened plants",   "High",      "Unresolved",
  "2",  "Online wildlife trade (social media) enforcement weak",  "High",      "Partial (DENR-NBI)",
  "3",  "No standalone Invasive Alien Species Act",               "High",      "Bill pending Congress",
  "4",  "DENR-BMB severely under-resourced for enforcement",      "High",      "Unresolved",
  "5",  "RA 9147 permits: slow processing impedes legitimate trade","Medium",  "IT system upgrade ongoing",
  "6",  "Timber species: monitoring outside protected areas weak", "High",     "Unresolved",
  "7",  "Rafflesia/Nepenthes: habitat protection inadequate",     "High",     "Partial (KBAs)",
  "8",  "Climate change adaptation for highland endemics absent",  "Medium",  "Unresolved"
) |>
  kbl(caption="Table 5. Regulatory Gaps for Plant Species in the Philippines", escape=FALSE, align="l") |>
  kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"), full_width=TRUE, font_size=12) |>
  column_spec(3, bold=TRUE, color=ifelse(c(TRUE,TRUE,TRUE,TRUE,FALSE,TRUE,TRUE,FALSE),"white","black"),
              background=ifelse(c(TRUE,TRUE,TRUE,TRUE,FALSE,TRUE,TRUE,FALSE),"#c0392b","#e67e22"))
```

## Online Wildlife Trade

Social media platforms — particularly Facebook groups — have become primary markets for illegally wild-collected orchids, pitcher plants, and other Philippine endemic plants. Despite joint operations between DENR and the National Bureau of Investigation (NBI), the scale of online trading exceeds current enforcement capacity. The TRAFFIC Southeast Asia office has documented the Philippines as a significant source country for illegally traded carnivorous plants and orchids.

## Rafflesia Conservation

*Rafflesia* species — with the world's largest individual flowers — are obligate endoparasites on *Tetrastigma* vines and cannot survive outside their host. They are effectively impossible to cultivate, meaning all trade is wild-collected. The Philippines hosts 13 *Rafflesia* species, more than any other country, but none has been formally bred in captivity. Habitat loss and plant collection continue to shrink their populations.

---

# Key References {#references}

| Resource | Link |
|----------|------|
| RA 9147 — Official Gazette | [officialgazette.gov.ph](https://www.officialgazette.gov.ph/2001/07/30/republic-act-no-9417/) |
| DENR DAO 2019-09 | [biodiversity.gov.ph](https://biodiversity.gov.ph/images/Biodiversity/Legal/DAO2019-09.pdf) |
| DENR-BMB Biodiversity Portal | [biodiversity.gov.ph](https://biodiversity.gov.ph) |
| Philippine GBIF Data | [gbif.org/country/PH](https://www.gbif.org/country/PH/summary) |
| IUCN Red List — Philippines Plants | [iucnredlist.org](https://www.iucnredlist.org) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm <- kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',
  if(nzchar(logo_sm)) paste0(logo_sm,'<br/>') else "",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; For informational purposes only &nbsp;&bull;&nbsp; Verify at DENR-BMB / CITES</p>',
  '</div>'))
```
