---
title: "Restricted and Protected Plant Species in Singapore"
subtitle: "Parks and Trees Act 2005 (Cap. 216), Singapore Red Data Book (3rd Ed.) — Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,            ~taxonomic_group, ~iucn_status,            ~cites_appendix,  ~legal_status_sg,                              ~restriction_type,              ~governing_law,                    ~managing_institution,   ~notes,
  # --- Heritage Trees & Parks and Trees Act 2005 (Cap. 216) ---
  "Tembusu",                 "Fagraea fragrans",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Featured on Singapore $5 note; iconic national tree; multiple Heritage Tree specimens; fine up to SGD 50,000 for damage",
  "Sea apple / Jambu ayer laut","Syzygium grande",         "Magnoliopsida",  "Near Threatened",       "Not listed",     "Heritage Tree / Parks and Trees Act 2005",    "Protected from removal",       "Parks and Trees Act 2005 (Cap.216)","NParks",               "Coastal heritage tree; important in Singapore's streetscape and park canopies; Heritage Tree programme",
  "Rain tree",               "Samanea saman",              "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Most common Heritage Tree species in Singapore; massive canopy spread; multiple specimens on Heritage Tree Register; iconic roadside tree",
  "Banyan tree",             "Ficus benghalensis",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Multiple Heritage Tree specimens including iconic trees at Botanic Gardens and Fort Canning; aerial root system culturally significant",
  "Malayan banyan",          "Ficus microcarpa",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Several Heritage Tree specimens; common in Singapore streetscape; important strangling fig with ecological value",
  "Sea almond",              "Terminalia catappa",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Beachfront heritage trees; distinctive pagoda-shaped canopy; Heritage Tree specimens at East Coast Park and other coastal areas",
  "Saga tree",               "Adenanthera pavonina",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Red saga seeds culturally important in Singapore; Heritage Tree specimens; formerly used as weights by goldsmiths",
  "Kapok tree",              "Ceiba pentandra",            "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Heritage Tree specimens in Singapore; massive buttress roots; one of the tallest Heritage Trees in the register",
  # --- Singapore Red Data Book — Nationally Threatened Species ---
  "Jelutong",                "Dyera costulata",            "Magnoliopsida",  "Vulnerable",            "Not listed",     "Vulnerable (Red Data Book) — protected in reserves","Habitat protection",      "Parks and Trees Act 2005",        "NParks",                "Lowland rainforest tree; latex historically used for chewing gum; VU; Bukit Timah NR populations protected",
  "Singapore walking-stick palm","Livistona saribus",      "Liliopsida",     "Vulnerable",            "Not listed",     "Vulnerable (Red Data Book) — protected in reserves","Collection prohibited",   "Parks and Trees Act 2005",        "NParks",                "Previously Livistona chinensis; native fan palm; Bukit Timah and CCNR populations; Singapore Red Data Book VU",
  "Meranti temak nipis",     "Shorea leprosula",           "Magnoliopsida",  "Endangered",            "Not listed",     "Endangered (Red Data Book) — nature reserve protection","Habitat protection",  "Parks and Trees Act 2005",        "NParks",                "Dipterocarp canopy tree; Bukit Timah NR remnant population; mass flowering events important for forest ecology; EN nationally",
  "Merawan siput jantan",    "Hopea sangal",               "Magnoliopsida",  "Critically Endangered", "Not listed",     "Critically Endangered (Red Data Book) — strict protection","Collection prohibited","Parks and Trees Act 2005",      "NParks",                "CR dipterocarp; fewer than 50 mature trees in Bukit Timah NR; one of Singapore's rarest native trees",
  "Simpoh air",              "Dillenia suffruticosa",      "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected in nature reserves",                "Habitat protection in reserves","Parks and Trees Act 2005",       "NParks",                "Native pioneer shrub; waxy red fruits; ecologically important in secondary forest succession; protected in reserves",
  "Tiup tiup",               "Adinandra dumosa",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected in nature reserves",                "Habitat protection in reserves","Parks and Trees Act 2005",       "NParks",                "Dominant tree of Singapore's adinandra belukar secondary forest; ecologically important habitat type; nature reserve protection",
  "Resam",                   "Dicranopteris linearis",     "Polypodiopsida", "Least Concern",         "Not listed",     "Protected in nature reserves",                "Habitat protection in reserves","Parks and Trees Act 2005",       "NParks",                "Native scrambling fern; dominant in open areas of nature reserves; important ground cover preventing erosion in CCNR",
  "Keruing",                 "Dipterocarpus costulatus",   "Magnoliopsida",  "Critically Endangered", "Not listed",     "Critically Endangered (Red Data Book) — strict protection","Collection prohibited","Parks and Trees Act 2005",      "NParks",                "CR dipterocarp; very few mature individuals remaining in Bukit Timah; winged fruits; globally threatened",
  # --- CITES-listed Species (Endangered Species Import and Export Act 2006) ---
  "Raffles' pitcher plant",  "Nepenthes rafflesiana",      "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (Singapore Red Data Book / CITES App. II)","Collection prohibited",  "Endangered Species (Import and Export) Act / Parks and Trees Act","NParks",  "Named for Stamford Raffles; Singapore's lowland pitcher plant; CITES Appendix II; collection and trade regulated",
  "Slender pitcher plant",   "Nepenthes gracilis",         "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II)",                   "Collection and trade regulated","Endangered Species (Import and Export) Act","NParks",    "Native to Singapore; found in heath forest and forest edges; all Nepenthes CITES App. II; trade requires permits",
  "Singapore orchid",        "Spathoglottis plicata",      "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II) — heritage significance","Collection regulated",   "Endangered Species (Import and Export) Act","NParks",      "Singapore Purple Orchid; all Orchidaceae CITES App. II; important in Singapore's horticultural heritage",
  "Pigeon orchid",           "Dendrobium crumenatum",      "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II)",                   "Collection and trade regulated","Endangered Species (Import and Export) Act","NParks",    "Common epiphytic orchid; mass flowering triggered by temperature drops; all Orchidaceae CITES App. II",
  "Tiger orchid",            "Grammatophyllum speciosum",   "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected (CITES App. II / Red Data Book VU)","Collection prohibited",        "Endangered Species (Import and Export) Act / Parks and Trees Act","NParks / SBGC","World's largest orchid; VU nationally; Heritage specimens in Singapore Botanic Gardens; CITES App. II",
  "Vanda Miss Joaquim",      "Papilionanthe Miss Joaquim", "Magnoliopsida",  "Not Assessed",          "Appendix II",    "National Flower — CITES App. II",             "Collection and trade regulated","Endangered Species (Import and Export) Act","NParks / SBGC","Singapore's national flower since 1981; hybrid orchid; all Orchidaceae CITES App. II; cultivated specimens in Botanic Gardens",
  "Agarwood / Gaharu",       "Aquilaria malaccensis",      "Magnoliopsida",  "Critically Endangered", "Appendix II",    "Protected (CITES App. II) — trade hub enforcement","Import/export permits required","Endangered Species (Import and Export) Act","NParks / AVS","Singapore is a major agarwood trade hub; CITES App. II since 2004; strict permit requirements; CR globally due to overexploitation",
  "Agarwood (secondary)",    "Aquilaria sinensis",         "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected (CITES App. II) — trade regulated", "Import/export permits required","Endangered Species (Import and Export) Act","NParks / AVS","Chinese agarwood; traded through Singapore; all Aquilaria spp. CITES App. II; VU globally",
  "King sago palm",          "Cycas revoluta",             "Cycadopsida",    "Least Concern",         "Appendix II",    "Protected (CITES App. II)",                   "Trade regulated",              "Endangered Species (Import and Export) Act","NParks / AVS","All Cycadaceae CITES App. II; popular ornamental in Singapore; import/export requires CITES permits",
  "Sago palm / Cycad",       "Cycas rumphii",              "Cycadopsida",    "Near Threatened",       "Appendix II",    "Protected (CITES App. II / native species)",  "Collection and trade regulated","Endangered Species (Import and Export) Act / Parks and Trees Act","NParks","Native cycad; coastal habitats; all Cycadaceae CITES App. II; NT globally; Singapore populations in nature reserves",
  "Venus flytrap",           "Dionaea muscipula",          "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected (CITES App. II) — trade regulated", "Import/export permits required","Endangered Species (Import and Export) Act","NParks / AVS","Not native to Singapore but traded; CITES App. II since 2024 (CoP19); Singapore enforces as CITES trade hub",
  # --- Mangrove and Coastal Protected Species ---
  "Nipah palm",              "Nypa fruticans",             "Liliopsida",     "Least Concern",         "Not listed",     "Protected (mangrove habitat protection)",     "Mangrove habitat protected",   "Parks and Trees Act / MPA regulations","NParks",            "Mangrove palm; Sungei Buloh and other coastal areas; Singapore mangrove restoration programme species",
  "Bakau minyak",            "Rhizophora apiculata",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (mangrove habitat protection)",     "Mangrove habitat protected",   "Parks and Trees Act 2005",        "NParks",                "Dominant mangrove species in Singapore; stilt roots; Sungei Buloh Wetland Reserve; protected under mangrove conservation policy",
  "Api-api putih",           "Avicennia alba",             "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (mangrove habitat protection)",     "Mangrove habitat protected",   "Parks and Trees Act 2005",        "NParks",                "Pioneer mangrove species; important in Singapore's mangrove reforestation; Sungei Buloh and Pulau Ubin",
  # --- Invasive Alien Species (NParks Management) ---
  "Mimosa vine",             "Mimosa pigra",               "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — active management",                "Control mandated",             "NParks invasive species programme","NParks",               "Giant sensitive plant; invades freshwater wetlands; Sungei Buloh and Pulau Ubin; outcompetes native vegetation",
  "Mile-a-minute weed",      "Mikania micrantha",          "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — active management",                "Control mandated",             "NParks invasive species programme","NParks",               "Smothers native vegetation in Singapore's secondary forests and forest edges; widespread in southern reserves",
  "Siam weed",               "Chromolaena odorata",        "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — active management",                "Control mandated",             "NParks invasive species programme","NParks",               "Invasive shrub from tropical Americas; outcompetes native pioneer species; NParks active removal in nature reserves and vacant land",
  "African tulip tree",      "Spathodea campanulata",      "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — active management",                "Removal recommended",          "NParks invasive species programme","NParks",               "IUCN 100 worst invasive species; aggressive coloniser of disturbed forest in Singapore; NParks removes from nature reserves",
  "Yellow flame / Albizia",  "Falcataria moluccana",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — active management and removal",    "Removal mandated",             "NParks invasive species programme","NParks",               "Fast-growing tree prone to collapse; NParks actively removes from nature reserves and road corridors; public safety hazard"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","singapore_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_sg,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Singapore retains approximately **2,280 native plant species** on its main island and associated islets — a remarkable number for a city-state of only 733 km². The country has lost over **95% of its original forest cover** since the 19th century, making the remaining patches in Bukit Timah Nature Reserve, the Central Catchment, and the Southern Ridges exceptionally important. Singapore's plant protection regime is anchored in the **Parks and Trees Act 2005 (Cap. 216)**, which empowers NParks to designate **Heritage Trees** and manage nature reserves. Singapore also operates one of the world's most comprehensive **Heritage Tree programmes**, with over **235 designated Heritage Trees** across the island.

<div class="alert alert-info" role="alert">
This report covers **`r n_species` legally restricted plant species** from the Singapore Red Data Book (3rd Edition, NParks 2021). NParks ([nparks.gov.sg](https://www.nparks.gov.sg)) publishes the authoritative species lists and heritage tree register.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Parks and Trees Act (Cap. 216) | 2005 (Rev. Ed.) | Heritage Trees designation; nature reserve management; plant protection; fines up to SGD 50,000 |
| National Parks Board Act (Cap. 198A) | 1996 (Rev. Ed.) | NParks powers; garden city management; ex-situ conservation |
| Endangered Species (Import and Export) Act (Cap. 92A) | 1989 (Rev. Ed.) | CITES implementation; import/export permits for listed species |
| Environmental Protection and Management Act | 1999 (Rev. Ed.) | Environmental quality; pollution control; integrated with NParks operations |
| Singapore Red Data Book (3rd Ed.) | 2021 | Scientific threat assessment for native flora and fauna (NParks publication) |

Full text: [NParks](https://www.nparks.gov.sg) | [Singapore Statutes Online](https://sso.agc.gov.sg) | [FAOLEX Singapore](https://www.fao.org/faolex/country-profiles/general-profile/en/?iso3=SGP)

---

# Subnational / Zonal Restrictions {#regional}

Singapore is a **city-state** with no states, provinces, or other subnational jurisdictions — national legislation applies uniformly across the entire 733 km² territory. There are no separate regional environmental agencies or subnational legislatures. However, functionally significant spatial variation exists through **NParks' designated conservation zones**, **Urban Redevelopment Authority (URA) conservation area plans**, and **park management regulations** that create a de facto zonal restriction system for plant species.

## NParks Conservation Areas — Management Zone System

**National Parks Board (NParks)** manages Singapore's parks, nature reserves, and green corridors under the **National Parks Board Act (Cap. 198A)**. The four **Nature Reserves** designated under the **Parks and Trees Act Cap. 216** have the strongest protections:

| Nature Reserve / Park | Location | Plant Species Restrictions |
|----------------------|----------|--------------------------|
| **Bukit Timah Nature Reserve** | Central Catchment | Primary rainforest remnant; absolute prohibition on plant collection, transplanting, or damage; *Shorea* dipterocarp species, *Hopea sangal*, and endemic *Livistona singapurensis* palm protected under NR designation; research access by NParks permit only |
| **Central Catchment Nature Reserve (CCNR)** | Central Singapore | Singapore's largest forest reserve (3,043 ha); *Livistona singapurensis*, *Hopea*, and *Shorea* species; no collection; buffer zones with additional permit requirements |
| **Labrador Nature Reserve** | Southern Ridge | Coastal cliff forest; rare coastal plants; collection prohibited; access controlled by NParks |
| **Sungei Buloh Wetland Reserve** | Northern coast | Mangroves and coastal wetland plants; Ramsar site; all plant collection prohibited; mangrove species research requires NParks permit |

## Parks and Trees Act — Street Tree and Heritage Tree Register

Beyond nature reserves, the **Parks and Trees Act** regulates:

1. **Streetside/roadside trees**: All trees along public roads are under NParks management; removal or pruning requires NParks approval regardless of the tree's protected status.

2. **Heritage Tree Register**: NParks maintains a Heritage Tree Register of trees with high cultural, historical, or ecological significance. As of 2024, **261 Heritage Trees** are listed, spanning 53 species including:
   - *Ficus benghalensis* (Banyan) — multiple specimens
   - *Terminalia catappa* (Sea almond) — beachfront heritage trees
   - *Tembusu* (*Fagraea fragrans*) — iconic Singapore tree on the $5 note
   - **Penalties for damaging Heritage Trees**: Up to SGD $50,000 fine under the Parks and Trees Act.

3. **Vacant Land (weed/invasive) regulations**: NEA and NParks jointly enforce clearance of invasive plants on vacant land; *Mimosa pigra*, *Mikania micrantha*, and *Chromolaena odorata* control is mandated.

## Urban Redevelopment Authority (URA) — Conservation Area Green Rules

URA's **Master Plan** designates conservation areas (e.g., Chinatown, Kampong Glam, Little India, Emerald Hill) where removal of existing mature trees on conserved buildings requires a **Conservation Technical Condition (CTC)** from URA — a layer above NParks requirements. This effectively creates additional tree protection in heritage districts.

<div class="alert alert-info" role="alert">
**Note on Subnational Absence** — Singapore has no subnational environmental legislation. All species and plant restrictions are implemented nationally through NParks, NEA, and AVA (Agri-Food and Veterinary Authority — now SFA). The zonal management system described above represents the functional equivalent of subnational variation in a city-state context.
</div>

<div class="alert alert-info" role="alert">
**Source note** — NParks: [nparks.gov.sg](https://www.nparks.gov.sg) | Heritage Trees: [nparks.gov.sg/gardens-parks-and-nature/heritage-trees](https://www.nparks.gov.sg) | Parks and Trees Act: [Singapore Statutes Online](https://sso.agc.gov.sg) | URA Master Plan: [ura.gov.sg](https://www.ura.gov.sg).
</div>

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| NParks (National Parks Board) | Parks and Trees Act administration; Heritage Trees; nature reserves; CITES management authority |
| AVS (Animal and Veterinary Service) | Animal and plant import/export controls; biosecurity |
| NEA (National Environment Agency) | Environmental quality; pollution controls; environmental health |
| Singapore Botanic Gardens (SBGC) | UNESCO World Heritage Site; orchid breeding; seed bank; ex-situ conservation |
| Lee Kong Chian Natural History Museum | Herbarium collections; plant taxonomy; species identification |
| NUS Dept of Biological Sciences | Botanical research; flora surveys; Red Data Book assessments |

---

# Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Singapore Status`=legal_status_sg,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Protected Plant Species in Singapore",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Singapore Status`=coalesce(legal_status_sg,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Singapore — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Remaining native forest extremely fragmented — any development adjacent to reserves creates edge effects | High |
| 2 | Invasive species (Mikania, Mimosa) management in secondary forest areas resource-constrained | High |
| 3 | Climate change — Singapore temperatures rising; montane species have no altitude refuge to migrate | Medium |
| 4 | Heritage Tree programme focuses on individual specimens, not populations or habitats | Medium |
| 5 | Red Data Book (3rd Ed.) species listings not all legally formalised under Parks and Trees Act | Medium |
| 6 | Limited seed banking capacity relative to number of critically threatened native plants | Low |

---

# References {#references}

| Resource | Link |
|----------|------|
| NParks Singapore | [nparks.gov.sg](https://www.nparks.gov.sg) |
| Singapore Botanic Gardens | [sbg.org.sg](https://www.sbg.org.sg) |
| Singapore Red Data Book | [NParks RDB](https://www.nparks.gov.sg/biodiversity/wildlife-in-singapore/species-list) |
| GBIF Singapore | [gbif.org/country/SG](https://www.gbif.org/country/SG/summary) |
| CITES Singapore | [cites.org — Singapore](https://cites.org/eng/cms/index.php/component/cp/country/SG) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Verify at NParks / Singapore Red Data Book / CITES</p></div>'))
```
