---
title: "Restricted and Prohibited Plant Species in South Africa"
subtitle: "NEMBA TOPS Regulations, AIS Lists, and CITES — A Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  fig.width  = 9,
  fig.height = 5,
  out.width  = "100%"
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(kableExtra)
  library(glue)
  library(janitor)
  library(rgbif)
})

# ── Load kwizbranding ─────────────────────────────────────────────────────────
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co  <- kqb_company("research")
kqb_col <- kqb_colours()

.kqb_logo_candidates <- c(
  file.path(here::here(), "docs", "www"),
  file.path(getwd(), "www"),
  file.path(getwd()),
  file.path(here::here())
)
.kqb_load_logo <- function(filename, cache_key) {
  for (d in .kqb_logo_candidates) {
    p <- file.path(d, filename)
    if (file.exists(p)) {
      raw  <- readBin(p, "raw", file.size(p))
      b64  <- base64enc::base64encode(raw)
      assign(cache_key, paste0("data:image/svg+xml;base64,", b64), envir = .GlobalEnv)
      return(invisible(TRUE))
    }
  }
  warning("[kwizbranding] Could not find ", filename)
}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",    ".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg", ".kqb_display_cache")

# ── Helper: styled kable ──────────────────────────────────────────────────────
styled_kable <- function(df, caption = NULL, scroll = TRUE, height = "450px") {
  k <- df |>
    kbl(caption = caption, escape = FALSE, align = "l") |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width        = TRUE,
      font_size         = 12
    )
  if (scroll) k <- scroll_box(k, height = height)
  k
}

# ── Embedded species data ─────────────────────────────────────────────────────
# Source: NEMBA TOPS Regulations (GN R 151/2007 updated), AIS Regulations (GN 1003/2020),
#         CITES Appendices (CoP19, 2022)
species <- tribble(
  ~common_name, ~scientific_name, ~taxonomic_group, ~iucn_status, ~cites_appendix, ~legal_status_sa, ~restriction_type, ~governing_law, ~managing_institution, ~notes, ~source_url,
  # ── TOPS — Cycads (Encephalartos spp. — ALL CITES Appendix I) ─────────────────
  "Albany cycad", "Encephalartos latifrons", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Most endangered cycad globally; heavily poached; Eastern Cape", "https://www.saflii.org",
  "Wood's cycad", "Encephalartos woodii", "Cycadopsida", "Extinct in the Wild", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Extinct in wild; only male plants known; clones in botanic gardens", "https://www.saflii.org",
  "Blue Drakensberg cycad", "Encephalartos nubimontanus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Drakensberg endemic; <100 plants in wild", "https://www.saflii.org",
  "Eastern Cape blue cycad", "Encephalartos princeps", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Eastern Cape endemic; extreme poaching pressure", "https://www.saflii.org",
  "Venda cycad", "Encephalartos hirsutus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Limpopo endemic; limited remaining wild population", "https://www.saflii.org",
  "Transvaal cycad", "Encephalartos dyerianus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Limpopo; very few in wild outside cultivation", "https://www.saflii.org",
  "Cerulean cycad", "Encephalartos cerinus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Limpopo; fewer than 60 plants known", "https://www.saflii.org",
  "Inopinus cycad", "Encephalartos inopinus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Limpopo; <200 plants; restricted rocky habitat", "https://www.saflii.org",
  "Dolomite cycad", "Encephalartos dolomiticus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Wolkberg dolomite cliffs; very few plants; climbing poachers", "https://www.saflii.org",
  "Barberton cycad", "Encephalartos brevifoliolatus", "Cycadopsida", "Extinct in the Wild", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Extinct in wild (poached); Barberton; only garden specimens survive", "https://www.saflii.org",
  "Heenani cycad", "Encephalartos heenanii", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Mpumalanga escarpment; <100 wild individuals", "https://www.saflii.org",
  "Middleburg cycad", "Encephalartos middelburgensis", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Mpumalanga grassland endemic; severely poached", "https://www.saflii.org",
  "Cupidus cycad", "Encephalartos cupidus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Limpopo; <300 individuals; cliff habitat", "https://www.saflii.org",
  "Laevifolius cycad", "Encephalartos laevifolius", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Mpumalanga/Limpopo; grassland; population plummeted from poaching", "https://www.saflii.org",
  "Modjadji cycad", "Encephalartos transvenosus", "Cycadopsida", "Near Threatened", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Modjadji Cycad Reserve, Limpopo; iconic Rain Queen forests", "https://www.saflii.org",
  "Bread palm", "Encephalartos altensteinii", "Cycadopsida", "Near Threatened", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Widely cultivated; wild populations threatened by collection", "https://www.saflii.org",
  "Natal cycad", "Encephalartos natalensis", "Cycadopsida", "Near Threatened", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "KZN coastal; common in trade; wild populations declining", "https://www.saflii.org",
  "Karoo cycad", "Encephalartos lehmannii", "Cycadopsida", "Near Threatened", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Eastern Cape; blue-leaved; heavily collected", "https://www.saflii.org",
  "Lebombo cycad", "Encephalartos lebomboensis", "Cycadopsida", "Near Threatened", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "KZN/Mpumalanga; Lebombo mountains; cliff dweller", "https://www.saflii.org",
  "Piet Retief cycad", "Encephalartos arenarius", "Cycadopsida", "Endangered", "Appendix I", "Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Eastern Cape coastal dunes; sand-dwelling; few hundred plants", "https://www.saflii.org",
  "Cycad (horridus)", "Encephalartos horridus", "Cycadopsida", "Endangered", "Appendix I", "Endangered (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Eastern Cape; spiny blue leaves; heavily traded in succulent market", "https://www.saflii.org",
  "Cycad (longifolius)", "Encephalartos longifolius", "Cycadopsida", "Near Threatened", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Eastern Cape; versatile habitat; poaching pressure", "https://www.saflii.org",
  "Cycad (villosus)", "Encephalartos villosus", "Cycadopsida", "Near Threatened", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Eastern seaboard; most common SA cycad; still legally protected", "https://www.saflii.org",
  "Cycad (eugene-maraisii)", "Encephalartos eugene-maraisii", "Cycadopsida", "Endangered", "Appendix I", "Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Waterberg, Limpopo; glaucous blue leaves; heavily poached", "https://www.saflii.org",
  "Cycad (msinganus)", "Encephalartos msinganus", "Cycadopsida", "Critically Endangered", "Appendix I", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "KZN Msinga area; discovered 2003; very restricted range", "https://www.saflii.org",
  # ── TOPS — Fynbos / Cape Flora ────────────────────────────────────────────────
  "Clanwilliam cedar", "Widdringtonia cedarbergensis", "Pinopsida", "Critically Endangered", "Not listed", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Cederberg endemic; <100 mature trees; fire and climate change", "https://www.saflii.org",
  "Fragrant sugarbush", "Protea odorata", "Magnoliopsida", "Critically Endangered", "Not listed", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "<250 mature plants in Swartland fynbos", "https://www.saflii.org",
  "Whorled heath", "Erica verticillata", "Magnoliopsida", "Extinct in the Wild", "Not listed", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Extinct in wild; maintained in Kirstenbosch ex-situ collections", "https://www.saflii.org",
  "Elim conebush", "Leucadendron elimense", "Magnoliopsida", "Critically Endangered", "Not listed", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Three surviving subpopulations; Agulhas Plain endemic", "https://www.saflii.org",
  "Marsh rose / Orothamnus", "Orothamnus zeyheri", "Magnoliopsida", "Endangered", "Not listed", "Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Fernkloof/Kogelberg endemic; marshy fynbos; very restricted range", "https://www.saflii.org",
  "Snow protea", "Protea cryophila", "Magnoliopsida", "Endangered", "Not listed", "Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Swartberg high-altitude endemic; climate change vulnerable", "https://www.saflii.org",
  "Sceptre protea", "Protea aristata", "Magnoliopsida", "Vulnerable", "Not listed", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Swartberg Pass endemic; fire-sensitive; few populations", "https://www.saflii.org",
  "Aulax cancellata", "Aulax cancellata", "Magnoliopsida", "Vulnerable", "Not listed", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "W Cape fynbos; channel-leaf female spicate proteoid; declining", "https://www.saflii.org",
  "Serruria furcellata", "Serruria furcellata", "Magnoliopsida", "Critically Endangered", "Not listed", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Scarlet spiderhead; Cape Flats lowland fynbos; urbanisation threat", "https://www.saflii.org",
  "Mimetes stokoei", "Mimetes stokoei", "Magnoliopsida", "Critically Endangered", "Not listed", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Kogelberg pagoda; rediscovered 2010 after believed extinct; <50 plants", "https://www.saflii.org",
  "Buchu", "Agathosma betulina", "Magnoliopsida", "Vulnerable", "Not listed", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Fynbos medicinal plant; over-harvested for export essential oil", "https://www.saflii.org",
  "Honeybush", "Cyclopia genistoides", "Magnoliopsida", "Vulnerable", "Not listed", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Fynbos tea plant; wild harvest regulated; several Cyclopia spp. threatened", "https://www.saflii.org",
  "Cape orchid (Disa)", "Disa uniflora", "Magnoliopsida", "Least Concern", "Appendix II", "Protected (TOPS)", "Regulated collection/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Pride of Table Mountain; all Orchidaceae CITES II; iconic", "https://www.saflii.org",
  "Blushing bride protea", "Serruria florida", "Magnoliopsida", "Vulnerable", "Not listed", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / CapeNature", "Wedding flower; Franschhoek/Stellenbosch fynbos; habitat loss", "https://www.saflii.org",
  # ── TOPS — Aloes and Succulents ───────────────────────────────────────────────
  "Bastard quiver tree", "Aloe pillansii", "Magnoliopsida", "Endangered", "Appendix II", "Endangered (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Northern Cape; declining from climate change; no recruitment observed", "https://www.saflii.org",
  "Spiral aloe", "Aloe polyphylla", "Magnoliopsida", "Vulnerable", "Appendix I", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Lesotho highlands endemic; most traded succulent globally; CITES I", "https://www.saflii.org",
  "Cape aloe", "Aloe ferox", "Magnoliopsida", "Least Concern", "Appendix II", "Protected (TOPS)", "Regulated harvest/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Commercially harvested for bitter aloe; permit required; sustainable", "https://www.saflii.org",
  "Quiver tree", "Aloidendron dichotomum", "Magnoliopsida", "Vulnerable", "Appendix II", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Iconic Namaqualand tree aloe; climate-driven mortality increasing", "https://www.saflii.org",
  "Giant quiver tree", "Aloidendron pillansii", "Magnoliopsida", "Endangered", "Appendix II", "Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Richtersveld; tallest aloe; <200 reproductive individuals", "https://www.saflii.org",
  "Pancake succulent", "Euphorbia obesa", "Magnoliopsida", "Vulnerable", "Appendix II", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Karoo endemic; heavily poached for succulent trade globally", "https://www.saflii.org",
  "Medusa's head euphorbia", "Euphorbia caput-medusae", "Magnoliopsida", "Least Concern", "Appendix II", "Protected (TOPS)", "Regulated collection/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "W Cape endemic; all succulent Euphorbia CITES II; ornamental trade", "https://www.saflii.org",
  "Living rocks", "Lithops spp.", "Magnoliopsida", "Various", "Appendix II", "Protected (TOPS)", "Regulated collection/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Entire genus; popular in succulent trade; poaching risk; Karoo/Namaqualand", "https://www.saflii.org",
  "Conophytum (stone plants)", "Conophytum spp.", "Magnoliopsida", "Various", "Appendix II", "Protected (TOPS)", "Regulated collection/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Genus-level CITES II listing 2022 (CoP19); massive poaching for Asian market", "https://www.saflii.org",
  "Hoodia (appetite suppressant)", "Hoodia gordonii", "Magnoliopsida", "Least Concern", "Appendix II", "Protected (TOPS)", "Regulated collection/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Kalahari/Northern Cape; San traditional use; patented; ABS dispute", "https://www.saflii.org",
  "Drakensberg agapanthus", "Agapanthus walshii", "Magnoliopsida", "Vulnerable", "Not listed", "Vulnerable (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "KZN endemic; heavily collected for horticulture", "https://www.saflii.org",
  "Pondo coconut", "Jubaeopsis caffra", "Magnoliopsida", "Endangered", "Not listed", "Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "World's rarest palm; <500 plants on Wild Coast, Eastern Cape", "https://www.saflii.org",
  # ── TOPS — Medicinal and other threatened ─────────────────────────────────────
  "African potato", "Hypoxis hemerocallidea", "Magnoliopsida", "Least Concern", "Not listed", "Protected (TOPS)", "Regulated harvest/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Heavily traded muthi plant; anti-retroviral claims drove demand", "https://www.saflii.org",
  "Wild ginger", "Siphonochilus aethiopicus", "Magnoliopsida", "Critically Endangered", "Not listed", "Critically Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Extinct from wild in KZN/Gauteng; muthi trade; rhizome harvested", "https://www.saflii.org",
  "Pepper-bark tree", "Warburgia salutaris", "Magnoliopsida", "Endangered", "Not listed", "Endangered (TOPS)", "Protected — no trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "Bark heavily harvested for traditional medicine; functionally extinct in KZN", "https://www.saflii.org",
  "African cherry / Prunus", "Prunus africana", "Magnoliopsida", "Vulnerable", "Appendix II", "Protected (TOPS)", "Protected — permit required", "NEMBA TOPS Regulations 2007 / CITES", "DFFE / SANBI", "Bark for prostate drug; CITES II; Mpumalanga/Limpopo populations", "https://www.saflii.org",
  "Pelargonium (umckaloabo)", "Pelargonium sidoides", "Magnoliopsida", "Least Concern", "Not listed", "Protected (TOPS)", "Regulated harvest/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI / ECPTA", "Root harvested for respiratory remedy; massive export to Germany", "https://www.saflii.org",
  "Sceletium (kanna)", "Sceletium tortuosum", "Magnoliopsida", "Least Concern", "Not listed", "Protected (TOPS)", "Regulated collection/trade", "NEMBA TOPS Regulations 2007", "DFFE / SANBI", "San/Khoi psychoactive plant; supplement industry; ABS requirement", "https://www.saflii.org",
  # ── Protected Trees (National Forests Act s.12) ───────────────────────────────
  "Stinkwood", "Ocotea bullata", "Magnoliopsida", "Vulnerable", "Not listed", "Protected Tree", "Protected — harvesting banned", "National Forests Act 1998 s.12", "DFFE / SANParks", "Valuable timber; Afromontane forests; permit for any disturbance", "https://www.dffe.gov.za",
  "Sneezewood", "Ptaeroxylon obliquum", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Highly durable timber; protected from unsustainable harvest", "https://www.dffe.gov.za",
  "Cape yellowwood", "Podocarpus latifolius", "Pinopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Largest indigenous tree in SA; restricted harvest", "https://www.dffe.gov.za",
  "Real yellowwood", "Podocarpus falcatus", "Pinopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "National tree of South Africa; felling prohibited without licence", "https://www.dffe.gov.za",
  "Outeniqua yellowwood", "Podocarpus elongatus", "Pinopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Breede River Valley and Cederberg; riverine conifer", "https://www.dffe.gov.za",
  "Wild fig", "Ficus spp.", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "All wild Ficus protected; keystone fruit tree; cultural significance", "https://www.dffe.gov.za",
  "Baobab", "Adansonia digitata", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE / LEDET", "Iconic Limpopo tree; some >2,000 yr old; bark harvesting regulated", "https://www.dffe.gov.za",
  "White stinkwood", "Celtis africana", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Urban heritage tree; frequently in municipal conflicts; felling banned", "https://www.dffe.gov.za",
  "Marula", "Sclerocarya birrea", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Fruit tree; Amarula liqueur; culturally significant; felling banned", "https://www.dffe.gov.za",
  "Cabbage tree", "Cussonia spicata", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "All Cussonia spp. protected; medicinal bark; garden specimen theft", "https://www.dffe.gov.za",
  "Ironwood", "Olea capensis", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Afromontane forest; extremely hard timber; slow growing", "https://www.dffe.gov.za",
  "Wild olive", "Olea europaea subsp. africana", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Indigenous olive; widespread; protected from commercial felling", "https://www.dffe.gov.za",
  "Karee", "Searsia lancea", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Karoo/grassland tree; common amenity tree; protected on private land", "https://www.dffe.gov.za",
  "Shepherd's tree", "Boscia albitrunca", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Kalahari iconic tree; witgat; drought-resistant; culturally important", "https://www.dffe.gov.za",
  "Leadwood", "Combretum imberbe", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Limpopo bushveld; extremely dense timber; slow growing; protected", "https://www.dffe.gov.za",
  "Wild pomegranate", "Burchellia bubalina", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Afromontane forest; showy tubular flowers; timber protected", "https://www.dffe.gov.za",
  "Camphor bush", "Tarchonanthus camphoratus", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Aromatic bushveld tree; widespread; protected from clearing", "https://www.dffe.gov.za",
  "Lavender tree", "Heteropyxis natalensis", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Aromatic riverside tree; Mpumalanga/Limpopo; timber protected", "https://www.dffe.gov.za",
  "Tamboti", "Spirostachys africana", "Magnoliopsida", "Least Concern", "Not listed", "Protected Tree", "Protected — permit required", "National Forests Act 1998 s.12", "DFFE", "Toxic latex; beautiful timber; bushveld; felling banned", "https://www.dffe.gov.za",
  # ── Invasive Alien Species (AIS) — Category 1a (compulsory eradication) ───────
  "Triffid weed", "Chromolaena odorata", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Major KZN/Mpumalanga invader; transforms forest margins; fire risk", "https://www.dffe.gov.za",
  "Pompom weed", "Campuloclinium macrocephalum", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Gauteng/Mpumalanga grassland invader; displaces native forbs", "https://www.dffe.gov.za",
  # ── Invasive Alien Species (AIS) — Category 1b (compulsory control) ───────────
  "Water hyacinth", "Eichhornia crassipes", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Major threat to water bodies; fastest growing aquatic weed", "https://www.dffe.gov.za",
  "Common lantana", "Lantana camara", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Major invasive in KZN and Limpopo riparian zones; toxic berries", "https://www.dffe.gov.za",
  "Black wattle", "Acacia mearnsii", "Magnoliopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit required; control outside designated areas", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Massive water user; transforms fynbos and grassland; some commercial use", "https://www.dffe.gov.za",
  "Silver wattle", "Acacia dealbata", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Invades riparian zones in Western Cape and KZN; fire risk", "https://www.dffe.gov.za",
  "Port Jackson willow", "Acacia saligna", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Transforms fynbos; biological control agents deployed (gall rust)", "https://www.dffe.gov.za",
  "Long-leaved wattle", "Acacia longifolia", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Coastal dune invader; Western/Eastern Cape; displaces dune fynbos", "https://www.dffe.gov.za",
  "Australian blackwood", "Acacia melanoxylon", "Magnoliopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit required; control outside designated areas", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Invades forest margins and riparian areas; some commercial forestry", "https://www.dffe.gov.za",
  "Rooikrans", "Acacia cyclops", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Dominant coastal dune invader in Western/Southern Cape", "https://www.dffe.gov.za",
  "Silky hakea", "Hakea sericea", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Dense fynbos invader; Cederberg and Agulhas; fire adapted", "https://www.dffe.gov.za",
  "Sweet hakea", "Hakea drupacea", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Invades fynbos; dense thickets; biocontrol limited", "https://www.dffe.gov.za",
  "Sweet prickly pear", "Opuntia ficus-indica", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Extensively naturalised in Karoo; Cactoblastis biocontrol ongoing", "https://www.dffe.gov.za",
  "Jointed cactus", "Opuntia aurantiaca", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Eastern Cape grassland invader; livestock injury; biocontrol deployed", "https://www.dffe.gov.za",
  "Queen of the night cactus", "Cereus jamacaru", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Limpopo/Northern Cape; tall columnar; spreading along fence lines", "https://www.dffe.gov.za",
  "Camphor tree", "Cinnamomum camphora", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Invades KZN/EC forests and coastal areas; bird-dispersed", "https://www.dffe.gov.za",
  "Honey mesquite", "Prosopis glandulosa", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Dryland invasive; millions of ha in Northern Cape; water table impact", "https://www.dffe.gov.za",
  "Velvet mesquite", "Prosopis velutina", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Northern Cape arid zone; hybridises with P. glandulosa", "https://www.dffe.gov.za",
  "Giant reed", "Arundo donax", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Invades rivers and wetlands nationwide; water consumption", "https://www.dffe.gov.za",
  "Spanish reed", "Arundo donax var. versicolor", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Variegated form; ornamental escape; same impacts as type", "https://www.dffe.gov.za",
  "Water lettuce", "Pistia stratiotes", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Floating aquatic weed; blocks waterways; Limpopo/Mpumalanga", "https://www.dffe.gov.za",
  "Salvinia", "Salvinia molesta", "Polypodiopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Floating fern; smothers dams and irrigation canals; biocontrol deployed", "https://www.dffe.gov.za",
  "Red water fern", "Azolla filiculoides", "Polypodiopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Floating fern; nitrogen fixer; deoxygenates water bodies", "https://www.dffe.gov.za",
  "Parrot's feather", "Myriophyllum aquaticum", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Submerged aquatic; chokes waterways; ornamental escape", "https://www.dffe.gov.za",
  "Bugweed", "Solanum mauritianum", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Major forest margin invader; KZN/Mpumalanga; bird-dispersed", "https://www.dffe.gov.za",
  "American bramble", "Rubus cuneifolius", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Grassland and forest margin; Mpumalanga/KZN; thorny thickets", "https://www.dffe.gov.za",
  "Syringa / Persian lilac", "Melia azedarach", "Magnoliopsida", "Least Concern", "Not listed", "Category 3 Invasive (no new planting)", "No new plantings; existing trees tolerated", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Widespread urban tree; toxic berries; riverbank invader", "https://www.dffe.gov.za",
  "Jacaranda", "Jacaranda mimosifolia", "Magnoliopsida", "Least Concern", "Not listed", "Category 3 Invasive (no new planting)", "No new plantings; existing trees tolerated", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Iconic Pretoria street tree; no new plantings allowed; heritage debate", "https://www.dffe.gov.za",
  "Castor oil plant", "Ricinus communis", "Magnoliopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit required; control where invading", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Riverbank invader; toxic seeds; commercial oil production permitted", "https://www.dffe.gov.za",
  "Mauritius thorn", "Caesalpinia decapetala", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Dense thorny thickets in KZN/Mpumalanga river valleys", "https://www.dffe.gov.za",
  "Scotch broom", "Cytisus scoparius", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Invades grasslands in Mpumalanga/KZN highveld", "https://www.dffe.gov.za",
  "Brazilian pepper", "Schinus terebinthifolia", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Coastal KZN invader; bird-dispersed; dense riverine thickets", "https://www.dffe.gov.za",
  "Pine species (radiata)", "Pinus radiata", "Pinopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit for plantation; control in fynbos", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Plantation species; invades fynbos and grassland; water impacts", "https://www.dffe.gov.za",
  "Pine species (patula)", "Pinus patula", "Pinopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit for plantation; control in fynbos", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Major forestry species; invades grassland and fynbos margins", "https://www.dffe.gov.za",
  "Pine species (pinaster)", "Pinus pinaster", "Pinopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit for plantation; control in fynbos", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Invades Western Cape fynbos extensively; Working for Water target", "https://www.dffe.gov.za",
  "Eucalyptus (various)", "Eucalyptus camaldulensis", "Magnoliopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit for plantation; control in riparian zones", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "River red gum; massive water consumer; riparian zone invader", "https://www.dffe.gov.za",
  "Poplar (various)", "Populus canescens", "Magnoliopsida", "Least Concern", "Not listed", "Category 2 Invasive (permit)", "Permit required; control in riparian zones", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Grey poplar; riparian invader; root suckers spread aggressively", "https://www.dffe.gov.za",
  "Kariba weed", "Salvinia auriculata", "Polypodiopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Floating fern; dam surfaces; confused with S. molesta; both controlled", "https://www.dffe.gov.za",
  "Parthenium weed", "Parthenium hysterophorus", "Magnoliopsida", "Least Concern", "Not listed", "Category 1b Invasive", "Compulsory control/removal", "NEMBA AIS Regulations 2020", "DFFE / DAFF", "Recent arrival spreading through Limpopo/Mpumalanga; allergen", "https://www.dffe.gov.za"
)

# ── Summary counts ────────────────────────────────────────────────────────────
n_species         <- nrow(species)
n_tops            <- sum(str_detect(species$governing_law, "TOPS"), na.rm = TRUE)
n_ais             <- sum(str_detect(species$governing_law, "AIS"), na.rm = TRUE)
n_protected_trees <- sum(str_detect(species$governing_law, "National Forests"), na.rm = TRUE)
n_cites_i         <- sum(species$cites_appendix == "Appendix I",  na.rm = TRUE)
n_cites_ii        <- sum(species$cites_appendix == "Appendix II", na.rm = TRUE)
n_cr              <- sum(species$iucn_status == "Critically Endangered", na.rm = TRUE)
n_en              <- sum(species$iucn_status == "Endangered", na.rm = TRUE)

# ── GBIF resolution ──────────────────────────────────────────────────────────
proj_root     <- here::here()
gbif_cache    <- file.path(proj_root, "output", "tables", "south_africa_regulated_plants_gbif.csv")

if (!file.exists(gbif_cache)) {
  message("[setup] Running GBIF backbone resolution for South Africa plant species ...")

  clean_name <- function(nm) {
    nm |> str_remove_all("\\s+spp\\.?$") |> str_remove_all("\\(.*?\\)") |> str_squish()
  }

  names_df <- tibble(
    name    = clean_name(species$scientific_name),
    kingdom = "Plantae"
  )

  backbone_res <- tryCatch(
    name_backbone_checklist(names_df),
    error = function(e) {
      map_dfr(names_df$name, function(nm) {
        tryCatch(as_tibble(name_backbone(name = nm, kingdom = "Plantae")),
                 error = function(e2) tibble(usageKey = NA_integer_,
                                             canonicalName = nm, matchType = "NONE"))
      })
    }
  )

  get_vernacular <- function(key) {
    if (is.na(key)) return(NA_character_)
    tryCatch({
      vn <- name_usage(key = as.numeric(key), data = "vernacularNames", limit = 100)
      if (is.null(vn$data) || nrow(vn$data) == 0) return(NA_character_)
      en <- vn$data |> filter(language %in% c("eng", "en")) |> pull(vernacularName) |> unique()
      if (length(en) == 0) NA_character_ else paste(en, collapse = ", ")
    }, error = function(e) NA_character_)
  }

  get_synonyms <- function(key) {
    if (is.na(key)) return(NA_character_)
    tryCatch({
      syn <- name_usage(key = as.numeric(key), data = "synonyms", limit = 100)
      if (is.null(syn$data) || nrow(syn$data) == 0) return(NA_character_)
      s <- syn$data |> pull(canonicalName) |> unique() |> na.omit()
      if (length(s) == 0) NA_character_ else paste(s, collapse = ", ")
    }, error = function(e) NA_character_)
  }

  backbone_clean <- backbone_res |>
    as_tibble() |>
    mutate(
      GBIFusageKey = as.character(usageKey),
      prefName     = coalesce(species, canonicalName, verbatim_name),
      gbif_family  = family, gbif_genus = genus
    )

  species_matched <- species |>
    mutate(clean_sci = clean_name(scientific_name)) |>
    bind_cols(
      backbone_clean |>
        select(GBIFusageKey, prefName, gbif_family, gbif_genus,
               gbif_matchType = matchType, gbif_confidence = confidence)
    )

  keys <- species_matched |>
    filter(!is.na(GBIFusageKey)) |>
    distinct(GBIFusageKey) |>
    pull(GBIFusageKey)

  vern_list <- list(); syn_list <- list()
  for (i in seq_along(keys)) {
    k <- keys[i]
    vern_list[[k]] <- get_vernacular(k)
    syn_list[[k]]  <- get_synonyms(k)
    if (i %% 5 == 0) Sys.sleep(0.2)
  }

  vern_df <- tibble(
    GBIFusageKey  = names(vern_list),
    gbif_english  = unlist(vern_list),
    gbif_synonyms = unlist(syn_list[names(vern_list)])
  )

  plants_enriched <- species_matched |>
    left_join(vern_df, by = "GBIFusageKey") |>
    mutate(
      englishName = coalesce(gbif_english, common_name),
      synonyms    = gbif_synonyms,
      family      = coalesce(gbif_family, NA_character_)
    )

  dir.create(file.path(proj_root, "output", "tables"), recursive = TRUE, showWarnings = FALSE)
  plants_enriched |>
    mutate(country="South Africa",region="",jurisdiction="NATIONAL",jurisdiction_group="",
      classification=taxonomic_group,taxonLevel="species",Note=notes,genus_I=word(prefName,1),is_aquatic="no",aquatic_type_I="") |>
    select(GBIFusageKey,country,region,jurisdiction,jurisdiction_group,prefName,classification,taxonLevel,family,englishName,synonyms,Note,genus_I,is_aquatic,aquatic_type_I,
           legal_status_sa, restriction_type, governing_law) |>
    write_csv(gbif_cache)
}

if (file.exists(gbif_cache)) {
  plants_gbif     <- read_csv(gbif_cache, col_types = cols(.default = col_character()),
                               show_col_types = FALSE)
  n_gbif_resolved <- sum(!is.na(plants_gbif$GBIFusageKey) & plants_gbif$GBIFusageKey != "")
  n_gbif_families <- n_distinct(plants_gbif$family, na.rm = TRUE)
} else {
  plants_gbif     <- NULL
  n_gbif_resolved <- 0L
  n_gbif_families <- 0L
}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity = 0.04, size = "280px auto"))
uri <- kqb_watermark_uri()
if (nzchar(uri)) {
  cat(paste0(
    '<style>\nbody::after {\n  content: "";\n  position: fixed;\n',
    '  top: 0; left: 0; width: 100%; height: 100%;\n  z-index: 9000;\n',
    '  pointer-events: none;\n  background-image: url("', uri, '");\n',
    '  background-repeat: repeat;\n  background-size: 280px auto;\n',
    '  opacity: 0.04;\n}\n</style>\n'
  ))
}
```

---

# Executive Summary {.unnumbered}

South Africa is one of the world's six recognised **megadiverse** countries, harbouring approximately **10% of all plant species on Earth** in just 2% of its land area. The Cape Floristic Region — a global biodiversity hotspot — is home to over **9,600 plant species**, of which approximately **70% are endemic**. Protecting this extraordinary flora requires a multi-layered legal framework that addresses threatened native species, invasive alien plants, regulated harvest, and international trade.

This report compiles **`r n_species` plant species** subject to legal restrictions under South African law. Of these, **`r n_tops`** are listed under the NEMBA Threatened or Protected Species (TOPS) Regulations, **`r n_ais`** are designated invasive alien species (AIS) subject to compulsory control, and **`r n_protected_trees`** are protected under the National Forests Act. **`r n_cites_i`** carry CITES Appendix I listings (commercial trade prohibited) and **`r n_cites_ii`** are on Appendix II. **`r n_cr`** are Critically Endangered and **`r n_en`** are Endangered on the IUCN Red List.

South Africa is one of the few African countries with a fully published, English-language threatened species list directly accessible via the Department of Forestry, Fisheries and Environment (DFFE) and SAFLII.

---

# Legal Framework {#legal-framework}

South Africa's plant protection regime rests on four primary legislative instruments.

## National Environmental Management: Biodiversity Act, 2004 (NEMBA) {#nemba}

The **National Environmental Management: Biodiversity Act, No. 10 of 2004** (NEMBA) is South Africa's primary legislation for biodiversity conservation. It establishes the framework for managing threatened and protected species, alien and invasive species, bioprospecting, and access and benefit sharing.

**Key NEMBA provisions for plants:**

- **Section 56 — Threatened or Protected Species:** Empowers the Minister to publish a list of threatened or protected species (TOPS). All specimens of listed species may only be possessed, traded, or otherwise processed under a TOPS permit.
- **Section 56(8) — Permit System:** TOPS permits are issued by provincial nature conservation agencies (e.g., CapeNature, Ezemvelo KZN Wildlife) and DFFE for national-level activities. Violations carry fines up to **R5 million or 10 years imprisonment**.
- **Chapter 5 — Alien and Invasive Species (AIS):** Designates species in four categories (1a compulsory eradication, 1b compulsory control, 2 permit-required, 3 prohibited) and imposes obligations on landowners to control Category 1 species.
- **Section 83 — Bioprospecting:** Regulates commercial use of indigenous biological resources (including plant-derived compounds), requiring prior informed consent and benefit-sharing agreements.

**TOPS Regulations** (Government Notice R 151, 2 February 2007, as amended): Publish the current list of threatened and protected species with their listing categories. The list is reviewed by SANBI and DFFE and updated via Gazette.

Full text: [SAFLII — NEMBA 2004](https://www.saflii.org/za/legis/num_act/nemba2004355/) | [DFFE](https://www.dffe.gov.za/legislation/actsregulations)

<div class="alert alert-warning" role="alert">
**⚠ Implementation Gap** — The TOPS Regulations were last comprehensively updated through GN 799/2022. However, SANBI's National Assessment of South Africa's Fauna and Flora (NASAFF, 2021) identified **~2,000 plant species** meeting threatened criteria under IUCN standards — far more than are currently listed in the TOPS Regulations. The TOPS list covers a fraction of IUCN-assessed threatened species, meaning the majority have no legal protection status in South Africa.
</div>

## National Forests Act, 1998 {#forests}

The **National Forests Act, No. 84 of 1998** protects indigenous forests and certain tree species.

**Section 12 — Protected Trees:** The Minister may, by Gazette notice, declare any tree species as protected. Felling, damage, removal, or attempted felling of a protected tree is prohibited without a licence from a Forest Officer. The current protected trees list was published in GN R 1030 of 2004 and includes approximately 120 species.

**Section 15 — Use of State Forests:** Regulates harvesting of any forest product from state forests.

Full text: [DFFE — National Forests Act](https://www.dffe.gov.za/legislation/actsregulations/forests)

## National Environmental Management: Biodiversity Act — AIS Regulations, 2020 {#ais}

The **Alien and Invasive Species Regulations** (GN No. R. 1003, 30 September 2020) under NEMBA designate invasive species in four categories:

| Category | Description | Obligation |
|----------|-------------|------------|
| **1a** | Species with no socio-economic benefit; must be eradicated | Compulsory removal and destruction |
| **1b** | Invasive; causes ecological/economic harm | Compulsory control on all land |
| **2** | Invasive in certain habitats; some permitted uses | Permit required for keeping or trade |
| **3** | Potentially invasive; prohibited from new planting | Existing specimens exempt under permit |

Categories 1a and 1b impose a legal duty on landowners to control listed species on their property. Failure to comply allows the responsible authority to carry out the work and recover costs from the landowner.

Full text: [GN R. 1003/2020](https://www.gov.za/documents/national-environmental-management-biodiversity-act-alien-and-invasive-species-regulations)

## CITES Implementation in South Africa {#cites}

South Africa ratified CITES in 1975. DFFE is the designated Management Authority. CITES is implemented domestically through NEMBA regulations.

South Africa hosts significant populations of CITES-listed plant species including all *Encephalartos* (Appendix I), all *Aloe* (Appendix II), all Orchidaceae, all Euphorbia (succulent spp.), and all Cactaceae.

<div class="alert alert-warning" role="alert">
**⚠ Implementation Gap** — South Africa is both a source and transit country for illegal wildlife products. Cycad poaching from wild populations has reached critical levels: in 2019–2022 alone, **over 100 TOPS permit violations** were documented by SANBI involving cycad theft from state land. The courts have imposed fines but rarely custodial sentences, reducing the deterrent effect of the legislation.
</div>

---

# Regional / Subnational Restrictions {#regional}

South Africa's Constitution (Schedule 4) places environment, nature conservation, and pollution as **concurrent national and provincial competencies**. Each of South Africa's 9 provinces has its own conservation legislation — mostly pre-democratic ordinances from the Apartheid era that have been partially retained, partially repealed, and partly superseded by post-1994 national laws. The **National Environmental Management: Biodiversity Act (NEMBA, 2004)** and **TOPS Regulations** establish national minimums; provinces may be stricter.

## Provincial Conservation Legislation

| Province | Key Instrument | Administering Body | Notable Plant Restrictions |
|---------|---------------|-------------------|--------------------------|
| **Western Cape** | Nature Conservation Ordinance 19/1974 (still partially active); Western Cape Nature Conservation Act 1974 + CapeNature regulations | **CapeNature** | Fynbos biome species — 3 protected plant lists maintained by CapeNature; *Cyclopia* (honeybush), *Aspalathus* (rooibos), and *Protea* spp. harvesting requires CapeNature permit; Rooibos harvesting in wild fynbos controlled under Western Cape permit system |
| **KwaZulu-Natal** | KZN Nature Conservation Ordinance 15/1974 + Ezemvelo KZN Wildlife regulations; KZN Biodiversity Act 2008 (Act 2 of 2008) | **Ezemvelo KZN Wildlife** | Cycad (*Encephalartos*) possession registers — Ezemvelo maintains strict cycad ownership records; Natal Fig and Umtambo (*Syzygium cordatum*) in protected areas; medicinal plant market (muthi) monitoring at Warwick Junction |
| **Gauteng** | GDARD (formerly DARD) Gauteng Nature Conservation Ordinance; Urban open space management plans | **GDARD (Gauteng DARD)** | *Aloe* collection in Gauteng requires GDARD permit; alien plant clearing mandated in all listed categories; urban tree removal requires local municipality and GDARD authorisation |
| **Limpopo** | Limpopo Environmental Management Act 7/2003 | **Limpopo Department of Economic Development, Environment and Tourism (LEDET)** | Baobab (*Adansonia digitata*) — cutting prohibited; mopane (*Colophospermum mopane*) harvest restrictions; Mapungubwe landscape species management |
| **Mpumalanga** | Mpumalanga Nature Conservation Act 10/1998; Lowveld NBG conservation regulations | **Mpumalanga Tourism and Parks Agency (MTPA)** | Cycad possession register maintained; mistletoe (*Viscum* spp.) and medicinal bark collection controlled; escarpment fern and orchid restrictions |
| **Eastern Cape** | Eastern Cape Nature Conservation Act 10/1974 (Old Transkei/Ciskei provisions partially retained) | **Eastern Cape Parks and Tourism Agency (ECPTA)** | Albany thicket species; succulent euphorbia collection; *Pelargonium sidoides* (umckaloabo) — provincial harvesting regulation under development given high trade pressure |
| **Northern Cape** | Ordinance 19/1974 successor + environmental management regulation | **Northern Cape DETEA** | Namaqualand succulent and Nama Karoo plant restrictions; *Hoodia gordonii* (appetite suppressant) — TOPS Critically Endangered listing + provincial monitoring; !Khomani San customary land use rights in Kalahari |
| **North West** | North West Nature Conservation Act 19/2014 | **North West DEDECT** | Bushveld species restrictions; Magaliesberg cycad (*Encephalartos dyerianus*) critical populations protected |
| **Free State** | Nature Conservation Ordinance 8/1969 (still active); OFS Nature Conservation Plan 2007 | **Free State DESTEA** | Grassland biome plant restrictions; *Aloe ferox* management; rooibos not applicable (not in biome) |

## Western Cape — Fynbos Biome Special Regime

The **Cape Floristic Region** (a UNESCO World Heritage site) has the world's highest plant diversity for its area — approximately 9,600 species in ~88,000 km². CapeNature maintains dedicated protected plant lists with three tiers:

1. **Protected Plant List A** — may not be collected, damaged, transported, or traded without CapeNature permit (applies outside protected areas)
2. **Protected Plant List B** — restricted harvest permits for commercial use
3. **Protected Plant List C** — notification required for commercial harvest

Commercially important species like *Agathosma betulina* (buchu — essential oil), *Aspalathus linearis* (rooibos — tea), and *Cyclopia* (honeybush) are subject to dedicated Wild Harvest Protocols under the **Rooibos Industry Development Trust** and CapeNature permit requirements.

## Traditional Use and Medicinal Plant Markets

South Africa's Constitution (Section 211) recognises traditional communities' rights. **Medicinal plant (muthi) trade** is a multi-billion rand industry — primarily regulated at provincial level:

- **KZN Ezemvelo**: Warwick Avenue medicinal plant market in Durban — periodic inspections and TOPS compliance checks
- **Gauteng GDARD**: Faraday Street muthi market — monitoring under Traditional Healers and Medicinal Plants Task Team
- Key traded species with provincial monitoring: *Prunus africana* bark, cycad seeds, *Hypoxis hemerocallidea* (African potato), *Siphonochilus aethiopicus* (wild ginger)

<div class="alert alert-info" role="alert">
**Source note** — CapeNature: [capenature.co.za](https://www.capenature.co.za) | Ezemvelo KZN Wildlife: [kznwildlife.com](https://www.kznwildlife.com) | SANBI TOPS: [sanbi.org/biodiversity/foundations-research/plant-conservation/tops/](https://www.sanbi.org) | GDARD: [gdard.gpg.gov.za](https://www.gdard.gpg.gov.za).
</div>

---

# Regulatory Institutions {#institutions}

```{r institutions-table}
institutions <- tribble(
  ~abbreviation, ~institution_name,                              ~parent_ministry,             ~mandate,
  "DFFE",        "Dept of Forestry, Fisheries and Environment",  "The Presidency",             "Primary environmental regulator; issues TOPS and AIS permits; manages protected trees; CITES Management Authority",
  "SANBI",       "South African National Biodiversity Institute","DFFE",                       "Biodiversity science; TOPS species assessments; GBIF node; manages 9 national botanical gardens; Red List authority",
  "SANParks",    "South African National Parks",                  "DFFE",                       "Manages 20 national parks; in-situ conservation; law enforcement within park boundaries",
  "CapeNature",  "Western Cape Nature Conservation Board",        "Western Cape Province",      "Issues TOPS permits in Western Cape; manages provincial nature reserves; fynbos species enforcement",
  "Ezemvelo",    "Ezemvelo KZN Wildlife",                         "KwaZulu-Natal Province",     "Issues TOPS permits in KZN; manages 80+ nature reserves; grassland and coastal species enforcement",
  "DAFF",        "Dept of Agriculture, Land Reform and Rural Dev","DALRRD",                     "Plant Health Directorate; AIS control on agricultural land; invasive species awareness",
  "SANBI-GBIF",  "South African GBIF Node (SANBI)",               "SANBI / DFFE",               "Biodiversity data infrastructure; maintains BioNET-SA portal; publishes GBIF occurrence data",
  "LAPES",       "Legal Affairs and Permits (NEMBA)",             "DFFE",                       "Administers TOPS and AIS national permit system; processes Appendix I CITES import/export applications"
)

institutions |>
  styled_kable(
    caption = "Table 1. Institutions with Plant Species Regulatory Mandates in South Africa",
    height  = "380px"
  )
```

---

# Comprehensive Plant Species Listing {#species}

The master listing covers **`r n_species` plant species** subject to legal restriction under South African law. This includes threatened and protected native species (TOPS), invasive alien species (AIS), protected trees, and CITES-listed taxa.

## Overview by Category

```{r species-category-fig, fig.cap="Species count by South African legal classification."}
species |>
  mutate(
    category = case_when(
      str_detect(governing_law, "AIS")             ~ "Invasive (AIS)",
      str_detect(governing_law, "National Forests") ~ "Protected Tree",
      str_detect(legal_status_sa, "Critically Endangered") ~ "TOPS — Critically Endangered",
      str_detect(legal_status_sa, "Endangered")    ~ "TOPS — Endangered",
      str_detect(legal_status_sa, "Vulnerable")    ~ "TOPS — Vulnerable",
      TRUE                                          ~ "TOPS — Protected"
    )
  ) |>
  count(category) |>
  mutate(category = fct_reorder(category, n)) |>
  ggplot(aes(x = n, y = category, fill = category)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "South Africa Restricted Plant Species by Legal Category",
    subtitle = glue("Total: {n_species} species across NEMBA TOPS, AIS Regulations, and National Forests Act"),
    x        = "Number of species", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

## Master Species Table

```{r master-species-table}
species |>
  arrange(governing_law, iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(
    `Common Name`        = common_name,
    `Scientific Name`,
    `Group`              = taxonomic_group,
    `IUCN Status`        = iucn_status,
    `CITES`              = cites_appendix,
    `SA Legal Status`    = legal_status_sa,
    `Governing Law`      = governing_law,
    `Managing Inst.`     = managing_institution,
    `Notes`              = notes
  ) |>
  styled_kable(
    caption = "Table 2. Comprehensive Restricted/Protected Plant Species in South Africa",
    height  = "600px"
  )
```

## TOPS — Critically Endangered and Endangered Species {#tops-threatened}

The TOPS Critically Endangered and Endangered categories represent the highest tier of domestic legal protection. Trade, possession, and collection of these species requires a TOPS permit (and Appendix I species additionally require CITES permits). In practice, for many Critically Endangered species, no trade is permitted.

```{r tops-threatened-table}
species |>
  filter(iucn_status %in% c("Critically Endangered", "Endangered", "Extinct in the Wild")) |>
  arrange(iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(
    `Common Name`     = common_name,
    `Scientific Name`,
    `IUCN`            = iucn_status,
    `CITES`           = cites_appendix,
    `SA Status`       = legal_status_sa,
    `Notes`           = notes
  ) |>
  styled_kable(
    caption = glue("Table 3. TOPS Critically Endangered and Endangered Plant Species ({sum(species$iucn_status %in% c('Critically Endangered','Endangered','Extinct in the Wild'), na.rm=TRUE)} species)"),
    height  = "500px"
  )
```

## Invasive Alien Species (AIS) {#invasive}

The AIS Regulations designate Category 1 invasive plants for compulsory control. Landowners are legally obligated to control or eradicate these species on their property. DFFE and DAFF manage the Working for Water programme, one of the world's largest invasive species control initiatives (~40,000 ha cleared annually).

```{r ais-table}
species |>
  filter(str_detect(governing_law, "AIS")) |>
  arrange(common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(
    `Common Name`      = common_name,
    `Scientific Name`,
    `AIS Category`     = legal_status_sa,
    `Restriction`      = restriction_type,
    `Notes`            = notes
  ) |>
  styled_kable(
    caption = glue("Table 4. Invasive Alien Plant Species in South Africa — AIS Regulations ({n_ais} species)"),
    height  = "400px"
  )
```

## GBIF Backbone Resolution {#gbif}

The table below resolves South Africa's regulated plant species against the GBIF Backbone Taxonomy. English vernacular names and synonyms are fetched from the GBIF `name_usage()` API.

**Resolution:** `r n_gbif_resolved` of `r n_species` species matched across `r n_gbif_families` families.

```{r gbif-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    mutate(
      gbif_url = if_else(!is.na(GBIFusageKey) & GBIFusageKey != "",
                         glue("https://www.gbif.org/species/{GBIFusageKey}"), NA_character_),
      iucn_url = paste0("https://www.iucnredlist.org/search?query=",
                        URLencode(str_squish(coalesce(prefName, "")), reserved = TRUE)),
      `Scientific Name` = if_else(!is.na(gbif_url),
                                   glue("<a href='{gbif_url}' target='_blank'><em>{prefName}</em></a>"),
                                   glue("<em>{coalesce(prefName, '')}</em>")),
      `IUCN` = glue("<a href='{iucn_url}' target='_blank'>IUCN &#x2197;</a>")
    ) |>
    transmute(
      `#`             = row_number(),
      `Scientific Name`,
      `Common Name`   = coalesce(englishName, ""),
      `Family`        = coalesce(family, ""),
      `GBIF Key`      = coalesce(GBIFusageKey, ""),
      `SA Status`     = coalesce(legal_status_sa, ""),
      `Synonyms`      = str_trunc(coalesce(synonyms, ""), 80),
      `IUCN`
    ) |>
    styled_kable(
      caption = glue("Table 5. South Africa Regulated Plants — GBIF Backbone Resolved ({n_gbif_resolved} of {n_species} matched | {n_gbif_families} families)"),
      height = "600px"
    )
} else {
  cat("*GBIF data not yet generated. The report will auto-generate on first render with internet access.*")
}
```

---

# Regulatory Gaps and Implementation Challenges {#gaps}

```{r gaps-table}
gaps <- tribble(
  ~`#`, ~`Gap Area`,                                              ~`Severity`, ~`Status (2025)`,
  "1",  "TOPS list covers fraction of IUCN-threatened plants",    "High",      "Ongoing gap",
  "2",  "Cycad poaching: enforcement prosecutions rare",          "High",      "Partial improvement",
  "3",  "AIS enforcement on private land inconsistent",           "High",      "Unresolved",
  "4",  "SANBI NASAFF assessments not auto-triggering TOPS updates","Medium",  "Unresolved",
  "5",  "Permit processing backlogs (TOPS and AIS)",              "Medium",    "Unresolved",
  "6",  "Benefit-sharing disputes over indigenous plant knowledge","Medium",   "ABS regulations active",
  "7",  "Protected trees list (GN R1030/2004) not updated",        "Medium",   "Review in progress",
  "8",  "Traditional healers' harvesting lacks legal clarity",    "Medium",    "Unresolved"
)

gaps |>
  kbl(caption = "Table 6. Summary of Regulatory Gaps for Plant Species in South Africa",
      escape = FALSE, align = "l") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed","responsive"),
                full_width = TRUE, font_size = 12) |>
  column_spec(3, bold = TRUE,
              color      = ifelse(gaps$Severity == "High", "white", "black"),
              background = ifelse(gaps$Severity == "High", "#c0392b", "#e67e22")) |>
  column_spec(4, color = ifelse(gaps$`Status (2025)` %in% c("Ongoing gap","Unresolved"),
                                "#c0392b", "#27ae60"))
```

## TOPS Species Coverage Gap

SANBI's 2021 NASAFF assessments evaluated ~7,000 plant species and identified approximately **2,000** meeting IUCN Critically Endangered, Endangered, or Vulnerable criteria. The current TOPS Regulations list a fraction of these. Species not listed in TOPS have no domestic legal protection regardless of IUCN status. The review process — requiring a Ministerial Gazette notice — is slow relative to the rate at which species are being assessed.

## Cycad Poaching Crisis

South Africa hosts the world's greatest cycad diversity with all *Encephalartos* species under CITES Appendix I. Despite this, cycad theft from botanical gardens, nature reserves, and private collections is pervasive. The value of a *E. woodii* plant can exceed R5 million on the black market. Convictions typically result in fines rather than imprisonment, providing insufficient deterrent. The Cycad Conservation Trust and SANBI have proposed mandatory RFID microchipping of all cycad specimens — legislation to mandate this is pending.

## Invasive Species Enforcement on Private Land

The AIS Regulations impose a legal duty on private landowners to control Category 1 species. However, compliance is largely voluntary in practice. DFFE has limited capacity to inspect and enforce on the estimated **1.5 million+ private properties** where Category 1 invasives are present. Working for Water targets primarily state land and contracted community land, leaving the private land gap largely unaddressed.

---

# Key References and Links {#references}

## Official Legal Texts

| Law | Link |
|-----|------|
| NEMBA, 2004 (Act No. 10 of 2004) | [SAFLII](https://www.saflii.org/za/legis/num_act/nemba2004355/) |
| NEMBA TOPS Regulations (GN R 151/2007) | [DFFE](https://www.dffe.gov.za/sites/default/files/legislations/nem_biodiversityact_threatspeciesregulations.pdf) |
| NEMBA AIS Regulations (GN R 1003/2020) | [Gov.za](https://www.gov.za/documents/national-environmental-management-biodiversity-act-alien-and-invasive-species-regulations) |
| National Forests Act, 1998 (No. 84 of 1998) | [SAFLII](https://www.saflii.org/za/legis/num_act/nfa1998209/) |
| National Forests Act Protected Trees (GN R 1030/2004) | [DFFE](https://www.dffe.gov.za/sites/default/files/legislations/nfa_protectedtreeslist.pdf) |

## Regulatory Agencies

| Institution | Website |
|-------------|---------|
| DFFE | [dffe.gov.za](https://www.dffe.gov.za) |
| SANBI | [sanbi.org](https://www.sanbi.org) |
| SANParks | [sanparks.org](https://www.sanparks.org) |
| CapeNature | [capenature.co.za](https://www.capenature.co.za) |

## Species and Conservation Databases

| Resource | Link |
|----------|------|
| CITES Appendices | [cites.org/eng/app/appendices.php](https://cites.org/eng/app/appendices.php) |
| IUCN Red List | [iucnredlist.org](https://www.iucnredlist.org) |
| SANBI Biodiversity Advisor | [sanbi.org/biodiversity-advisor](https://www.sanbi.org/biodiversity-advisor/) |
| BioNET-SA (GBIF node) | [biodiversityadvisor.sanbi.org](https://biodiversityadvisor.sanbi.org) |
| Cycad Conservation Trust | [cycad.co.za](https://www.cycad.co.za) |

---

# Session Information {.appendix}

```{r session-info}
sessionInfo()
```

---

*Report generated with R and knitr. Species data: embedded tribble sourced from NEMBA TOPS Regulations (GN R 151/2007 updated), NEMBA AIS Regulations (GN R 1003/2020), National Forests Act GN R 1030/2004, and CITES Appendices (CoP19, 2022). GBIF taxonomy: rgbif package. For current permit requirements consult DFFE or relevant provincial conservation agency.*

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm <- kqb_logo_img(height = "28px")
cat(paste0(
  '<div class="kqb-report-footer">',
  if (nzchar(logo_sm)) paste0(logo_sm, '<br/>') else "",
  '<p style="font-weight:600;color:#0d2240;">', kqb_co$name, ' &mdash; ', kqb_co$tagline, '</p>',
  '<p>', kqb_co$legal_name, ' | ', kqb_co$country, '</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">',
  'Generated ', format(Sys.Date(), "%d %B %Y"),
  ' &nbsp;&bull;&nbsp; For informational purposes only',
  ' &nbsp;&bull;&nbsp; Verify current listings at DFFE, SANBI, and CITES',
  '</p>',
  '</div>'
))
```
