---
title: "Restricted and Protected Plant Species in Tanzania"
subtitle: "Wildlife Conservation Act, Forests Act, Environmental Management Act, and CITES"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})

if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"), file.path(getwd(),"www"),
                            file.path(getwd()), file.path(here::here()))
.kqb_load_logo <- function(fn, ck) {
  for (d in .kqb_logo_candidates) { p <- file.path(d,fn)
    if (file.exists(p)) { assign(ck, paste0("data:image/svg+xml;base64,",
      base64enc::base64encode(readBin(p,"raw",file.size(p)))), envir=.GlobalEnv); return(invisible(TRUE)) } }
}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")

styled_kable <- function(df, caption=NULL, scroll=TRUE, height="450px") {
  k <- df |> kbl(caption=caption, escape=FALSE, align="l") |>
    kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"), full_width=TRUE, font_size=12)
  if (scroll) k <- scroll_box(k, height=height); k
}

# ── Embedded species data ─────────────────────────────────────────────────────
# Sources: Wildlife Conservation Act 2009 (Cap. 283); Forests Act 2002 (Cap. 323);
#          Environmental Management Act 2004; Plant Protection Act 1997;
#          CITES CoP19 2022; Tanzania Government Gazette.
species <- tribble(
  ~common_name,                     ~scientific_name,               ~taxonomic_group, ~iucn_status,            ~cites_appendix, ~legal_status_tz,                        ~restriction_type,              ~governing_law,                     ~managing_institution,   ~notes,
  # ── Forests Act 2002 — Protected / Reserved Timber Species ──────────────────
  "African blackwood / mpingo",     "Dalbergia melanoxylon",        "Magnoliopsida",  "Least Concern",          "Appendix II",   "Protected Timber Species",              "Export quota / permit req.",    "Forests Act 2002 / CITES",         "TFS / CITES Secretariat","Iconic Tanzanian timber; CITES Appendix II since 2017; global instrument-making",
  "East African rosewood",          "Dalbergia nitidula",           "Magnoliopsida",  "Least Concern",          "Appendix II",   "Protected Timber Species",              "Export permit required",        "Forests Act 2002 / CITES",         "TFS",                   "All Dalbergia spp. CITES App II since 2017; traded as rosewood",
  "East African sandalwood",        "Osyris lanceolata",            "Magnoliopsida",  "Least Concern",          "Not listed",    "Export Banned (TZ Govt Order)",         "Harvesting/export prohibited",  "Ministerial Order (2004)",         "TFS / MNRT",            "Total export ban due to severe over-exploitation across East Africa",
  "African camphor",                "Ocotea usambarensis",          "Magnoliopsida",  "Vulnerable",            "Not listed",    "Vulnerable — Forests Act Protected",    "Harvest restricted",            "Forests Act 2002",                 "TFS",                   "Usambara Mountains; prized timber; declining due to illegal logging",
  "Miombo woodland tree (miombo)",  "Brachystegia spiciformis",     "Magnoliopsida",  "Least Concern",          "Not listed",    "Protected Forest Species",              "Charcoal/harvest restricted",   "Forests Act 2002",                 "TFS",                   "Dominant miombo woodland tree; charcoal production major threat",
  "Brachystegia (mwanga)",          "Brachystegia longifolia",      "Magnoliopsida",  "Least Concern",          "Not listed",    "Protected Forest Species",              "Harvest regulated",             "Forests Act 2002",                 "TFS",                   "Dry miombo woodland; timber and charcoal; restricted harvest",
  "Julbernardia (msani)",           "Julbernardia globiflora",      "Magnoliopsida",  "Least Concern",          "Not listed",    "Protected Forest Species",              "Harvest regulated",             "Forests Act 2002",                 "TFS",                   "Key miombo canopy species; heavily harvested for charcoal and poles",
  "African mahogany",               "Khaya anthotheca",             "Magnoliopsida",  "Vulnerable",            "Not listed",    "Protected Timber Species",              "Harvest restricted / permit",   "Forests Act 2002",                 "TFS",                   "High-value timber; heavily logged; populations declining across range",
  "East African cedar (mlangarua)",  "Juniperus procera",           "Pinopsida",      "Least Concern",          "Not listed",    "Protected Forest Species",              "Harvest restricted",            "Forests Act 2002",                 "TFS",                   "Montane forest species; Mt Kilimanjaro and Southern Highlands; slow-growing",
  "Mvule / iroko",                  "Milicia excelsa",              "Magnoliopsida",  "Near Threatened",        "Not listed",    "Protected Timber Species",              "Harvest permit required",       "Forests Act 2002",                 "TFS",                   "Premier timber tree of East Africa; heavily exploited; restricted felling",
  "African teak (mkora)",           "Pterocarpus angolensis",       "Magnoliopsida",  "Near Threatened",        "Not listed",    "Protected Timber Species",              "Harvest/export restricted",     "Forests Act 2002",                 "TFS",                   "High-value timber; slow growth; log export ban in effect",
  "Mninga / ambila",                "Pterocarpus tinctorius",       "Magnoliopsida",  "Least Concern",          "Not listed",    "Protected Timber Species",              "Harvest regulated",             "Forests Act 2002",                 "TFS",                   "Miombo woodland timber; used for furniture; regulated under Forests Act",
  "Podo / yellowwood",              "Afrocarpus gracilior",         "Pinopsida",      "Least Concern",          "Not listed",    "Protected Forest Species",              "Harvest restricted",            "Forests Act 2002",                 "TFS",                   "Montane forest conifer; Eastern Arc and Southern Highlands; timber valued",
  "African olive",                  "Olea capensis",                "Magnoliopsida",  "Least Concern",          "Not listed",    "Protected Forest Species",              "Harvest regulated",             "Forests Act 2002",                 "TFS",                   "Eastern Arc montane forests; important canopy species; timber and poles",
  "Mangrove (mkoko)",               "Rhizophora mucronata",         "Magnoliopsida",  "Least Concern",          "Not listed",    "Protected Forest Species",              "Harvest permit required",       "Forests Act 2002",                 "TFS",                   "Mangrove forests protected under Forests Act; pole harvest regulated by TFS",
  "Highland bamboo",                "Arundinaria alpina",           "Liliopsida",     "Least Concern",          "Not listed",    "Protected Forest Species",              "Harvest permit required",       "Forests Act 2002",                 "TFS",                   "Southern Highlands and Kilimanjaro; regulated harvest for construction",
  # ── Wildlife Conservation Act 2009 — Protected Plants ───────────────────────
  "Tanzanian violet",               "Saintpaulia ionantha",         "Magnoliopsida",  "Critically Endangered", "Not listed",    "Critically Endangered (WCA 2009)",      "Collection/trade prohibited",   "Wildlife Conservation Act 2009",   "TAWA / TAWIRI",         "African violet; wild ancestor of popular ornamental; East Usambara endemic",
  "African violet (Nguru)",         "Saintpaulia grotei",           "Magnoliopsida",  "Endangered",            "Not listed",    "Endangered — habitat loss",             "Collection prohibited",         "Wildlife Conservation Act 2009",   "TAWA",                  "Nguru Mountains endemic Saintpaulia; extremely restricted range",
  "Saintpaulia (Uluguru)",          "Saintpaulia goetzeana",        "Magnoliopsida",  "Endangered",            "Not listed",    "Endangered — Eastern Arc endemic",      "Collection prohibited",         "Wildlife Conservation Act 2009",   "TAWA",                  "Uluguru Mountains endemic; forest habitat declining; global ornamental trade ancestor",
  # ── CITES-listed Tanzanian Plants ───────────────────────────────────────────
  "Prunus africana",                "Prunus africana",              "Magnoliopsida",  "Least Concern",          "Appendix II",   "Regulated (CITES Appendix II)",         "Export permit required",        "CITES Regs / Forests Act",         "TFS / CITES Authority", "Over-exploited for European prostate medication market; Kilimanjaro populations",
  "Usambara cycad",                 "Encephalartos hildebrandtii",  "Cycadopsida",    "Vulnerable",            "Appendix I",    "Protected (CITES Appendix I)",          "Collection/trade prohibited",   "Wildlife Conservation Act 2009",   "TAWA",                  "Coastal Tanzania cycad; CITES App I; significant poaching pressure",
  "Uluguru cycad",                  "Encephalartos turneri",        "Cycadopsida",    "Critically Endangered", "Appendix I",    "Protected (CITES Appendix I)",          "Collection/trade prohibited",   "Wildlife Conservation Act 2009",   "TAWA",                  "Uluguru Mountains endemic cycad; fewer than 50 mature individuals; CITES App I",
  "East African aloe",              "Aloe rabaiensis",              "Magnoliopsida",  "Near Threatened",        "Appendix II",   "Regulated (CITES Appendix II)",         "Export permit required",        "CITES Regs / WCA 2009",            "TAWA / CITES Authority","Coastal Tanzania aloe; all Aloe spp. regulated under CITES App II",
  "Aloe (spiral aloe relative)",    "Aloe dorotheae",               "Magnoliopsida",  "Critically Endangered", "Appendix II",   "Critically Endangered (CITES App II)",  "Export permit required",        "CITES Regs / WCA 2009",            "TAWA / CITES Authority","Tanzania endemic; extremely small wild population; ornamental trade threat",
  "Aloe (Usambara)",                "Aloe boscawenii",              "Magnoliopsida",  "Critically Endangered", "Appendix II",   "Critically Endangered (CITES App II)",  "Export permit required",        "CITES Regs / WCA 2009",            "TAWA / CITES Authority","East Usambara endemic aloe; coastal forest edge habitat; very restricted range",
  "Tanzanian orchid",               "Eulophia petersii",            "Magnoliopsida",  "Least Concern",          "Appendix II",   "Regulated (CITES Appendix II)",         "Collection permit required",    "CITES Regs / WCA 2009",            "TAWA",                  "All Orchidaceae CITES Appendix II; collection pressure in Eastern Arc",
  "Epiphytic orchid (Eastern Arc)",  "Polystachya paniculata",      "Magnoliopsida",  "Least Concern",          "Appendix II",   "Regulated (CITES Appendix II)",         "Collection permit required",    "CITES Regs / WCA 2009",            "TAWA",                  "Widespread East African orchid; all orchids CITES App II; trade regulated",
  "Succulent euphorbia",            "Euphorbia antiquorum",         "Magnoliopsida",  "Least Concern",          "Appendix II",   "Regulated (CITES Appendix II)",         "Export permit required",        "CITES Regs",                       "TAWA / CITES Authority","Succulent Euphorbia spp. CITES App II; ornamental collection risk",
  "Candelabra euphorbia",           "Euphorbia candelabrum",        "Magnoliopsida",  "Least Concern",          "Appendix II",   "Regulated (CITES Appendix II)",         "Export permit required",        "CITES Regs",                       "TAWA / CITES Authority","Large tree euphorbia; CITES App II; used in traditional medicine and as live fence",
  # ── Eastern Arc Mountains Endemics ──────────────────────────────────────────
  "Usambara fig",                   "Ficus usambarensis",           "Magnoliopsida",  "Endangered",            "Not listed",    "Eastern Arc endemic — forest protected", "Forest reserve protections",   "Forests Act 2002",                 "TFS / TAWIRI",          "East Usambara endemic fig; restricted to submontane forest; habitat declining",
  "Cola (East Usambara)",           "Cola usambarensis",            "Magnoliopsida",  "Vulnerable",            "Not listed",    "Eastern Arc endemic — forest protected", "Forest reserve protections",   "Forests Act 2002",                 "TFS / TAWIRI",          "East Usambara endemic; understorey tree; habitat loss from agriculture",
  "Impatiens (East Usambara)",      "Impatiens usambarensis",       "Magnoliopsida",  "Endangered",            "Not listed",    "Eastern Arc endemic — forest protected", "Forest reserve protections",   "Forests Act 2002 / WCA 2009",      "TAWA / TFS",            "Narrow endemic herb; moist montane forest; extremely vulnerable to disturbance",
  "Coffea (wild Usambara coffee)",  "Coffea pseudozanguebariae",    "Magnoliopsida",  "Vulnerable",            "Not listed",    "Wild genetic resource — protected",     "Collection restricted",         "Forests Act 2002",                 "TFS / TAWIRI",          "Wild coffee relative; valuable genetic resource for coffee breeding; forest dependent",
  # ── Plant Protection Act 1997 — Noxious Weeds ──────────────────────────────
  "Black jack",                     "Bidens pilosa",                "Magnoliopsida",  "Least Concern",          "Not listed",    "Noxious Weed",                          "Control in croplands",          "Plant Protection Act 1997",        "MAFSC",                 "Widespread weed in agricultural zones; priority in Plant Protection Act schedules",
  "Striga / witchweed",             "Striga asiatica",              "Magnoliopsida",  "Least Concern",          "Not listed",    "Noxious Weed — Quarantine Pest",        "Quarantine / control mandated", "Plant Protection Act 1997",        "MAFSC",                 "Parasitic weed devastating cereal crops; quarantine pest under PPA schedules",
  "Dodder",                         "Cuscuta campestris",           "Magnoliopsida",  "Least Concern",          "Not listed",    "Noxious Weed — Quarantine Pest",        "Quarantine / control mandated", "Plant Protection Act 1997",        "MAFSC",                 "Parasitic weed on legumes and vegetables; seed-borne quarantine pest",
  # ── Invasive Alien Species ─────────────────────────────────────────────────
  "Chromolaena / Siam weed",        "Chromolaena odorata",          "Magnoliopsida",  "Least Concern",          "Not listed",    "Invasive Species",                      "Control mandated",              "EMA 2004",                         "NEMC / MAFSC",          "Spreading invasive in coastal Tanzania and Zanzibar; replaces native vegetation",
  "Water hyacinth",                 "Eichhornia crassipes",         "Magnoliopsida",  "Least Concern",          "Not listed",    "Invasive Species",                      "Eradication mandated",          "EMA 2004 / WCA 2009",              "NEMC / TAWA",           "Lake Victoria and other water bodies; requires regular mechanical control",
  "Lantana",                        "Lantana camara",               "Magnoliopsida",  "Least Concern",          "Not listed",    "Invasive Species",                      "Control in protected areas",    "Wildlife Conservation Act 2009",   "TAWA / TANAPA",         "Major invasive in Serengeti and other national parks; biocontrol partially deployed",
  "Prosopis (mesquite)",            "Prosopis juliflora",           "Magnoliopsida",  "Least Concern",          "Not listed",    "Invasive — semi-arid zones",            "Control programmes active",     "EMA 2004 / NEMC",                  "NEMC / MAFSC",          "Invasive in semi-arid northern Tanzania; Arusha and Dodoma regions",
  "Parthenium weed",                "Parthenium hysterophorus",     "Magnoliopsida",  "Least Concern",          "Not listed",    "Invasive Species",                      "Control mandated",              "EMA 2004 / Plant Protection Act",  "NEMC / MAFSC",          "Rapidly spreading invasive; allergenic; displaces pasture; Arusha and northern TZ",
  "Mexican sunflower",              "Tithonia diversifolia",        "Magnoliopsida",  "Least Concern",          "Not listed",    "Invasive — colonising disturbed land",  "Control in protected areas",    "EMA 2004",                         "NEMC / TAWA",           "Aggressive coloniser of forest edges and disturbed sites; Eastern Arc threat",
  "Elephant grass",                 "Cenchrus purpureus",           "Liliopsida",     "Least Concern",          "Not listed",    "Potentially Invasive Grass",            "Control in protected areas",    "EMA 2004 / Wildlife Act",          "NEMC / TAWA",           "Widely used for fodder; becoming invasive in disturbed forest edges"
)

n_species  <- nrow(species)
n_protected<- sum(str_detect(species$governing_law, regex("Wildlife Conservation|Forests Act|WCA", ignore_case=TRUE)), na.rm=TRUE)
n_invasive <- sum(str_detect(species$restriction_type, regex("invasive|eradication|control|noxious|mandated", ignore_case=TRUE)), na.rm=TRUE)
n_cites_i  <- sum(species$cites_appendix == "Appendix I", na.rm=TRUE)
n_cites_ii <- sum(species$cites_appendix == "Appendix II", na.rm=TRUE)
n_cr       <- sum(species$iucn_status == "Critically Endangered", na.rm=TRUE)

proj_root  <- here::here()
gbif_cache <- file.path(proj_root, "output", "tables", "tanzania_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  clean_name <- function(nm) nm |> str_remove_all("\\s+spp\\.?$") |> str_remove_all("\\(.*?\\)") |> str_squish()
  names_df <- tibble(name=clean_name(species$scientific_name), kingdom="Plantae")
  backbone_res <- tryCatch(name_backbone_checklist(names_df), error=function(e)
    map_dfr(names_df$name, function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),
      error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  get_vern <- function(key) { if(is.na(key)) return(NA_character_)
    tryCatch({ vn <- name_usage(key=as.numeric(key),data="vernacularNames",limit=100)
      if(is.null(vn$data)||nrow(vn$data)==0) return(NA_character_)
      en <- vn$data |> filter(language %in% c("eng","en")) |> pull(vernacularName) |> unique()
      if(length(en)==0) NA_character_ else paste(en,collapse=", ") }, error=function(e) NA_character_) }
  get_syn <- function(key) { if(is.na(key)) return(NA_character_)
    tryCatch({ syn <- name_usage(key=as.numeric(key),data="synonyms",limit=100)
      if(is.null(syn$data)||nrow(syn$data)==0) return(NA_character_)
      s <- syn$data |> pull(canonicalName) |> unique() |> na.omit()
      if(length(s)==0) NA_character_ else paste(s,collapse=", ") }, error=function(e) NA_character_) }
  bc <- backbone_res |> as_tibble() |>
    mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species |> mutate(clean_sci=clean_name(scientific_name)) |>
    bind_cols(bc |> select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm |> filter(!is.na(GBIFusageKey)) |> distinct(GBIFusageKey) |> pull(GBIFusageKey)
  vl <- list(); sl <- list()
  for (i in seq_along(keys)) { k <- keys[i]; vl[[k]] <- get_vern(k); sl[[k]] <- get_syn(k); if(i%%5==0) Sys.sleep(0.2) }
  vd <- tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  enriched <- sm |> left_join(vd,by="GBIFusageKey") |>
    mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  enriched |> mutate(country="Tanzania",region="",jurisdiction="NATIONAL",jurisdiction_group="",
    classification=taxonomic_group,taxonLevel="species",Note=notes,genus_I=word(prefName,1),is_aquatic="no",aquatic_type_I="") |>
    select(GBIFusageKey,country,region,jurisdiction,jurisdiction_group,prefName,classification,taxonLevel,family,englishName,synonyms,Note,genus_I,is_aquatic,aquatic_type_I,legal_status_tz,restriction_type,governing_law) |>
    write_csv(gbif_cache)
}
if (file.exists(gbif_cache)) {
  plants_gbif     <- read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE)
  n_gbif_resolved <- sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="")
  n_gbif_families <- n_distinct(plants_gbif$family, na.rm=TRUE)
} else { plants_gbif <- NULL; n_gbif_resolved <- 0L; n_gbif_families <- 0L }
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04, size="280px auto"))
uri <- kqb_watermark_uri()
if (nzchar(uri)) cat(paste0('<style>\nbody::after {\n  content: "";\n  position: fixed;\n',
  '  top:0; left:0; width:100%; height:100%;\n  z-index:9000;\n  pointer-events:none;\n',
  '  background-image: url("', uri, '");\n  background-repeat:repeat;\n  background-size:280px auto;\n',
  '  opacity:0.04;\n}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Tanzania is home to approximately **14,000 plant species**, with an estimated **1,100 endemics**, concentrated particularly in the **Eastern Arc Mountains** — one of Africa's most irreplaceable biodiversity hotspots. The country's plant diversity spans tropical lowland forests, montane forests, miombo and acacia woodlands, mangroves, and alpine moorlands. Tanzania also harbours globally significant populations of *Dalbergia* (rosewood/mpingo) species, which drive major international timber trade.

This report covers **`r n_species` legally restricted plant species** under legal restriction in Tanzania. **`r n_protected`** are protected under the Wildlife Conservation Act 2009 or Forests Act 2002. **`r n_invasive`** are notified invasive species. **`r n_cites_i`** hold CITES Appendix I status and **`r n_cites_ii`** are on CITES Appendix II. **`r n_cr`** are IUCN Critically Endangered.

---

# Legal Framework {#legal-framework}

## Wildlife Conservation Act, 2009 (Cap. 283) {#wca}

The **Wildlife Conservation Act, 2009** (Cap. 283, Revised Edition 2016) is Tanzania's primary wildlife statute. It covers both fauna and certain flora, providing:

- Protection of nationally significant wildlife resources including specific plant species
- Regulation of collection, trade, and export of wildlife products
- CITES implementation for all appendix-listed species
- Management of Game Reserves and Game Controlled Areas

Plant species may be listed via subsidiary regulations and Gazette notices under this Act.

Full text: [TanzLII — WCA 2009](https://tanzlii.org/tz/legislation/act/2009/5/)

## Forests Act, 2002 (Cap. 323) {#forests}

The **Forests Act, 2002** governs all forest resources in Tanzania Mainland. Key provisions:

- **Section 46:** Regulates the harvesting of timber and non-timber forest products in forest reserves
- **Gazette Notices:** The Minister may gazette specific tree species as protected
- **Forest Reserves:** Covering approximately 9.5 million ha; all forest products within reserves require licences
- Regulates export of logs, poles, and processed timber; export bans may be declared by Minister

**Charcoal and wood fuel:** The 2020 Charcoal Regulations under the Forests Act regulate commercial charcoal production. Miombo woodland species (*Brachystegia* spp.) are primary charcoal sources with significant pressure on reserves.

## Environmental Management Act, 2004 {#ema}

The **Environmental Management Act, 2004** (Cap. 191) provides the overarching environmental framework, establishing:

- **National Environmental Management Council (NEMC)** as the environmental regulator
- Environmental Impact Assessment requirements for development activities
- Provisions for invasive species management and biological diversity conservation
- Section 57: Prohibition on introducing invasive species without Environmental Impact Assessment

## Plant Protection Act, 1997 {#ppa}

The **Plant Protection Act, 1997** empowers the Ministry of Agriculture, Food Security and Cooperatives (MAFSC) to:
- Control importation of plant material and quarantine pests
- Gazette noxious weeds for compulsory control
- Regulate movement of plant material within Tanzania to prevent spread of pests and diseases

## CITES Implementation {#cites}

Tanzania ratified CITES in 1980 and is a significant party, having been repeatedly at the centre of elephant ivory and rosewood trade debates. The **CITES Management Authority** for plants operates under MNRT (Ministry of Natural Resources and Tourism).

Critical plant-related CITES milestone: **COP17 (2016)** listed all *Dalbergia* species on Appendix II, directly impacting Tanzania's mpingo (*D. melanoxylon*) export trade. Tanzania had previously relied on *Dalbergia* for significant export revenue and was required to establish and comply with Non-Detriment Findings (NDFs).

<div class="alert alert-warning" role="alert">
**⚠ Implementation Gap** — Tanzania seized 8,069 logs of CITES-regulated rosewood (*Dalbergia* spp.) from Madagascar in 2014. Multiple trials and appeals resulted in courts ordering the logs returned to the shippers in 2023. This case — analysed in [*Diversity* 16(3):183, 2024](https://www.mdpi.com/1424-2818/16/3/183) — revealed that Tanzanian and regional courts lack adequate understanding of CITES Appendix II plant trade provisions.
</div>

---

# Regional / Subnational Restrictions {#regional}

Tanzania is a **United Republic** comprising the mainland (Tanzania Proper) and **Zanzibar** (the semi-autonomous islands of Unguja and Pemba). Zanzibar has its own constitution, parliament, and government with independent authority over most matters including environmental management. This creates a dual regulatory system for plant species. The mainland is divided into 31 administrative regions, each with a **Regional Administrative Secretary** and zonal TAWIRI offices, but most environmental regulation remains centralised at the national level.

## Zanzibar — Separate Legal Framework

Zanzibar's environmental legislation is entirely distinct from mainland Tanzania:

| Instrument | Year | Plant Species Scope |
|-----------|------|---------------------|
| **Zanzibar Environmental Management Act No. 3/2013 (ZEMA)** | 2013 | Framework environmental law for Zanzibar; Zanzibar Environmental Management Authority (ZEMA) regulates biodiversity access; species collection requires ZEMA permit |
| **Zanzibar Forests and Non-Farm Trees Act No. 10/1996** | 1996 | Governs collection of forest products in Zanzibar forests; Jozani-Chwaka Bay National Park (the only NP in Zanzibar) has absolute no-collection zones |
| **Zanzibar Decree Conservation of Wild Plants No. 1/1991** | 1991 | Protects specific plant species in Zanzibar; *Coelacanth* seagrass and mangrove species protected |
| **Zanzibar Spice Industry regulations** | Ongoing | Cloves (*Syzygium aromaticum*) — export monopoly through Zanzibar State Trading Corporation (ZSTC); vanilla, black pepper and nutmeg also regulated for quality and export |
| **Zanzibar Blue Economy / Marine Conservation** | 2021 | Seagrass and mangrove collection restricted under Blue Economy Policy; fishing communities have traditional use rights but commercial collection requires ZEMA and Ministry of Blue Economy permits |

Zanzibar's **Jozani-Chwaka Bay National Park** (established 2004 under the National Parks Act No. 3/2004, Zanzibar) protects coral rag forest with distinct plant communities including endemic *Pandanus* palms and *Uvaria*. Collection within the park requires the Zanzibar National Parks and Reserves authority (ZPNR) permit — TANAPA has no jurisdiction here.

## Mainland Tanzania — Regional Variation

While mainland environmental legislation is primarily national, two significant subnational layers exist:

1. **Wildlife Management Areas (WMAs)**: Community-based areas managed by registered Village Natural Resources Committees (VNRCs). WMAs cover approximately 3.6 million hectares. Non-timber forest product collection within WMAs requires village council approval under WMA rules; commercial plant collection prohibited without additional TAMISEMI authorisation.

2. **Ngorongoro Conservation Area (NCA)**: The Ngorongoro Conservation Area Authority (NCAA) administers the 8,094 km² NCA under the **Ngorongoro Conservation Area Ordinance Cap. 413**. Plant collection (including medicinal plants) within the NCA is prohibited except for authorised researchers; Maasai traditional use of specific plant species (*Acacia* bark, *Olea africana*) is subject to NCAA permit.

3. **Selous Game Reserve / Nyerere National Park**: Despite conversion to national park status in 2019, the buffer zone remains under Tanzania Forest Services (TFS) management with specific NTF product collection restrictions different from national park rules.

4. **Southern Highlands Forest Reserves**: Kilolo, Mufindi, Mufindi highlands — TFS administers forest reserves in Iringa and Njombe regions with specific collection permits for timber, medicinal plants, and bamboo (*Arundinaria alpina*).

<div class="alert alert-info" role="alert">
**Source note** — ZEMA (Zanzibar): [zema.go.tz](https://www.zema.go.tz) | TAWIRI: [tawiri.or.tz](https://www.tawiri.or.tz) | TFS Tanzania: [tfs.go.tz](https://www.tfs.go.tz) | Ngorongoro Conservation Area: [ncaa.go.tz](https://www.ncaa.go.tz).
</div>

---

# Regulatory Institutions {#institutions}

```{r institutions-table}
tribble(
  ~`Abbrev.`, ~`Institution`,                                    ~`Ministry`,     ~`Mandate`,
  "TAWA",     "Tanzania Wildlife Management Authority",           "MNRT",          "Game Reserves management; WCA enforcement; wildlife permits; CITES authority (wildlife)",
  "TANAPA",   "Tanzania National Parks Authority",                "MNRT",          "National park management; invasive species control in parks; tourism regulation",
  "TFS",      "Tanzania Forest Services Agency",                  "MNRT",          "Forests Act enforcement; timber licensing; charcoal regulation; protected tree species",
  "NEMC",     "National Environmental Management Council",        "VP Office",     "Environmental regulation; EMA enforcement; invasive species monitoring; EIA review",
  "TAWIRI",   "Tanzania Wildlife Research Institute",             "MNRT",          "Wildlife research; biodiversity assessments; monitoring; Red List contributions for Tanzania",
  "MAFSC",    "Ministry of Agriculture — Plant Health Directorate","MAFSC",        "Plant Protection Act enforcement; phytosanitary; quarantine; noxious weed control",
  "TOSCI",    "Tanzania Official Seed Certification Institute",   "MAFSC",         "Seed certification; plant genetic resources; seed trade regulation"
) |>
  styled_kable(caption="Table 1. Institutions with Plant Regulatory Mandates in Tanzania", height="360px")
```

---

# Comprehensive Plant Species Listing {#species}

## Master Species Table

```{r master-table}
species |>
  arrange(governing_law, iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`, `Group`=taxonomic_group,
         `IUCN`=iucn_status, `CITES`=cites_appendix,
         `Tanzania Status`=legal_status_tz,
         `Managing Inst.`=managing_institution, `Notes`=notes) |>
  styled_kable(caption="Table 2. Restricted/Protected Plant Species in Tanzania", height="600px")
```

## CITES-Listed and Regulated Timber Species {#timber}

Tanzania's *Dalbergia melanoxylon* (African blackwood, mpingo) is one of the world's most valuable timbers by weight, prized for wind and string instrument manufacturing. Following COP17's listing of all *Dalbergia* on CITES Appendix II, Tanzania established a Non-Detriment Finding (NDF) process and an annual export quota.

```{r timber-table}
species |>
  filter(cites_appendix %in% c("Appendix I","Appendix II") |
         str_detect(governing_law, regex("Forests Act|CITES", ignore_case=TRUE))) |>
  arrange(cites_appendix, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`,
         `IUCN`=iucn_status, `CITES`=cites_appendix,
         `Tanzania Status`=legal_status_tz, `Notes`=notes) |>
  styled_kable(caption="Table 3. CITES-Listed and Protected Timber/Forest Species in Tanzania", height="400px")
```

## Invasive Species {#invasive}

```{r invasive-table}
species |>
  filter(str_detect(restriction_type, regex("invasive|eradication|control|noxious|mandated", ignore_case=TRUE))) |>
  arrange(common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`,
         `Tanzania Status`=legal_status_tz, `Notes`=notes) |>
  styled_kable(caption=glue("Table 4. Invasive Plant Species in Tanzania ({n_invasive} species)"), height="350px")
```

## GBIF Backbone Resolution {#gbif}

```{r gbif-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    mutate(
      gbif_url = if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",
                         glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
      iucn_url = paste0("https://www.iucnredlist.org/search?query=",
                        URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
      `Scientific Name` = if_else(!is.na(gbif_url),
        glue("<a href='{gbif_url}' target='_blank'><em>{prefName}</em></a>"),
        glue("<em>{coalesce(prefName,'')}</em>")),
      `IUCN` = glue("<a href='{iucn_url}' target='_blank'>IUCN &#x2197;</a>")
    ) |>
    transmute(`#`=row_number(), `Scientific Name`, `Common Name`=coalesce(englishName,""),
              `Family`=coalesce(family,""), `GBIF Key`=coalesce(GBIFusageKey,""),
              `Tanzania Status`=coalesce(legal_status_tz,""),
              `Synonyms`=str_trunc(coalesce(synonyms,""),80), `IUCN`) |>
    styled_kable(caption=glue("Table 5. Tanzania Regulated Plants — GBIF Backbone Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} families)"), height="600px")
} else { cat("*GBIF data will be generated on first render with internet access.*") }
```

---

# Complete Species Table (Template Format) {#complete-template}

The table below presents all species in the **18-column `regulated_plants_v9.7` template format** — the canonical client deliverable schema. This is the same data as the downloadable CSV.

```{r complete-template-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    styled_kable(
      caption = "Complete Species Listing — 18-Column Template Format (GBIFusageKey, country, region, jurisdiction, jurisdiction_group, prefName, classification, taxonLevel, family, englishName, synonyms, Note, genus_I, is_aquatic, aquatic_type_I, legal_status, restriction_type, governing_law)",
      height = "650px"
    )
}
```

---

# Regulatory Gaps {#gaps}

```{r gaps-table}
tribble(
  ~`#`, ~`Gap Area`,                                               ~`Severity`, ~`Status (2025)`,
  "1",  "No published comprehensive threatened plant species list", "High",     "Unresolved",
  "2",  "Dalbergia/rosewood: court enforcement failures on CITES",  "High",     "Documented failure (2023 acquittal)",
  "3",  "Mpingo (D. melanoxylon): export quota monitoring gaps",    "High",     "NDF process active",
  "4",  "Charcoal production: miombo woodland degradation",         "High",     "Partial regulations",
  "5",  "Eastern Arc endemic plants — most lack legal protection",  "High",     "Unresolved",
  "6",  "NEMC IAS enforcement limited capacity",                    "Medium",   "Unresolved",
  "7",  "Sandalwood ban enforcement at border posts",               "Medium",   "Partial",
  "8",  "EIA process does not require plant species surveys",       "Medium",   "Unresolved"
) |>
  kbl(caption="Table 6. Regulatory Gaps for Plant Species in Tanzania", escape=FALSE, align="l") |>
  kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"), full_width=TRUE, font_size=12) |>
  column_spec(3, bold=TRUE, color=ifelse(c(TRUE,TRUE,TRUE,TRUE,TRUE,FALSE,FALSE,FALSE),"white","black"),
              background=ifelse(c(TRUE,TRUE,TRUE,TRUE,TRUE,FALSE,FALSE,FALSE),"#c0392b","#e67e22"))
```

## Eastern Arc Mountains — A Plant Biodiversity Hotspot Without Adequate Protection

The Eastern Arc Mountains (EAM) — a chain of 13 ancient forest blocks from southern Kenya to southern Tanzania — are recognised as one of the world's top biodiversity hotspots, with over **2,000 endemic plant species**. Despite this global significance, the vast majority of EAM endemic plant species lack specific domestic legal protection in Tanzania. The WCA 2009 and Forests Act 2002 protect the forest reserves where these plants occur, but species-level listing and monitoring is absent. IUCN assessments document significant threats to EAM endemics including habitat loss, climate change, and collection.

---

# Key References {#references}

| Resource | Link |
|----------|------|
| Wildlife Conservation Act 2009 — TanzLII | [tanzlii.org](https://tanzlii.org/tz/legislation/act/2009/5/) |
| Forests Act 2002 (Cap. 323) — TanzLII | [tanzlii.org](https://tanzlii.org) |
| Environmental Management Act 2004 | [tanzlii.org](https://tanzlii.org) |
| Tanzania Forest Services | [tfs.go.tz](https://www.tfs.go.tz) |
| GBIF Tanzania | [gbif.org/country/TZ](https://www.gbif.org/country/TZ/summary) |
| EAM — Conservation International | [conservation.org](https://www.conservation.org) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm <- kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',
  if(nzchar(logo_sm)) paste0(logo_sm,'<br/>') else "",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; For informational purposes only &nbsp;&bull;&nbsp; Verify at TAWA / TFS / NEMC</p>',
  '</div>'))
```
