---
title: "Restricted and Protected Plant Species in Colombia"
subtitle: "Decreto 1076/2015, Resolución 1912/2017 (MADS), Ley 99/1993 — Preliminary Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,            ~taxonomic_group, ~iucn_status,            ~cites_appendix,  ~legal_status_co,                           ~restriction_type,              ~governing_law,                    ~managing_institution,   ~notes,
  "Quinine tree",            "Cinchona pubescens",         "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Vulnerable (Res. 1912/2017)",              "Regulated collection/trade",   "Res. 1912/2017 / CITES",          "MADS / IAVH",           "Bark source of quinine; Andean cloud forests; over-harvested historically",
  "Rosewood Colombia",       "Aniba rosaeodora",           "Magnoliopsida",  "Endangered",            "Appendix II",    "Endangered (Res. 1912/2017)",              "Collection/trade regulated",   "Res. 1912/2017 / CITES",          "MADS / CAR",            "Amazon timber; essential oil (linalool); heavily exploited for fragrance industry",
  "Wax palm",                "Ceroxylon quindiuense",      "Magnoliopsida",  "Endangered",            "Not listed",     "Endangered (Res. 1912/2017) — National Tree","Collection prohibited",        "Res. 1912/2017 / Decreto 1076",   "MADS / CAR",            "World's tallest palm; Colombia national tree; habitat loss and frond harvest for Palm Sunday",
  "Christmas orchid",        "Cattleya trianae",           "Magnoliopsida",  "Endangered",            "Appendix II",    "Endangered (Res. 1912/2017) — National Flower","Collection/trade regulated","Res. 1912/2017 / CITES",          "MADS / IAVH",           "Colombia national flower; Andean epiphytic orchid; prized in horticulture; wild populations depleted",
  "Frailejón",               "Espeletia grandiflora",      "Magnoliopsida",  "Endangered",            "Not listed",     "Endangered (Res. 1912/2017)",              "Habitat protection",           "Res. 1912/2017 / Decreto 1076",   "MADS / PNN",            "Paramo keystone species; water regulation; climate-sensitive; >3,000 m altitude",
  "Helosis",                 "Helosis cayennensis",        "Magnoliopsida",  "Least Concern",         "Not listed",     "Not specifically listed — under monitoring","Scientific collection only",  "Decreto 1076/2015",               "IAVH",                  "Holoparasitic plant; Amazon; part of broader forest monitoring under SINAP",
  "Giant Amazon lily",       "Victoria amazonica",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected habitat (Amazon biome)",         "Habitat protection",           "Decreto 1076/2015 / Ley 99/1993", "MADS / CARs Amazonia",  "Iconic Amazon species; wetland habitats protected under Colombian Amazon framework",
  "Coca plant",              "Erythroxylum coca",          "Magnoliopsida",  "Least Concern",         "Not listed",     "Controlled — illegal crop (Law 30/1986)", "Cultivation prohibited (illegal)","Law 30/1986 / Decreto 1076",    "MADS / UNODC",          "Legal wild populations exist; commercial cultivation strictly illegal; eradication programmes ongoing",
  "Water hyacinth",          "Eichhornia crassipes",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — management mandated",          "Control mandated",             "Decreto 1076/2015",               "MADS / CARs",           "Major invasive in Magdalena R. and Pacific rivers; covers 30,000+ ha seasonally",
  "Tagua palm",              "Phytelephas aequatorialis",  "Magnoliopsida",  "Near Threatened",       "Not listed",     "Near Threatened (Res. 1912/2017)",         "Regulated harvest",            "Res. 1912/2017",                  "MADS / CAR Chocó",      "Vegetable ivory; Pacific lowlands; overharvested for traditional crafts and export"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","colombia_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_co,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Colombia is one of the world's most biodiverse nations, recognised as a **megadiverse country** with approximately **25,000–30,000 vascular plant species** — around 10% of global plant diversity, of which roughly 30–35% are endemic. The Colombian Andes, Amazon, Chocó, and Orinoquia biomes host extraordinary floristic richness. Plant protection is coordinated under the Ministry of Environment and Sustainable Development (MADS) through Resolución 1912/2017, which lists nationally threatened species across IUCN categories.

<div class="alert alert-info" role="alert">
**Preliminary Report** — Full species listing from Resolución 1912/2017 (MADS national threatened flora list) to be completed. Current report covers `r n_species` representative species. Legislation is primarily in Spanish; SiB Colombia ([sibcolombia.net](https://sibcolombia.net)) provides an English-language biodiversity data interface.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Ley 99/1993 | 1993 | Framework environmental law; establishes MADS and CARs; basis for biodiversity protection |
| Decreto Único Reglamentario Ambiental (Decreto 1076/2015) | 2015 | Consolidated environmental regulations; governs wildlife and flora management |
| Resolución 1912/2017 (MADS) | 2017 | National list of threatened species (CR/EN/VU/NT/DD categories) for fauna and flora |
| Ley 17/1981 | 1981 | CITES implementation in Colombia |
| Law 30/1986 | 1986 | Narcotics statute — coca cultivation restrictions |

Full text: [SiB Colombia](https://sibcolombia.net) | [MADS](https://www.minambiente.gov.co) | [FAOLEX Colombia](https://faolex.fao.org/colombia)

---

# Regional / Subnational Restrictions {#regional}

Colombia's 33 **Regional Autonomous Corporations (Corporaciones Autónomas Regionales — CARs)** are constitutionally mandated agencies that exercise regulatory authority over natural resources in their territories. Under **Ley 99/1993** and **Decreto 1076/2015**, CARs have independent powers to restrict plant collection, set harvest quotas, and designate regional protected areas (*áreas protegidas regionales*) beyond national law. CARs do not merely implement national policy — they are autonomous legal entities with their own budgets and enforcement capacity.

## Key CARs and Their Plant-Species Restrictions

| CAR | Territory | Notable Plant Restrictions |
|-----|-----------|--------------------------|
| **CAR Cundinamarca** | Cundinamarca + Bogotá | Wax palm (*Ceroxylon quindiuense*) frond harvest strictly controlled; Páramo species habitat restrictions; urban tree cutting requires CAR permit |
| **CVC** (Valle del Cauca) | Valle del Cauca + Pacific | *Aniba rosaeodora* (rosewood) — harvest suspension orders; Chocó forest species collection regulated; tagua harvest quotas under PNN coordination |
| **Corantioquia** | Antioquia | *Cattleya trianae* orchid — collection bans in Antioquia range; Alto Cauca riparian forest protection; endemic Antioquia species under restriction |
| **Corpoamazonia** | Putumayo, Caquetá, Amazonas | Amazon plant collection permits issued by Corpoamazonia; *Victoria amazonica* wetland habitats protected; coca eradication coordination |
| **CDA** | Guainía, Vaupés, Vichada | Remote Amazon CARs — minimal enforcement capacity but legal collection permits required; coca and epena (*Virola calophylla*) monitored |
| **CDMB** | Santander (paramos) | Páramo ecosystem species — *Espeletia* spp., orchids; collection prohibited in Páramos Almorzadero and Santurbán |
| **Cornare** | Eastern Antioquia | Protection of Bosque de Niebla (cloud forest) species; endemic orchid collection controls |
| **CAM** | Huila + Alto Magdalena | Wax palm protections in San Agustín area; *Cinchona* regulation; access to botanical collections requires CAM permit |

## Páramo Ecosystem — Special Regional Regime

Colombian páramos (high-altitude moorlands above ~3,200 m) benefit from a heightened legal protection regime enacted through **Resolución 1769/2023** (MADS) and Constitutional Court rulings (Sentencia T-622/2016 on the Atrato River). Key points:

- All páramo ecosystems are designated **exclusion zones** for agriculture, mining, and hydrocarbon extraction under **Ley 1930/2018** (Ley de Páramos).
- Plant species characteristic of páramos (*Espeletia*, *Polylepis*, mosses and cushion plants) are indirectly protected through habitat exclusion.
- Regional CARs must map and delimit páramos in their territories and enforce the exclusion regime — enforcement quality varies substantially by CAR capacity.

## Amazon Resguardos — Indigenous Territorial Restrictions

Indigenous territories (*resguardos*) cover over 25 million hectares of Colombian Amazon and Orinoco basin. Under **Ley 21/1991** (ILO Convention 169) and Constitutional jurisprudence, indigenous communities exercise territorial control over plant resources. Commercial extraction of any plant from a resguardo without community consent (via *consulta previa*) constitutes a legal violation. Several Amazonian communities have self-declared bans on commercial *ayahuasca* vine (*Banisteriopsis caapi*) extraction by non-indigenous operators.

<div class="alert alert-info" role="alert">
**Source note** — CAR portals vary widely in accessibility. Key national portal: [SIAC (Sistema de Información Ambiental de Colombia)](https://www.siac.gov.co) | CAR registry: [Minambiente — CARs](https://www.minambiente.gov.co/departamento/corporaciones-autonomas-regionales/) | Ley Páramos: [Ley 1930/2018 — Colombia Compra Eficiente](https://www.colombiacompra.gov.co).
</div>

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| MADS (Ministry of Environment and Sustainable Development) | National policy, threatened species listing, CITES management authority |
| IDEAM (Institute of Hydrology, Meteorology and Environmental Studies) | Environmental monitoring and information systems |
| IAVH (Instituto de Investigación de Recursos Biológicos Alexander von Humboldt) | Biodiversity research, Red List assessments, SiB Colombia coordination |
| CAR (Corporaciones Autónomas Regionales) | Regional environmental management, permit issuance, enforcement |
| PNN (Parques Nacionales Naturales de Colombia) | Protected areas management (57 national parks) |
| SINCHI (Amazonian Scientific Research Institute) | Amazonian flora research and monitoring |

---

# Preliminary Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Colombia Status`=legal_status_co,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Preliminary — Protected Plant Species in Colombia",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Colombia Status`=coalesce(legal_status_co,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Colombia — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Deforestation rate — Colombia lost ~170,000 ha/year of forest (2010–2020 avg.); enforcement in FARC-demobilised zones weak | High |
| 2 | Res. 1912/2017 covers ~1,500 threatened species but enforcement capacity at CAR level is limited | High |
| 3 | Wax palm (national tree) still harvested for Palm Sunday fronds despite protection | High |
| 4 | Coca eradication programmes (glyphosate aerial spraying) harm non-target native flora including orchids | High |
| 5 | Orchid trade via social media platforms largely unmonitored | Medium |
| 6 | Páramo protection fragmented — multiple overlapping jurisdictions | Medium |

---

# References {#references}

| Resource | Link |
|----------|------|
| MADS — Ministry of Environment | [minambiente.gov.co](https://www.minambiente.gov.co) |
| SiB Colombia — Biodiversity Data | [sibcolombia.net](https://sibcolombia.net) |
| IAVH — Humboldt Institute | [humboldt.org.co](https://www.humboldt.org.co) |
| Resolución 1912/2017 (full text) | [MADS resolutions portal](https://www.minambiente.gov.co/normativa/resoluciones/) |
| GBIF Colombia | [gbif.org/country/CO](https://www.gbif.org/country/CO/summary) |
| CITES Colombia | [cites.org — Colombia](https://cites.org/eng/cms/index.php/component/cp/country/CO) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Preliminary — Verify at MADS / SiB Colombia / CITES</p></div>'))
```
