---
title: "Restricted and Protected Plant Species in Brazil"
subtitle: "Portaria MMA 443/2014 (National Threatened Plants), Lei 12.651/2012 (Native Vegetation Protection Law), Lei 13.123/2015 (Biodiversity Law) — Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name, ~scientific_name, ~taxonomic_group, ~iucn_status, ~cites_appendix, ~legal_status_br, ~restriction_type, ~governing_law, ~managing_institution, ~notes,
  # ── Atlantic Forest Threatened Trees ──────────────────────────────────────────
  "Pau-brasil / Brazilwood", "Paubrasilia echinata", "Magnoliopsida", "Endangered", "Appendix II", "Endangered (Port. 443/2014) — National Tree", "Collection/trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "National tree; bow-making wood; Atlantic Forest endemic; <10% habitat remains",
  "Paraná pine / Araucária", "Araucaria angustifolia", "Pinopsida", "Critically Endangered", "Appendix I", "CR (Port. 443/2014) — CITES App. I", "Felling/collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Atlantic Forest conifer; >97% decline; seeds (pinhão) culturally important",
  "Brazilian rosewood", "Dalbergia nigra", "Magnoliopsida", "Vulnerable", "Appendix I", "VU (Port. 443/2014) — CITES App. I", "All trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Most restricted timber globally; CITES Appendix I since 1992; Atlantic Forest endemic",
  "Peroba-rosa", "Aspidosperma polyneuron", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Premium Atlantic Forest timber; heavily logged; slow regeneration",
  "Canela-preta", "Ocotea catharinensis", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Atlantic Forest laurel; prized timber; severe population decline in SC/PR",
  "Canela-sassafrás", "Ocotea odorifera", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging/extraction prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Sassafras oil source; over-exploited for essential oil industry",
  "Cabreúva", "Myrocarpus frondosus", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Balsam-scented timber; Atlantic Forest; severely reduced by logging",
  "Cedro / Spanish cedar", "Cedrela fissilis", "Magnoliopsida", "Endangered", "Appendix III", "EN (Port. 443/2014)", "Harvest regulated / export controlled", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Premium cabinet timber; CITES App. III; population fragmented across biomes",
  "Imbuia", "Ocotea porosa", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Araucaria Forest associate; prized furniture wood; critically reduced in PR/SC",
  "Jacarandá-da-bahia", "Dalbergia hortensis", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014)", "Trade regulated / export prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Bahian rosewood; all Dalbergia spp. now CITES App. II; smuggling concern",
  "Ipê-roxo", "Handroanthus impetiginosus", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Lapacho timber; medicinal bark (lapachol); heavy international demand",
  "Pau-ferro", "Caesalpinia leiostachya", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Ironwood; extremely dense timber; Atlantic Forest remnants; slow-growing",
  "Jequitibá-rosa", "Cariniana legalis", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Largest Atlantic Forest tree (to 50 m); old-growth indicator species",
  "Virola / Ucuúba", "Virola surinamensis", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Amazon/Atlantic Forest; plywood timber; severely over-logged in Pará",
  "Pau-de-balsa / Balsa", "Ochroma pyramidale", "Magnoliopsida", "Least Concern", "Not listed", "Forest Code protection in APPs", "APP harvest prohibited", "Lei 12.651/2012", "ICMBio / IBAMA", "Lightest commercial timber; riparian zones protected under Forest Code",
  # ── Amazon Timber & NTFPs ─────────────────────────────────────────────────────
  "Brazilian mahogany", "Swietenia macrophylla", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Harvest/export regulated", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "King of tropical timbers; CITES II since 2003; management plans required",
  "Pau-rosa / Rosewood oil", "Aniba rosaeodora", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Harvest/export regulated", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Essential oil (linalool) for perfumery; Chanel No. 5 ingredient; replanting required",
  "Brazil nut", "Bertholletia excelsa", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014) — Protected tree", "Felling prohibited", "Port. MMA 443/2014 / Lei 12.651", "ICMBio / IBAMA", "Amazon icon; felling illegal since 1994; nut harvest sustainable; pollinator-dependent",
  "Acapu", "Vouacapoua americana", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Dense Amazon timber; illegal logging hotspot in Pará; extremely slow growth",
  "Muiracatiara / Tigerwood", "Astronium lecointei", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Premium flooring timber; Amazon; management plan required for harvest",
  "Massaranduba", "Manilkara huberi", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Bullet wood; extremely durable Amazon timber; heavy logging pressure in Pará",
  "Itaúba", "Mezilaurus itauba", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Shipbuilding timber; naturally rot-resistant; over-harvested in Mato Grosso",
  "Jatobá", "Hymenaea courbaril", "Magnoliopsida", "Least Concern", "Not listed", "Forest Code protected", "Harvest regulated", "Lei 12.651/2012", "ICMBio / IBAMA", "Multi-use Amazon/Atlantic Forest tree; resin, timber, food; slow regeneration",
  "Castanha-do-maranhão", "Bombax paraense", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Amazon emergent tree; timber and kapok fiber; declining populations",
  "Pau-rainha", "Brosimum rubescens", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Bloodwood / satinwood; deep red timber; Amazon; logged for specialty markets",
  # ── Palms ─────────────────────────────────────────────────────────────────────
  "Juçara palm", "Euterpe edulis", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated / felling prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Atlantic Forest palm heart source; unsustainable harvest decimated populations",
  "Açaí palm", "Euterpe oleracea", "Magnoliopsida", "Least Concern", "Not listed", "Commercially regulated — ABS", "Commercial use regulated", "Lei 13.123/2015", "MMA / IBAMA", "Amazon superfruit; multi-billion dollar industry; sustainable harvest model",
  "Piaçava palm", "Attalea funifera", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Bahia endemic; fiber palm; over-harvested for broom bristles",
  "Licuri palm", "Syagrus coronata", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Caatinga/Cerrado palm; key food for Lear's macaw (EN); habitat conversion threat",
  "Babaçu palm", "Attalea speciosa", "Magnoliopsida", "Least Concern", "Not listed", "Commercially regulated — Babaçu Law", "Free access for traditional communities", "Lei 12.651/2012 / State Babaçu Laws", "MMA / IBAMA", "Amazon/Cerrado transitional palm; oil and charcoal; quebradeiras rights",
  "Carnaúba wax palm", "Copernicia prunifera", "Magnoliopsida", "Least Concern", "Not listed", "Commercially regulated", "Sustainable harvest permitted", "Lei 12.651/2012 / State regulations", "IBAMA / State agencies", "Tree of life (Caatinga/Cerrado); industrial wax; geographic indication",
  "Buriti palm", "Mauritia flexuosa", "Magnoliopsida", "Least Concern", "Not listed", "APP species — Forest Code", "Felling prohibited in veredas", "Lei 12.651/2012", "ICMBio / IBAMA", "Cerrado vereda palm; keystone species; habitat (veredas) are APPs under Forest Code",
  # ── Orchids (CITES Appendix II) ───────────────────────────────────────────────
  "Cattleya orchid", "Cattleya labiata", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "National orchid; Pernambuco Atlantic Forest; wild collection major threat",
  "Cattleya walkeriana", "Cattleya walkeriana", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Collection/trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Cerrado rupestrian orchid; Minas Gerais/Goiás; prized in horticulture",
  "Cattleya leopoldii", "Cattleya leopoldii", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Collection/trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Santa Catarina Atlantic Forest; large bifoliate; habitat loss",
  "Laelia purpurata", "Cattleya purpurata", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Southern Atlantic Forest; large showy flowers; state flower of SC",
  "Miltonia spectabilis", "Miltonia spectabilis", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Pansy orchid; Atlantic Forest lowlands; fragmented populations",
  "Constantia cipoensis", "Constantia cipoensis", "Magnoliopsida", "Critically Endangered", "Appendix II", "CR (Port. 443/2014) — CITES App. II", "Collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Micro-orchid; Serra do Cipó endemic; rupestrian grasslands; <100 plants",
  "Hoffmannseggella crispata", "Hoffmannseggella crispata", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Rupícola orchid; iron quadrangle Minas Gerais; mining threatens habitat",
  "Zygopetalum mackayi", "Zygopetalum mackayi", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Fragrant terrestrial orchid; Atlantic Forest; over-collected for fragrance",
  "Brasilidium crispum", "Brasilidium crispum", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Formerly Oncidium crispum; Atlantic Forest epiphyte; Rio de Janeiro mountains",
  "Isabelia virginalis", "Isabelia virginalis", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Miniature cloud forest orchid; MG/ES Atlantic Forest; fog-dependent",
  "Phragmipedium vittatum", "Phragmipedium vittatum", "Magnoliopsida", "Critically Endangered", "Appendix I", "CR (Port. 443/2014) — CITES App. I", "All trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Brazilian slipper orchid; only Phragmipedium in Atlantic Forest; <200 plants",
  "Hadrolaelia pumila", "Hadrolaelia pumila", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Miniature Cattleya relative; coastal Atlantic Forest; Rio/ES states",
  "Grandiphyllum divaricatum", "Grandiphyllum divaricatum", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Dancing lady orchid; Atlantic Forest epiphyte; fragmented populations",
  "Epidendrum secundum", "Epidendrum secundum", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "ICMBio / IBAMA", "Widespread but all Orchidaceae CITES II; rupestrian and Atlantic Forest forms",
  # ── Cacti (CITES Appendix I & II) ─────────────────────────────────────────────
  "Uebelmannia", "Uebelmannia pectinifera", "Magnoliopsida", "Endangered", "Appendix I", "EN (Port. 443/2014) — CITES App. I", "All trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Minas Gerais quartzite endemic; entire genus CITES I; poaching for collectors",
  "Discocactus horstii", "Discocactus horstii", "Magnoliopsida", "Critically Endangered", "Appendix I", "CR (Port. 443/2014) — CITES App. I", "All trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Tiny globular cactus; Grão Mogol MG endemic; fewer than 500 plants",
  "Melocactus violaceus", "Melocactus violaceus", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Turk's cap cactus; Atlantic coastal restinga; habitat destruction",
  "Pilosocereus azulensis", "Pilosocereus azulensis", "Magnoliopsida", "Critically Endangered", "Appendix II", "CR (Port. 443/2014) — CITES App. II", "Collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Blue columnar cactus; Bahia endemic; fewer than 250 individuals",
  "Arthrocereus melanurus", "Arthrocereus melanurus", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Night-blooming cactus; iron quadrangle MG; mining destroys habitat",
  "Cipocereus bradei", "Cipocereus bradei", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Serra do Cipó columnar cactus; quartzite specialist; restricted range",
  "Brasilicereus markgrafii", "Brasilicereus markgrafii", "Magnoliopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Collection prohibited", "Port. MMA 443/2014 / CITES", "ICMBio", "Caatinga/Cerrado cactus; arborescent; declining due to land conversion",
  "Tacinga inamoena", "Tacinga inamoena", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "ICMBio / IBAMA", "Caatinga Opuntia relative; all Cactaceae CITES II; ornamental trade",
  "Rhipsalis baccifera", "Rhipsalis baccifera", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "ICMBio / IBAMA", "Mistletoe cactus; only cactus native to Old World; epiphytic; Atlantic Forest",
  "Pereskia grandifolia", "Pereskia grandifolia", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "ICMBio / IBAMA", "Leafy cactus; ora-pro-nóbis; edible leaves; Atlantic Forest; all Cactaceae CITES II",
  # ── Bromeliads ────────────────────────────────────────────────────────────────
  "Alcantarea imperialis", "Alcantarea imperialis", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Collection prohibited", "Port. MMA 443/2014", "ICMBio", "Giant bromeliad; Serra dos Órgãos RJ endemic; ornamental poaching",
  "Vriesea hieroglyphica", "Vriesea hieroglyphica", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Collection prohibited", "Port. MMA 443/2014", "ICMBio", "King of bromeliads; hieroglyphic leaf markings; Atlantic Forest; prized ornamental",
  "Nidularium innocentii", "Nidularium innocentii", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Collection regulated", "Port. MMA 443/2014", "ICMBio", "Bird's nest bromeliad; Atlantic Forest understorey; declining with habitat loss",
  "Fernseea itatiaiae", "Fernseea itatiaiae", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Collection prohibited", "Port. MMA 443/2014", "ICMBio", "High-altitude bromeliad; Itatiaia NP endemic; climate change vulnerable",
  "Dyckia distachya", "Dyckia distachya", "Magnoliopsida", "Critically Endangered", "Not listed", "CR (Port. 443/2014)", "Collection prohibited", "Port. MMA 443/2014", "ICMBio", "Rheophytic bromeliad; Uruguay River rapids; drowned by Itá/Machadinho dams",
  "Cryptanthus zonatus", "Cryptanthus zonatus", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Collection regulated", "Port. MMA 443/2014", "ICMBio", "Earth star; Pernambuco Atlantic Forest; terrestrial; over-collected for trade",
  "Tillandsia stricta", "Tillandsia stricta", "Magnoliopsida", "Least Concern", "Not listed", "Forest Code protection", "Harvest regulated in APPs", "Lei 12.651/2012", "ICMBio / IBAMA", "Widespread air plant; Atlantic Forest epiphyte; harvested for ornamental market",
  "Aechmea nudicaulis", "Aechmea nudicaulis", "Magnoliopsida", "Least Concern", "Not listed", "Forest Code protection", "Harvest regulated", "Lei 12.651/2012", "ICMBio / IBAMA", "Restinga bromeliad; coastal Atlantic Forest; habitat loss from urbanisation",
  # ── Ferns & Lycophytes ────────────────────────────────────────────────────────
  "Xaxim / Tree fern", "Dicksonia sellowiana", "Polypodiopsida", "Endangered", "Appendix II", "EN (Port. 443/2014) — CITES App. II", "Harvest/trade prohibited", "Port. MMA 443/2014 / CITES", "ICMBio / IBAMA", "Southern tree fern; trunks used as planters (xaxim); harvest banned since 2001",
  "Golden tree fern", "Cyathea phalerata", "Polypodiopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Atlantic Forest tree fern; all Cyatheaceae CITES II; montane cloud forest",
  # ── Cerrado & Caatinga Threatened Species ─────────────────────────────────────
  "Aroeira-do-sertão", "Myracrodruon urundeuva", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Caatinga/Cerrado timber; medicinal bark; over-exploited for fence posts",
  "Braúna", "Schinopsis brasiliensis", "Magnoliopsida", "Endangered", "Not listed", "EN (Port. 443/2014)", "Logging prohibited", "Port. MMA 443/2014", "ICMBio / IBAMA", "Caatinga hardwood; tannin source; severe decline from charcoal and timber",
  "Amburana / Cerejeira", "Amburana cearensis", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Caatinga/Cerrado; cumaru timber; medicinal bark; population declining",
  "Pequi", "Caryocar brasiliense", "Magnoliopsida", "Least Concern", "Not listed", "Forest Code protected", "Felling prohibited", "Lei 12.651/2012", "ICMBio / IBAMA", "Cerrado fruit tree; culturally essential in MG/GO; felling banned since 2006",
  "Barbatimão", "Stryphnodendron adstringens", "Magnoliopsida", "Least Concern", "Not listed", "Commercially regulated", "Bark harvest regulated", "Lei 12.651/2012 / State regulations", "IBAMA / State agencies", "Cerrado medicinal bark; tannins for wound healing; over-harvested",
  "Faveira / Fava d'anta", "Dimorphandra mollis", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Cerrado tree; rutin source (pharmaceutical); over-harvested for industry",
  "Umbuzeiro", "Spondias tuberosa", "Magnoliopsida", "Least Concern", "Not listed", "Forest Code protected", "Felling prohibited", "Lei 12.651/2012 / Caatinga regulations", "ICMBio / IBAMA", "Caatinga fruit tree; culturally sacred in sertão; drought-adapted; tuber stores water",
  "Imburana-de-cheiro", "Commiphora leptophloeos", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Caatinga aromatic tree; resin and bark used in folk medicine; declining",
  "Sucupira-preta", "Bowdichia virgilioides", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Cerrado/Atlantic Forest legume tree; durable timber; over-exploited",
  "Gonçalo-alves", "Astronium fraxinifolium", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Cerrado/Caatinga timber; decorative wood; population fragmented",
  # ── Medicinal & ABS-Regulated Species ─────────────────────────────────────────
  "Wild cacao", "Theobroma cacao", "Magnoliopsida", "Least Concern", "Not listed", "ABS regulated — Lei 13.123", "ABS permit required", "Lei 13.123/2015", "MMA / IBAMA", "Wild cacao relatives in Amazon; ABS for commercial genetic use",
  "Guaraná", "Paullinia cupana", "Magnoliopsida", "Least Concern", "Not listed", "ABS regulated — geographic indication", "Commercial use regulated", "Lei 13.123/2015 / INPI", "MMA / IBAMA / INPI", "Amazon stimulant; GI for Maués region; ABS compliance required",
  "Rubber tree", "Hevea brasiliensis", "Magnoliopsida", "Least Concern", "Not listed", "ABS regulated — Lei 13.123", "ABS for genetic resources", "Lei 13.123/2015", "MMA / IBAMA", "Amazon native; rubber industry; wild genetic diversity protected under ABS",
  "Espinheira-santa", "Monteverdia ilicifolia", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Atlantic Forest medicinal holly; anti-ulcer properties; over-harvested",
  "Marapuama / Muirapuama", "Ptychopetalum olacoides", "Magnoliopsida", "Least Concern", "Not listed", "ABS regulated — Lei 13.123", "ABS for commercial use", "Lei 13.123/2015", "MMA / IBAMA", "Amazon aphrodisiac herb; traditional knowledge associated; ABS applies",
  "Jaborandi", "Pilocarpus microphyllus", "Magnoliopsida", "Vulnerable", "Not listed", "VU (Port. 443/2014)", "Harvest regulated", "Port. MMA 443/2014", "ICMBio / IBAMA", "Pilocarpine source (glaucoma drug); Maranhão; wild harvest declining",
  "Andiroba", "Carapa guianensis", "Magnoliopsida", "Least Concern", "Not listed", "ABS regulated — Lei 13.123", "ABS for oil products", "Lei 13.123/2015", "MMA / IBAMA", "Amazon medicinal tree; seed oil for cosmetics/medicine; traditional knowledge",
  "Copaíba", "Copaifera langsdorffii", "Magnoliopsida", "Least Concern", "Not listed", "ABS regulated — Lei 13.123", "ABS for oleoresin", "Lei 13.123/2015", "MMA / IBAMA", "Cerrado/Amazon; oleoresin anti-inflammatory; sustainable tapping possible",
  # ── Succulents & Other CITES-listed ───────────────────────────────────────────
  "Euphorbia phosphorea", "Euphorbia phosphorea", "Magnoliopsida", "Vulnerable", "Appendix II", "VU (Port. 443/2014) — CITES App. II", "Collection/trade regulated", "Port. MMA 443/2014 / CITES", "ICMBio", "Caatinga succulent euphorbia; candlestick-shaped; Bahia endemic",
  "Giant Amazon water lily", "Victoria amazonica", "Magnoliopsida", "Least Concern", "Not listed", "Protected habitat — Amazon regulations", "Habitat protection", "Lei 12.651/2012 / SNUC", "ICMBio / IBAMA", "Amazonian icon; floating leaves to 3 m; habitat protected under Forest Code",
  # ── Invasive Alien Species ────────────────────────────────────────────────────
  "Water hyacinth", "Eichhornia crassipes", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — management mandated", "Control mandated", "Lei 9.605/1998 / IBAMA regulations", "IBAMA / ANA", "Covers >100,000 ha of waterways; hydroelectric and navigation impact",
  "Leucaena", "Leucaena leucocephala", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — control required", "Control mandated in reserves", "IBAMA / ICMBio management plans", "ICMBio / IBAMA", "Major invasive in Atlantic Forest fragments; nitrogen-fixer alters soil",
  "Slash pine", "Pinus elliottii", "Pinopsida", "Least Concern", "Not listed", "Invasive — control in south Brazil", "Removal mandated in PAs", "CONAMA Resolution 369/2006", "ICMBio / State agencies", "Plantation escapee; invades Araucaria forest and grasslands in PR/SC/RS",
  "African oil palm", "Elaeis guineensis", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — Atlantic Forest concern", "Monitoring / control in PAs", "IBAMA regulations", "ICMBio / IBAMA", "Plantation escapee in Bahia/Pará Atlantic Forest margins; displaces native palms",
  "Jackfruit", "Artocarpus heterophyllus", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — Atlantic Forest", "Removal in protected areas", "ICMBio management plans", "ICMBio", "Dominant invasive in Tijuca NP (RJ) and other Atlantic Forest parks",
  "White ginger lily", "Hedychium coronarium", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — riparian zones", "Control mandated", "Lei 9.605/1998 / IBAMA regulations", "IBAMA / ICMBio", "Lírio-do-brejo; invades Atlantic Forest wetlands and riverbanks",
  "Ironwood she-oak", "Casuarina equisetifolia", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — coastal dunes", "Control in restinga ecosystems", "IBAMA regulations", "IBAMA / ICMBio", "Invasive on coastal dunes and restinga in northeast Brazil",
  "Yellow bells", "Tecoma stans", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — disturbed areas", "Control in reserves", "IBAMA / ICMBio management plans", "ICMBio / IBAMA", "Invasive in Cerrado/Caatinga margins; colonises degraded and urban areas",
  "Signal grass / Braquiária", "Urochloa decumbens", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — Cerrado/Amazon", "Control in protected areas", "IBAMA regulations", "ICMBio / IBAMA", "Pasture grass invading Cerrado and Amazon forest edges; fire promoter",
  "Brazilian pepper tree", "Schinus terebinthifolia", "Magnoliopsida", "Least Concern", "Not listed", "Native — invasive outside Brazil", "Export monitored under biosafety", "Lei 12.651/2012", "IBAMA", "Brazilian native; highly invasive in Florida/Caribbean/Hawaii"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","brazil_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>mutate(country="Brazil",region="",jurisdiction="NATIONAL",jurisdiction_group="",
    classification=taxonomic_group,taxonLevel="species",Note=notes,genus_I=word(prefName,1),is_aquatic="no",aquatic_type_I="")|>
    select(GBIFusageKey,country,region,jurisdiction,jurisdiction_group,prefName,classification,taxonLevel,family,englishName,synonyms,Note,genus_I,is_aquatic,aquatic_type_I,legal_status_br,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Brazil is the world's most biodiverse nation, harbouring approximately **50,000 plant species** — roughly **20% of all plant species on Earth**. The country contains six major biomes: Amazon, Atlantic Forest, Cerrado, Caatinga, Pantanal, and Pampa. The Atlantic Forest is one of the world's most threatened and species-rich biomes, with less than 12% of original cover remaining. Brazil's national threatened plant list — **Portaria MMA 443/2014** — covers 2,113 species across IUCN threat categories. The **Native Vegetation Protection Law (Lei 12.651/2012)** — the Forest Code — governs vegetation protection on private land and defines Permanent Preservation Areas (APPs) and Legal Reserve requirements.

<div class="alert alert-info" role="alert">
This report covers **`r n_species` legally restricted plant species** listed under Portaria MMA 443/2014 (National List of Threatened Plant Species). Legislation is primarily in Portuguese; the REFLORA database ([reflora.jbrj.gov.br](http://reflora.jbrj.gov.br)) provides an English-language taxonomic interface.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Portaria MMA 443/2014 | 2014 | National List of Threatened Plant Species — 2,113 species in CR/EN/VU/NT/DD categories |
| Lei 12.651/2012 — Native Vegetation Protection Law (Forest Code) | 2012 | Permanent Preservation Areas (APPs); Legal Reserve (RL) requirements; vegetation protection on private land |
| Lei 13.123/2015 — Biodiversity Law | 2015 | Access and benefit sharing (ABS) for genetic resources; replaces MP 2.186-16/2001 |
| Lei 9.605/1998 — Environmental Crimes Law | 1998 | Criminal penalties for flora crimes; illegal deforestation; invasive species non-management |
| Sistema Nacional de Unidades de Conservação (SNUC) — Lei 9.985/2000 | 2000 | National protected areas system; 12 categories of conservation units |

Full text: [ICMBio](https://www.icmbio.gov.br) | [MMA](https://www.gov.br/mma) | [IBAMA](https://www.ibama.gov.br) | [FAOLEX Brazil](https://www.fao.org/faolex/country-profiles/general-profile/en/?iso3=BRA)

---

# Regional / Subnational Restrictions {#regional}

Brazil's federated system grants states (and the Federal District) significant environmental authority. States may enact their own threatened species lists, expand protected categories, and impose stricter harvest or trade controls than federal law — they may not be *weaker* than federal law. The most significant state instruments for plant species are summarised below.

| State | Key Instrument | Plants Scope | Enforcing Body |
|-------|---------------|-------------|----------------|
| **São Paulo** | Resolução SMA 57/2016 (*Lista de Espécies da Flora Ameaçadas*) | 314 threatened plant species listed; stricter than Portaria 443/2014 for many taxa; collection and commercialisation prohibited without CETESB/SMA permit | SEMA-SP (Secretaria de Meio Ambiente) / CETESB |
| **Minas Gerais** | Deliberação COPAM 367/2008 (*Lista da Flora Ameaçada de Extinção*) | 406 threatened plants listed; Cerrado species especially; major overlap with Atlantic Forest and Cerrado biomes | SUPRAM / SEMAD |
| **Paraná** | Decreto Estadual 3148/2004 (*Lista Vermelha da Flora Ameaçada — PR*) | ~236 species; araucaria-mixed forest species; forestry licensing conditioned on rare plant surveys | IAP (Instituto Ambiental do Paraná) |
| **Santa Catarina** | Resolução CONSEMA 51/2008 | 160+ threatened plant species; araucaria and Atlantic Forest taxa | FATMA (Fundação do Meio Ambiente) |
| **Rio Grande do Sul** | Decreto 52.109/2014 (*Lista de Espécies da Flora Nativa Ameaçadas*) | 300+ taxa listed; grassland (campo) species notably included; native grassland conversion restricted | SEMA-RS / FEPAM |
| **Amazonas** | Lei Estadual 3135/2007 (*Política Estadual sobre Mudanças Climáticas*) + Decreto 24.375/2004 (*SEUC-AM*) | State conservation units may restrict plant collection beyond federal rules; ABS provisions apply in state PAs | SDS Amazonas (now SEMA-AM) |
| **Pará** | SEMA-PA management plans for state conservation units | Timber and NTFP harvest restrictions within state PAs; enforcement of Forest Code APP and RL in the Amazon | SEMA-Pará / IDEFLOR-Bio |
| **Bahia** | IMA-BA (Instituto do Meio Ambiente e Recursos Hídricos) | Atlantic Forest remnants; Caatinga endemic plants; licensing conditioning | IMA Bahia |

## Key State-Level Mechanisms

1. **State threatened plant lists** — São Paulo, Minas Gerais, Paraná, Santa Catarina, and Rio Grande do Sul each maintain official state lists that add hundreds of species not covered by the federal Portaria MMA 443/2014. Commercial use, collection, and trade of listed species require state environmental permits.

2. **State Conservation Units (SEUC)** — Each state has its own system of conservation units (parks, ecological stations, APA areas) with individual management plans that may impose species-specific collection bans. Amazonian states (Amazônas, Pará, Mato Grosso) have extensive state PA networks.

3. **CONAMA Resolutions** — National-level (federal) but implemented regionally: CONAMA Resolution 369/2006 governs intervention in Permanent Preservation Areas (APPs); states add site-specific restrictions. CONAMA Resolution 423/2010 addresses Atlantic Forest species and native grassland.

4. **Forest Code compliance conditionality** — State environmental agencies (e.g., CETESB in SP, FEPAM in RS) enforce CAR (Cadastro Ambiental Rural) registration, a precondition for land use. Rural properties with irregular CAR status face restrictions on timber and NTFP harvest from native vegetation.

<div class="alert alert-info" role="alert">
**Source note** — State threatened plant lists are enacted by state secretariat resolutions or COPAM deliberations and may be updated independently of federal lists. Always verify the current state list with the relevant state environmental secretariat. Key portal: [Sistema Nacional de Informação sobre Meio Ambiente (SINIMA)](https://www.gov.br/mma/pt-br/acesso-a-informacao/sistemas-e-ferramentas/sinima) | [INMA — Instituto Nacional da Mata Atlântica](https://www.inma.gov.br).
</div>

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| ICMBio (Chico Mendes Institute for Biodiversity Conservation) | Protected areas management; threatened species action plans; CITES management authority |
| IBAMA (Brazilian Institute of Environment and Renewable Natural Resources) | Environmental licensing; enforcement; invasive species; timber trade monitoring |
| MMA (Ministry of Environment and Climate Change) | Biodiversity policy; Forest Code; threatened species listing; Portaria 443 |
| Jardim Botânico do Rio de Janeiro (JBRJ) | National herbarium; REFLORA database; plant taxonomy; Red List assessments |
| INPA (National Institute for Amazonian Research) | Amazonian flora research; botanical surveys |
| EMBRAPA | Agricultural biodiversity; crop wild relative conservation; genetic resources |

---

# Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Brazil Status`=legal_status_br,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Protected Plant Species in Brazil",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Brazil Status`=coalesce(legal_status_br,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Brazil — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Complete Species Table (Template Format) {#complete-template}

The table below presents all species in the **18-column `regulated_plants_v9.7` template format** — the canonical client deliverable schema. This is the same data as the downloadable CSV.

```{r complete-template-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    styled_kable(
      caption = "Complete Species Listing — 18-Column Template Format (GBIFusageKey, country, region, jurisdiction, jurisdiction_group, prefName, classification, taxonLevel, family, englishName, synonyms, Note, genus_I, is_aquatic, aquatic_type_I, legal_status, restriction_type, governing_law)",
      height = "650px"
    )
}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Amazon deforestation — 11,568 km² lost in 2022 (INPE); enforcement against illegal clearing inadequate | High |
| 2 | Forest Code (Lei 12.651/2012) 2012 amnesty provisions reduce legal liability for pre-2008 deforestation | High |
| 3 | Portaria 443/2014 covers 2,113 species but enforcement capacity insufficient relative to Atlantic Forest area | High |
| 4 | Araucaria angustifolia (CR) — illegal logging for timber continues in southern Brazil buffer zones | High |
| 5 | Lei 13.123/2015 ABS implementation for wild plant genetic resources incomplete — many sectors non-compliant | Medium |
| 6 | Invasive Pinus and Leucaena removal in protected areas severely underfunded | Medium |

---

# References {#references}

| Resource | Link |
|----------|------|
| ICMBio | [icmbio.gov.br](https://www.icmbio.gov.br) |
| IBAMA | [ibama.gov.br](https://www.ibama.gov.br) |
| JBRJ — REFLORA Database | [reflora.jbrj.gov.br](http://reflora.jbrj.gov.br) |
| Flora do Brasil 2020 | [floradobrasil.jbrj.gov.br](http://floradobrasil.jbrj.gov.br) |
| GBIF Brazil | [gbif.org/country/BR](https://www.gbif.org/country/BR/summary) |
| CITES Brazil | [cites.org — Brazil](https://cites.org/eng/cms/index.php/component/cp/country/BR) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Verify at ICMBio / IBAMA / JBRJ / CITES</p></div>'))
```
