---
title: "Restricted and Protected Plant Species in Malaysia"
subtitle: "Wildlife Conservation Act 2010, National Forestry Act 1984, and CITES — Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,                 ~taxonomic_group, ~iucn_status,            ~cites_appendix, ~legal_status_my,                       ~restriction_type,             ~governing_law,                ~managing_institution,    ~notes,
  # --- WCA 2010 First Schedule: Totally Protected Plants (Peninsular Malaysia) ---
  "Rafflesia (Peninsular)",  "Rafflesia kerrii",               "Magnoliopsida",  "Critically Endangered", "Not listed",    "Totally Protected (WCA 2010)",         "Collection/trade prohibited",  "WCA 2010 First Schedule",     "PERHILITAN",             "Malaysia endemic Rafflesia; Peninsular Malaysia; flower collapses within days",
  "Rafflesia (Sabah)",       "Rafflesia keithii",              "Magnoliopsida",  "Endangered",            "Not listed",    "Totally Protected (Sabah WCE 1997)",   "Collection/trade prohibited",  "Sabah WCE 1997 Schedule 1",   "Sabah Wildlife Dept",    "Sabah endemic; Cabinet Order 2004; Poring Hot Springs population monitored",
  "Rafflesia (Sabah)",       "Rafflesia pricei",               "Magnoliopsida",  "Endangered",            "Not listed",    "Totally Protected (Sabah WCE 1997)",   "Collection/trade prohibited",  "Sabah WCE 1997 Schedule 1",   "Sabah Wildlife Dept",    "Sabah endemic; restricted to Crocker Range; Cabinet Order protected",
  "Rafflesia (Sabah)",       "Rafflesia tengku-adlinii",       "Magnoliopsida",  "Critically Endangered", "Not listed",    "Totally Protected (Sabah WCE 1997)",   "Collection/trade prohibited",  "Sabah WCE 1997 Schedule 1",   "Sabah Wildlife Dept",    "Discovered 2005; extremely limited range in Sabah highlands",
  "Rafflesia (Sarawak)",     "Rafflesia tuan-mudae",           "Magnoliopsida",  "Endangered",            "Not listed",    "Totally Protected (Sarawak WLPO 1998)","Collection/trade prohibited",  "Sarawak WLPO 1998 Sched 1",   "Sarawak Forestry Corp",  "Sarawak largest-flowered Rafflesia; Gunung Gading NP flagship species",
  # --- CITES Appendix I: Nepenthes (pitcher plants) ---
  "Rajah pitcher plant",     "Nepenthes rajah",                "Magnoliopsida",  "Vulnerable",            "Appendix I",    "Totally Protected (WCA 2010)",         "Collection/trade prohibited",  "WCA 2010 / CITES",            "Sabah Parks",            "Largest pitcher plant; Sabah Mt Kinabalu endemic; CITES Appendix I",
  "Kinabalu pitcher plant",  "Nepenthes villosa",              "Magnoliopsida",  "Vulnerable",            "Appendix II",   "Totally Protected (WCA 2010)",         "Collection/trade prohibited",  "WCA 2010 / CITES",            "Sabah Parks",            "Mt Kinabalu cloud forest endemic; upper montane specialist above 2300m",
  "Lowii pitcher plant",     "Nepenthes lowii",                "Magnoliopsida",  "Least Concern",         "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs / WCA 2010",       "PERHILITAN / Sabah Parks","Borneo highland pitcher plant; mutualism with tree shrews",
  "Macfarlanei pitcher plant","Nepenthes macfarlanei",          "Magnoliopsida",  "Vulnerable",            "Appendix II",   "Protected (WCA 2010)",                 "Collection permit required",   "WCA 2010 / CITES",            "PERHILITAN",             "Peninsular Malaysia endemic; Cameron Highlands and Genting; overcollection threat",
  "Hookeriana pitcher plant", "Nepenthes hookeriana",           "Magnoliopsida",  "Least Concern",         "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs",                  "PERHILITAN",             "Natural hybrid; peat swamp forests; all Nepenthes CITES II",
  # --- CITES Appendix I: Paphiopedilum (slipper orchids) ---
  "Rothschild slipper orchid","Paphiopedilum rothschildianum", "Magnoliopsida",  "Critically Endangered", "Appendix I",    "Totally Protected (WCA 2010)",         "Collection/trade prohibited",  "WCA 2010 / CITES",            "Sabah Parks",            "Gold of Kinabalu; 1 specimen = thousands USD on black market; CITES I",
  "Sandal slipper orchid",   "Paphiopedilum sanderianum",      "Magnoliopsida",  "Endangered",            "Appendix I",    "Totally Protected (WCA 2010)",         "Collection/trade prohibited",  "WCA 2010 / CITES",            "Sarawak Forestry Corp",  "World longest petals (up to 90cm); Sarawak limestone; rediscovered 1978",
  "Hooker slipper orchid",   "Paphiopedilum hookerae",         "Magnoliopsida",  "Vulnerable",            "Appendix I",    "Totally Protected (WCA 2010)",         "Collection/trade prohibited",  "WCA 2010 / CITES",            "Sabah Wildlife Dept",    "Borneo endemic; limestone forest specialist; all Paphiopedilum CITES I",
  "Lowii slipper orchid",    "Paphiopedilum lowii",            "Magnoliopsida",  "Vulnerable",            "Appendix I",    "Totally Protected (WCA 2010)",         "Collection/trade prohibited",  "WCA 2010 / CITES",            "PERHILITAN",             "Peninsular Malaysia and Borneo; epiphytic slipper orchid; highly traded",
  # --- Other CITES-listed orchids (Appendix II — all Orchidaceae) ---
  "Pigeon orchid",           "Dendrobium crumenatum",          "Magnoliopsida",  "Least Concern",         "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs / WCA 2010",       "PERHILITAN",             "Common epiphyte; mass-flowering after temperature drops; all orchids CITES II",
  "Tiger orchid",            "Grammatophyllum speciosum",      "Magnoliopsida",  "Least Concern",         "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs / WCA 2010",       "PERHILITAN",             "World largest orchid; clumps up to 2 tonnes; lowland forests declining",
  "Bulbophyllum",            "Bulbophyllum lobbii",            "Magnoliopsida",  "Least Concern",         "Appendix II",   "Protected (CITES Appendix II)",        "Collection permit required",   "CITES Regs / WCA 2010",       "PERHILITAN",             "Largest orchid genus; Malaysia major source of traded orchids",
  "Vanda orchid",            "Vanda dearei",                   "Magnoliopsida",  "Vulnerable",            "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs / Sabah WCE",      "Sabah Wildlife Dept",    "Borneo endemic Vanda; fragrant yellow flowers; overcollected for horticulture",
  # --- CITES-listed: Aquilaria (agarwood) ---
  "Agarwood (karas)",        "Aquilaria malaccensis",          "Magnoliopsida",  "Vulnerable",            "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs / NFA 1984",       "PERHILITAN / MTIB",      "Gaharu; resinous heartwood extremely valuable; CITES II since 2005; illegal harvesting widespread",
  "Agarwood (Borneo)",       "Aquilaria beccariana",           "Magnoliopsida",  "Vulnerable",            "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs / NFA 1984",       "Sarawak Forestry Corp",  "Borneo agarwood species; peat swamp forests; all Aquilaria CITES II",
  # --- CITES-listed: Cycads ---
  "Cycad (Malaysian)",       "Cycas rumphii",                  "Cycadopsida",    "Least Concern",         "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs",                  "PERHILITAN",             "Coastal cycad; all Cycas CITES II (Appendix I for some species); ornamental trade",
  "Pectinata cycad",         "Cycas pectinata",                "Cycadopsida",    "Vulnerable",            "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs",                  "PERHILITAN",             "Limestone hill cycad; Peninsular Malaysia; habitat loss from quarrying",
  # --- Timber: Protected and export-restricted species ---
  "Ramin",                   "Gonystylus bancanus",            "Magnoliopsida",  "Vulnerable",            "Appendix II",   "Protected (CITES Appendix II)",        "Export permit required",       "CITES Regs / NFA 1984",       "PERHILITAN / SFC",       "Peat swamp forest timber; CITES Appendix II since 2004; illegal logging continues",
  "Borneo ironwood",         "Eusideroxylon zwageri",          "Magnoliopsida",  "Vulnerable",            "Not listed",    "Totally Protected Timber",             "Felling prohibited",           "National Forestry Act 1984",   "Forestry Dept / SFD",    "Belian; hardest tropical timber; harvesting ban in multiple states; poaching continues",
  "Meranti (light red)",     "Shorea leprosula",               "Magnoliopsida",  "Endangered",            "Not listed",    "Protected Forest Species",             "Harvest regulated",            "National Forestry Act 1984",   "Forestry Dept",          "Major commercial dipterocarp; IUCN Endangered; dominant canopy tree in lowland forest",
  "Meranti (Sarawak)",       "Shorea macrophylla",             "Magnoliopsida",  "Vulnerable",            "Not listed",    "Protected Forest Species",             "Harvest regulated",            "National Forestry Act 1984",   "Sarawak Forestry Corp",  "Engkabang; major dipterocarp; seed traded for illipe butter",
  "Keruing",                 "Dipterocarpus costulatus",       "Magnoliopsida",  "Critically Endangered", "Not listed",    "Protected Forest Species",             "Harvest regulated",            "National Forestry Act 1984",   "Forestry Dept",          "Keruing timber species; IUCN CR; lowland dipterocarp forest; oil from resin",
  "Chengal",                 "Neobalanocarpus heimii",         "Magnoliopsida",  "Endangered",            "Not listed",    "Protected Premium Timber",             "Felling restricted",           "National Forestry Act 1984",   "Forestry Dept",          "Most prized Malaysian timber; extremely slow-growing; harvest quotas strictly enforced",
  "Merbau",                  "Intsia palembanica",             "Magnoliopsida",  "Vulnerable",            "Not listed",    "Protected Forest Species",             "Harvest regulated",            "National Forestry Act 1984",   "Forestry Dept / SFC",    "Premium hardwood for furniture; heavily logged; Peninsular and Borneo",
  "Kapur",                   "Dryobalanops aromatica",         "Magnoliopsida",  "Endangered",            "Not listed",    "Protected Forest Species",             "Harvest regulated",            "National Forestry Act 1984",   "Forestry Dept",          "Camphor-producing dipterocarp; Peninsular Malaysia lowlands; severely depleted",
  "Kempas",                  "Koompassia malaccensis",         "Magnoliopsida",  "Least Concern",         "Not listed",    "Protected (Honey Tree)",               "Felling prohibited in some states","NFA 1984 / State rules",  "Forestry Dept",          "Tualang; tallest tropical tree; protected as honey tree for Apis dorsata colonies",
  # --- Malaysian endemics and rare species ---
  "Johannesteijsmannia palm","Johannesteijsmannia altifrons",  "Magnoliopsida",  "Vulnerable",            "Not listed",    "Protected (WCA 2010)",                 "Collection restricted",        "WCA 2010 / NFA 1984",         "PERHILITAN / Forestry",  "Diamond-shaped undivided leaf palm; Peninsular Malaysia understorey endemic",
  "Jelutong",                "Dyera costulata",                "Magnoliopsida",  "Least Concern",         "Not listed",    "Protected Forest Species",             "Harvest regulated",            "National Forestry Act 1984",   "Forestry Dept",          "Latex-producing tree; historically over-tapped; NFA regulated species",
  "Tongkat ali",             "Eurycoma longifolia",            "Magnoliopsida",  "Near Threatened",       "Not listed",    "Harvest regulated",                    "Collection permit required",   "NFA 1984 / State rules",      "Forestry Dept / FRIM",   "High-value medicinal plant; wild-harvesting pressure; FRIM conservation programme",
  # --- Sabah and Sarawak specific protections ---
  "Sabah Rafflesia",         "Rafflesia arnoldii",             "Magnoliopsida",  "Endangered",            "Not listed",    "Totally Protected (Sabah WCE 1997)",   "Collection/trade prohibited",  "Sabah WCE 1997 Schedule 1",   "Sabah Wildlife Dept",    "World largest single flower; Sabah populations; WCE Schedule 1",
  "Jade vine (Sabah)",       "Strongylodon macrobotrys",       "Magnoliopsida",  "Vulnerable",            "Not listed",    "Protected (Sabah WCE 1997)",           "Collection permit required",   "Sabah WCE 1997 Schedule 2",   "Sabah Wildlife Dept",    "Turquoise flowers; planted in Sabah gardens; wild populations declining",
  "Sarawak pitcher plant",   "Nepenthes northiana",            "Magnoliopsida",  "Endangered",            "Appendix II",   "Totally Protected (Sarawak WLPO 1998)","Collection/trade prohibited",  "Sarawak WLPO 1998 / CITES",   "Sarawak Forestry Corp",  "Sarawak limestone endemic; spectacular pitcher; overcollected historically",
  # --- Invasive alien species ---
  "Water hyacinth",          "Eichhornia crassipes",           "Magnoliopsida",  "Least Concern",         "Not listed",    "Invasive Species",                     "Control mandated",             "EQA 1974 / State rules",      "DOE / JPS",              "Major invasive in Malaysian reservoirs and rivers; waterway blockage",
  "Mikania vine",            "Mikania micrantha",              "Magnoliopsida",  "Least Concern",         "Not listed",    "Invasive Species",                     "Control in plantations",       "State Forestry Acts",          "DOF",                    "Mile-a-minute vine; major problem in oil palm and rubber plantations",
  "Giant mimosa",            "Mimosa pigra",                   "Magnoliopsida",  "Least Concern",         "Not listed",    "Invasive Species",                     "Control mandated",             "State Agriculture Dept rules", "DOA",                    "Thorny shrub invading wetlands; displaces native vegetation in Peninsular Malaysia",
  "Siam weed",               "Chromolaena odorata",            "Magnoliopsida",  "Least Concern",         "Not listed",    "Invasive Species",                     "Control in plantations",       "State Agriculture Dept rules", "DOA",                    "Aggressive weed in disturbed land; fire hazard in dry season; oil palm plantations"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","malaysia_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_my,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Malaysia hosts approximately **15,000 plant species** across Peninsular Malaysia, Sabah, and Sarawak, with some of the world's richest tropical forest biodiversity. The country is a global centre of Rafflesia, Nepenthes, orchid, and dipterocarp diversity.

<div class="alert alert-info" role="alert">
This report covers **`r n_species` legally restricted plant species** under WCA 2010 First and Second Schedules and National Forestry Act protected species lists.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law | Year | Key Provision |
|-----|------|---------------|
| Wildlife Conservation Act (Act 716) | 2010 | First Schedule (Totally Protected); Second Schedule (Protected); 10 year prison / RM100,000 fine |
| National Forestry Act (Act 313) | 1984 | Protected tree species; timber licensing; forest reserves |
| Environmental Quality Act (Act 127) | 1974 | Environmental framework; pollution and degradation controls |
| Sarawak Wildlife Protection Ordinance | 1998 | State-level wildlife protection (Sarawak jurisdiction) |
| Sabah Wildlife Conservation Enactment | 1997 | State-level wildlife protection (Sabah jurisdiction) |

Full text: [MyBIS — WCA 2010](https://www.mybis.gov.my/pb/675) | [FAOLEX](https://faolex.fao.org/docs/pdf/mal107883.pdf)

---

# Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Malaysia Status`=legal_status_my,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Protected Plant Species in Malaysia",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Malaysia Status`=coalesce(legal_status_my,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Malaysia — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Regional / Subnational Restrictions {#regional}

Malaysia's constitutional structure fundamentally divides environmental and forestry jurisdiction: **Peninsular Malaysia** (11 states + 2 Federal Territories) operates under federal legislation including the **Protection of Wildlife Act 1972** and **National Forestry Act 1984**, while **Sabah** and **Sarawak** on Borneo are constitutionally autonomous over forests and wildlife. Federal laws **do not apply** to Sabah and Sarawak for most natural resource matters — each has its own comprehensive legislation.

## Sabah — Wildlife Conservation Enactment 1997 (WCE) + Forestry Enactment

| Instrument | Year | Key Plant Provisions |
|-----------|------|---------------------|
| Wildlife Conservation Enactment 1997 (WCE) | 1997 (amended 2012, 2016) | Three-schedule protected species lists; Schedule 1 (totally protected) includes *Rafflesia* spp. — collection, trade, possession absolutely prohibited; Schedule 2 and 3 species require Director of Wildlife permits |
| Forest Enactment 1968 (revised 1992) | 1968/1992 | Sabah Forestry Department (SFD) administers all forest reserves; Class I (Protection Forests) exclude commercial harvest; timber species licenses issued by SFD only |
| Sabah Parks Enactment 1984 | 1984 | Kinabalu Park, Tunku Abdul Rahman Marine Park — no plant collection without Sabah Parks permit; *Nepenthes rajah* and all lowii-group pitcher plants fully protected within parks |
| *Rafflesia* Cabinet Orders | 2004, updated | *Rafflesia keithii*, *R. pricei*, *R. tengku-adlinii* — individually listed by Cabinet Order; trade, transport, damage, and collection a criminal offence under WCE |
| Sabah Biodiversity Enactment 2000 | 2000 | Access and benefit sharing — any commercial bioprospecting requires Director-General of Sabah Biodiversity Centre (SaBC) approval; plant genetic resource access specifically regulated |

## Sarawak — Wild Life Protection Ordinance 1998 (WLPO) + Forests Ordinance

| Instrument | Year | Key Plant Provisions |
|-----------|------|---------------------|
| Wild Life Protection Ordinance 1998 (WLPO) Cap. 26 | 1998 | Three-schedule protected species lists similar to Sabah; Schedule 1 totally protected — all *Rafflesia* spp. in Sarawak; Schedule 2 and 3 require Director permits |
| Forests Ordinance 2015 (replacing 1954 Ordinance) | 2015 | Sarawak Forestry Corporation (SFC) administers timber licensing; Protected Forest, Permanent Forest Estate, National Park classifications; harvest in Protected Forests prohibited |
| Sarawak Biodiversity Centre Ordinance 1998 | 1998 | Commercial access to Sarawak biological resources requires SBC Access Permit (AP) and benefit-sharing agreement; applies to all plant species in Sarawak |
| Wildlife Sanctuary / National Park Regulations | Ongoing | Gunung Mulu, Bako, Lambir Hills, Kubah NP — absolute prohibition on plant collection; orchid, pitcher plant, and *Rafflesia* populations in NPs strictly protected |

## Peninsular Malaysia — State Forestry Variation

Within Peninsular Malaysia, each of the 11 states has its own **State Forestry Department** operating under the National Forestry Act 1984. States manage Permanent Forest Estates (PFE) and may designate additional protection:

- **Pahang**: Cameron Highlands Forest Reserve — highest-elevation forest in Peninsular Malaysia; orchid collection enforcement by State Forestry Department
- **Kelantan / Terengganu**: Taman Negara buffer zone state forests; rattan and medicinal plant collection quotas
- **Selangor**: Klang Gates Quartz Ridge Nature Reserve (state gazette) — unique quartz ridge species restrictions
- **Johor**: Endau-Rompin State Park — no plant collection; *Livistona* palms and endemic species strictly protected within state park boundaries

<div class="alert alert-info" role="alert">
**Source note** — Sabah Biodiversity Centre: [sabahbiodiversity.org](https://www.sabahbiodiversity.org) | Sabah Forestry Department: [forest.sabah.gov.my](https://forest.sabah.gov.my) | Sarawak Forestry Corporation: [sarawakforestry.com](https://www.sarawakforestry.com) | Sarawak Biodiversity Centre: [sbc.org.my](https://www.sbc.org.my) | PERHILITAN (Peninsular): [wildlife.gov.my](https://www.wildlife.gov.my).
</div>

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Deforestation rate in Sarawak/Sabah — ongoing habitat loss | High |
| 2 | WCA Second Schedule under-utilised relative to species diversity | High |
| 3 | Nepenthes online trade via social media | High |
| 4 | Ramin (CITES App II) — illegal logging continues | High |
| 5 | Sarawak/Sabah state laws inconsistent with federal WCA | Medium |

---

# References {#references}

| Resource | Link |
|----------|------|
| WCA 2010 — MyBIS | [mybis.gov.my](https://www.mybis.gov.my/pb/675) |
| PERHILITAN | [wildlife.gov.my](https://www.wildlife.gov.my) |
| GBIF Malaysia | [gbif.org/country/MY](https://www.gbif.org/country/MY/summary) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Verify at PERHILITAN / CITES</p></div>'))
```
