---
title: "Restricted and Protected Plant Species in Thailand"
subtitle: "Plants Act B.E. 2518 (1975), Wild Animal and Plant Conservation and Protection Act B.E. 2562 (2019) — Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,            ~taxonomic_group, ~iucn_status,            ~cites_appendix,  ~legal_status_th,                              ~restriction_type,              ~governing_law,                    ~managing_institution,   ~notes,
  # --- DALBERGIA / ROSEWOOD (CITES highest concern) ---
  "Siamese rosewood",        "Dalbergia cochinchinensis",  "Magnoliopsida",  "Critically Endangered", "Appendix II",    "Conserved plant (Plants Act B.E. 2518) / CITES App. II","Collection/trade prohibited",  "Plants Act B.E. 2518 / CITES",    "DNP / RFD",             "Phayung; most trafficked timber species globally; >90% population decline; Mekong region endemic; primary target of rosewood syndicates",
  "Tamhieng rosewood",       "Dalbergia oliveri",          "Magnoliopsida",  "Endangered",            "Appendix II",    "Conserved plant (Plants Act B.E. 2518) / CITES App. II","Collection/trade prohibited",  "Plants Act B.E. 2518 / CITES",    "DNP / RFD",             "Chingchan; major rosewood timber alongside D. cochinchinensis; heavy illegal logging in Isan and border provinces",
  "Burma padauk",            "Pterocarpus macrocarpus",    "Magnoliopsida",  "Least Concern",         "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Pradu; high-value timber species; covered under 1989 blanket logging ban; substitution target when rosewood unavailable",
  "Indian rosewood",         "Dalbergia latifolia",        "Magnoliopsida",  "Vulnerable",            "Appendix II",    "CITES App. II regulated",                     "Export/import permit required","CITES / WACPPA B.E. 2562",         "DNP",                   "Plantation-grown specimens in Thailand; all Dalbergia spp. listed CITES App. II since CoP17 (2016); annotation #15",
  "Sissoo",                  "Dalbergia sissoo",           "Magnoliopsida",  "Least Concern",         "Appendix II",    "CITES App. II regulated",                     "Export/import permit required","CITES / WACPPA B.E. 2562",         "DNP",                   "Cultivated in northern Thailand; all Dalbergia spp. CITES App. II; plantation timber with export documentation required",
  # --- AGARWOOD (Aquilaria) ---
  "Agarwood",                "Aquilaria malaccensis",      "Magnoliopsida",  "Critically Endangered", "Appendix II",    "Conserved plant (Plants Act B.E. 2518) / CITES App. II","Collection/trade prohibited from wild","Plants Act B.E. 2518 / CITES","DNP / RFD",        "Mai kritsana; heartwood resin (oud) extremely valuable; wild harvest prohibited; plantation cultivation expanding in southern Thailand",
  "Second-class agarwood",   "Aquilaria crassna",          "Magnoliopsida",  "Critically Endangered", "Appendix II",    "Conserved plant (Plants Act B.E. 2518) / CITES App. II","Collection/trade prohibited from wild","Plants Act B.E. 2518 / CITES","DNP / RFD",        "Mai ha; principal Thai agarwood species; plantations in Trat, Chanthaburi; wild populations severely depleted",
  # --- ORCHIDS (Orchidaceae — all CITES App. II minimum) ---
  "Blue vanda orchid",       "Vanda coerulea",             "Magnoliopsida",  "Endangered",            "Appendix I",     "Conserved plant (Plants Act B.E. 2518) / CITES App. I","Collection/trade prohibited",  "Plants Act B.E. 2518 / CITES",    "DNP",                   "Royal blue orchid; Chiang Mai highlands; wild collection for horticulture devastated wild populations",
  "Lady's slipper orchid",   "Paphiopedilum bellatulum",   "Magnoliopsida",  "Endangered",            "Appendix I",     "Conserved plant (Plants Act B.E. 2518) / CITES App. I","Collection/trade prohibited",  "Plants Act B.E. 2518 / CITES",    "DNP",                   "All Paphiopedilum spp. CITES App. I; limestone karst endemic; northern Thailand; highly collected for orchid trade",
  "Charming lady's slipper", "Paphiopedilum charlesworthii","Magnoliopsida",  "Endangered",            "Appendix I",     "Conserved plant (Plants Act B.E. 2518) / CITES App. I","Collection/trade prohibited",  "Plants Act B.E. 2518 / CITES",    "DNP",                   "Myanmar-Thailand border endemic; CITES App. I; limestone habitats; highly prized in orchid trade; wild populations critically low",
  "Exul lady's slipper",     "Paphiopedilum exul",         "Magnoliopsida",  "Endangered",            "Appendix I",     "Conserved plant (Plants Act B.E. 2518) / CITES App. I","Collection/trade prohibited",  "Plants Act B.E. 2518 / CITES",    "DNP",                   "Thai endemic Paphiopedilum; Peninsular Thailand; restricted limestone range; CITES App. I; conservation breeding at Queen Sirikit Botanic Garden",
  "Necklace orchid",         "Dendrobium lindleyi",        "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II / Plants Act B.E. 2518)","Collection/trade regulated",   "CITES / Plants Act B.E. 2518",    "DNP",                   "Ueang kham; all Orchidaceae CITES App. II; widespread epiphyte; mass-collected from forests for Songkran decorations",
  "Black orchid",            "Coelogyne pandurata",        "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II)",                   "Collection/trade regulated",   "CITES / Plants Act B.E. 2518",    "DNP",                   "Spectacular Borneo-Thai orchid; black lip; epiphytic; all CITES App. II orchids regulated under Plants Act",
  "Foxtail orchid",          "Rhynchostylis gigantea",     "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II / Plants Act B.E. 2518)","Collection/trade regulated",   "CITES / Plants Act B.E. 2518",    "DNP",                   "Chang kra; Thailand's provincial flower of Chiang Rai; heavily collected; all orchids CITES App. II",
  "Tiger orchid",            "Grammatophyllum speciosum",   "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II / Plants Act B.E. 2518)","Collection/trade regulated",   "CITES / Plants Act B.E. 2518",    "DNP",                   "World's largest orchid species; southern Thailand; CITES App. II; habitat loss from deforestation",
  # --- CYCADS (all CITES-listed) ---
  "Thai sago cycad",         "Cycas siamensis",            "Cycadopsida",    "Near Threatened",       "Appendix II",    "Protected (CITES App. II / Plants Act B.E. 2518)","Collection/trade regulated",   "CITES / Plants Act B.E. 2518",    "DNP",                   "Prong pa; all Cycas spp. CITES App. II; slow-growing; over-collected for ornamental trade; limestone karsts in Kanchanaburi",
  "Pectinate cycad",         "Cycas pectinata",            "Cycadopsida",    "Vulnerable",            "Appendix II",    "Protected (CITES App. II / Plants Act B.E. 2518)","Collection/trade regulated",   "CITES / Plants Act B.E. 2518",    "DNP",                   "Northern Thailand montane forests; CITES App. II; declining due to habitat loss and collection",
  # --- CARNIVOROUS / PITCHER PLANTS ---
  "Nepenthes (Mekong)",      "Nepenthes mirabilis",        "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (WACPPA B.E. 2562 / CITES App. II)","Collection/trade regulated",   "WACPPA B.E. 2562 / CITES",        "DNP",                   "Widespread tropical pitcher plant; Thailand populations on CITES Appendix II; popular in carnivorous plant trade",
  "Nepenthes (Thai)",        "Nepenthes smilesii",         "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (WACPPA B.E. 2562 / CITES App. II)","Collection/trade regulated",   "WACPPA B.E. 2562 / CITES",        "DNP",                   "Thai endemic pitcher plant; Isan region peat swamps; all Nepenthes CITES App. II; habitat threatened by agriculture",
  # --- PARASITIC / ICONIC SPECIES ---
  "Rafflesia",               "Rafflesia kerrii",           "Magnoliopsida",  "Critically Endangered", "Not listed",     "Conserved plant (Plants Act B.E. 2518)",       "Collection/disturbance prohibited","Plants Act B.E. 2518",          "DNP",                   "World's largest individual flower; Khao Sok NP; parasitic on Tetrastigma vines; highly threatened; ecotourism icon",
  # --- TIMBER SPECIES (Logging Ban 1989) ---
  "Teak",                    "Tectona grandis",            "Magnoliopsida",  "Least Concern",         "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Royal Forest Department ban on all commercial logging 1989; teak still illegally harvested near Myanmar border; plantation teak export permitted",
  "Ironwood",                "Hopea odorata",              "Magnoliopsida",  "Vulnerable",            "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Takian thong; sacred tree in Thai culture; Dipterocarpaceae; logging ban 1989; specimen trees protected at temple sites",
  "Yang na",                 "Dipterocarpus alatus",       "Magnoliopsida",  "Endangered",            "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Yang na; provincial tree of Kanjanaburi; dominant canopy dipterocarp; oleo-resin historically harvested; logging ban 1989",
  "Keruing",                 "Dipterocarpus costatus",     "Magnoliopsida",  "Endangered",            "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Yang khon; important timber dipterocarp; eastern and southern Thailand forests; protected under blanket logging ban",
  "Makha mong",              "Afzelia xylocarpa",          "Magnoliopsida",  "Endangered",            "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Makha; premium hardwood; deciduous forest species; Isan and central Thailand; protected under 1989 ban; illegal logging persists",
  "Siamese sal",             "Shorea obtusa",              "Magnoliopsida",  "Least Concern",         "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Rang; dominant dry dipterocarp forest species; northern and northeastern Thailand; timber ban 1989",
  "White meranti",           "Shorea roxburghii",          "Magnoliopsida",  "Least Concern",         "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Phayom; flowers used in Buddhist ceremonies; dry evergreen and mixed deciduous forests; logging ban 1989",
  # --- PLANTS ACT CONSERVED LIST / NATIONAL PARK ACT ---
  "Golden chicken fern",     "Cibotium barometz",          "Polypodiopsida", "Least Concern",         "Appendix II",    "Conserved plant (Plants Act B.E. 2518) / CITES App. II","Collection prohibited",        "Plants Act B.E. 2518 / CITES",    "DNP",                   "Khut kra jok; tree fern; rhizome traded in traditional medicine; CITES App. II; northern Thailand montane forests",
  "Siam tulip",              "Curcuma alismatifolia",      "Magnoliopsida",  "Least Concern",         "Not listed",     "Permitted commercial export — regulated",     "Trade regulated (export permit)","Plants Act B.E. 2518 — prohibited export list","DOA",      "Thailand's major commercial flower export; wild populations protected; commercial cultivation permitted; Chaiyaphum provincial flower",
  "Kratom",                  "Mitragyna speciosa",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Decriminalised 2021; regulated cultivation",  "Cultivation/sale regulated",   "Kratom Act B.E. 2565 (2022)",      "FDA / DOA",             "Formerly Narcotics Act Schedule V; decriminalised Aug 2021; Kratom Act 2022 regulates commercial production and age restrictions",
  "Opium poppy",             "Papaver somniferum",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Prohibited (Narcotics Act B.E. 2522)",        "Cultivation/possession prohibited","Narcotics Act B.E. 2522 (1979)","ONCB / FDA",           "Schedule IV narcotic; highland eradication programme; Royal Project substitution crops; northern border provinces",
  "Sacred lotus",            "Nelumbo nucifera",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected in national parks (NP Act B.E. 2562)","Collection prohibited in NPs", "National Park Act B.E. 2562",      "DNP",                   "Bua luang; culturally significant; protected within national park boundaries; Thale Noi wetlands Ramsar site",
  "Sacred fig",              "Ficus religiosa",            "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (heritage tree / NP Act B.E. 2562)","Felling prohibited at heritage sites","NP Act B.E. 2562 / Ancient Monuments Act","DNP / Fine Arts Dept","Ton pho; Buddhist significance; protected at temple and heritage sites under multiple laws; Ayutthaya and Sukhothai by-laws",
  # --- MANGROVE SPECIES ---
  "Mangrove apple",          "Sonneratia caseolaris",      "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (mangrove forest regulations)",     "Mangrove felling prohibited",  "Mangrove Management regulations / WACPPA","DNP",            "Keystone mangrove species; coastal fisheries habitat; mangrove loss in Thailand ~25% since 1990",
  "Nipa palm",               "Nypa fruticans",             "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (mangrove forest regulations)",     "Mangrove felling prohibited",  "Mangrove Management regulations / WACPPA","DNP / RFD",      "Chak; estuarine palm; Rayong and Trat mangrove zones; frond harvest permitted under community use plans only",
  "Tall-stilt mangrove",     "Rhizophora apiculata",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (mangrove forest regulations)",     "Mangrove felling prohibited",  "Mangrove Management regulations / WACPPA","DNP / RFD",      "Kong kang bai lek; dominant Thai mangrove; charcoal production historically; now protected; replanting programmes active",
  # --- THAILAND RED DATA BOOK / RARE SPECIES ---
  "Thai aloe",               "Aloe vera",                  "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II)",                   "Export permit required",       "CITES / WACPPA B.E. 2562",        "DNP",                   "All Aloe spp. CITES App. II except A. vera (exempt from annotation); cultivated commercially; wild Aloe not native to Thailand",
  "Siamese crocodile tree",  "Dracaena cochinchinensis",   "Magnoliopsida",  "Not Evaluated",         "Not listed",     "Preserved plant — logging banned (1989)",     "All logging prohibited in natural forests","Logging Ban Policy 1989 / Forest Act","RFD",         "Chan pha; dragon's blood resin; eastern Thailand; rare in wild; Red Data list; sometimes confused with D. draco",
  "Elephant apple",          "Dillenia indica",            "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected in national parks (NP Act B.E. 2562)","Collection prohibited in NPs", "National Park Act B.E. 2562",      "DNP",                   "Ma tad; large-fruited tree of riparian forests; NP-protected; fruits used in Thai cuisine; habitat loss from dam construction",
  "Khoi paper tree",         "Streblus asper",             "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (Plants Act B.E. 2518 — controlled)","Export regulated",             "Plants Act B.E. 2518",             "DOA",                   "Khoi; bark used for traditional khoi paper (Thai manuscripts); controlled plant under Plants Act; cultural heritage value",
  # --- INVASIVE SPECIES (ONEP management) ---
  "Chromolaena",             "Chromolaena odorata",        "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — management required",              "Control mandated in reserves", "DNP management plans",             "DNP / ONEP",            "Siam weed; major invasive in Thai forest margins and buffer zones; displaces native ground flora rapidly; fire-prone",
  "Water hyacinth",          "Eichhornia crassipes",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — control required",                 "Control mandated",             "ONEP / RID regulations",           "ONEP / RID",            "Phak tob chawa; clogs Thai waterways and reservoirs; >100,000 ha affected; control programme ongoing",
  "Giant mimosa",            "Mimosa pigra",               "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — control required",                 "Control mandated in wetlands", "ONEP / DNP management plans",      "ONEP / DNP",            "Mai yara phon; IUCN 100 worst invasive species; invades wetlands and Mekong floodplains; thorny thickets displace native riparian flora",
  "Leucaena",                "Leucaena leucocephala",      "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — management required",              "Control in protected areas",   "ONEP management plans",            "ONEP",                  "Krathin; IUCN 100 worst invasives; nitrogen-fixer forming monocultures in disturbed forest; widespread in degraded Thai lowlands",
  "Mesquite",                "Prosopis juliflora",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — management required",              "Control in protected areas",   "ONEP management plans",            "ONEP",                  "Introduced for agroforestry; invasive in dry areas of Isan; forms impenetrable thickets; groundwater depletion concerns"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","thailand_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_th,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Thailand hosts approximately **11,000–12,000 plant species** within a diverse landscape spanning tropical rainforests, monsoon forests, mangroves, and montane cloud forests. The country is part of the Indo-Burma biodiversity hotspot. Thailand's plant protection regime operates under the **Plants Act B.E. 2518 (1975)**, which classifies plants into three categories — **conserved** (strictly protected), **controlled** (export/import regulated), and **prohibited** (import banned) — and the newer **Wild Animal and Plant Conservation and Protection Act B.E. 2562 (2019)**, which implements CITES obligations domestically.

<div class="alert alert-info" role="alert">
This report covers **`r n_species` legally restricted plant species** under Thailand's Plants Act B.E. 2518 conserved and controlled plant lists. The Plants Act list is maintained by the Department of Agriculture (DOA) with DNP coordinating CITES implementation.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Plants Act B.E. 2518 | 1975 (amended) | Three-tier classification: conserved (protected), controlled (permit), prohibited (import ban); DOA administration |
| Wild Animal and Plant Conservation and Protection Act B.E. 2562 | 2019 | CITES implementation; replaces Wildlife Enhancement and Conservation Act B.E. 2535; App. I/II/III obligations |
| National Forest Act B.E. 2484 | 1941 (amended) | Forest reserves; timber licensing; RFD management |
| Environmental Quality Promotion and Conservation Act B.E. 2535 | 1992 | EIA; environmental quality management; ONEP coordination |
| Logging Ban Policy | 1989 | Blanket ban on commercial logging in natural forests |

Full text: [DNP Thailand](https://www.dnp.go.th) | [DOA Thailand](https://www.doa.go.th) | [FAOLEX Thailand](https://www.fao.org/faolex/country-profiles/general-profile/en/?iso3=THA)

---

# Regional / Subnational Restrictions {#regional}

Thailand is a unitary state with 77 provinces administered by centrally-appointed governors, but the **Department of National Parks, Wildlife and Plant Conservation (DNP)** and **Royal Forest Department (RFD)** maintain regional offices that implement national legislation. Provincial and local administrative organisations (**PAO / Tambon Administrative Organisations — TAOs**) have limited environmental powers under the **Decentralisation Act 1999**. TAOs may designate local environment protection zones and enforce littering/clearing bans, but species-specific restrictions remain national.

## Regional Offices — DNP and RFD

Each of Thailand's 6 geographic regions has **DNP Regional Conservation Area Offices (RCAO)** and **RFD Regional Forest Offices**:

| Region | Key RCAOs | Notable Species Focus |
|--------|----------|----------------------|
| **Northern Region** (Chiang Mai, Chiang Rai) | RCAO 1 (Chiang Mai) | Teak (*Tectona grandis*) in Doi Inthanon, Doi Pui, and Chiang Dao National Parks; *Cibotium barometz* (golden chicken fern); opium poppy (*Papaver somniferum*) — Highland alternative crops programme replaces cultivation |
| **Northeastern Region** (Isan) | RCAO 7 (Udon Thani), RCAO 9 | Mukdahan RAMSAR site — aquatic plant restrictions; *Pterocarpus macrocarpus* (padauk) harvest restrictions; Phu Wiang National Park rare plant management |
| **Central Region** | RCAO 3, RCAO 4 | Khao Yai WH Site — strict research permit regime; orchid and tree fern collection prohibited; *Hopea odorata* and Dipterocarp species in protected forests |
| **Eastern Region** (Chonburi, Rayong) | RCAO 2 | Khao Chamao and Khao Wong NP — mangrove restoration zones; *Nypa fruticans* (mangrove palm) restricted; Rayong mangrove forest has special provincial conservation zone |
| **Western Region** (Kanchanaburi) | RCAO 3 (shared) | Thong Pha Phum National Park — extensive research permit requirements; Karen indigenous plant use in Umphang WS |
| **Southern Region** (Surat Thani to Songkhla) | RCAO 5, RCAO 6 | *Rafflesia kerrii* in Khao Sok NP — absolute no-collection zone; *Coelogyne* and *Dendrobium* orchids; *Aquilaria malaccensis* (agarwood) enforcement by RFD |

## Northern Highlands — Highland Community Forest Programme

Under the **Community Forest Act B.E. 2562 (2019)**, Thailand formalised community forest rights. Over 11,000 community forests have been registered with the RFD. In northern highland communities (Karen, Akha, Hmong, Lisu), **traditional resource use zones** within community forests may restrict or allow plant collection beyond national DNP regulations:

- Community forest management plans may designate **conservation zones** (no collection) and **use zones** (NTFP harvest permitted).
- The Northern Upland Area Development Centre (NUADC) coordinates with communities on medicinal plant access and conservation.
- Opium replacement crops (coffee, macadamia, cherry) are managed under royal project schemes — some highland *Camellia sinensis* (tea) collection from wild populations is regulated under Royal Project Foundation rules.

## Muang Boran / Historical Zones — Local Authority Tree Protection

Under the **Ancient Monuments, Antiques, Objects of Art and National Museums Act BE 2504**, heritage trees in the precincts of:
- **Sukhothai Historical Park** (UNESCO World Heritage Site): Any vegetation modification requires Fine Arts Department approval.
- **Ayutthaya Historical Park**: Sacred fig (*Ficus religiosa*) and other temple trees protected; TAO Ayutthaya enforces by-law with fines.

<div class="alert alert-info" role="alert">
**Source note** — DNP Thailand: [dnp.go.th](https://www.dnp.go.th) | RFD: [forest.go.th](https://www.forest.go.th) | Community Forest Act 2019: [FAOLEX Thailand](https://www.fao.org/faolex/country-profiles/general-profile/en/?iso3=THA) | Royal Project Foundation: [royalprojectthailand.com](https://www.royalprojectthailand.com).
</div>

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| DNP (Department of National Parks, Wildlife and Plant Conservation) | WACPPA administration; CITES management authority; national parks; conserved plants |
| DOF (Department of Forestry / Royal Forest Department, RFD) | Timber licensing; forest management; teak and timber species oversight |
| DOA (Department of Agriculture) | Plants Act administration; conserved/controlled plant lists; import permits |
| ONEP (Office of Natural Resources and Environment Policy and Planning) | Environmental policy; biodiversity strategy; NBSAP coordination |
| Queen Sirikit Botanic Garden | Ex-situ plant conservation; orchid collections; seed bank |
| Forest Herbarium (BKF) | National herbarium; plant taxonomy; species identification support |

---

# Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Thailand Status`=legal_status_th,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Protected Plant Species in Thailand",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Thailand Status`=coalesce(legal_status_th,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Thailand — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Siamese rosewood trafficking — one of world's most seized timber species; organised criminal syndicates involved | High |
| 2 | WACPPA 2019 enforcement capacity insufficient; border controls for plant trafficking inadequate | High |
| 3 | Orchid wild collection continues despite Plants Act; online trade via Line/Facebook unmonitored | High |
| 4 | Mangrove coastal forest loss ongoing from aquaculture and coastal development | High |
| 5 | Chromolaena odorata invasion in national park buffer zones inadequately resourced | Medium |
| 6 | Plants Act conserved plant list review overdue relative to current IUCN threat assessments | Medium |

---

# References {#references}

| Resource | Link |
|----------|------|
| DNP Thailand | [dnp.go.th](https://www.dnp.go.th) |
| Department of Agriculture (DOA) | [doa.go.th](https://www.doa.go.th) |
| Royal Forest Department (RFD) | [forest.go.th](https://www.forest.go.th) |
| GBIF Thailand | [gbif.org/country/TH](https://www.gbif.org/country/TH/summary) |
| CITES Thailand | [cites.org — Thailand](https://cites.org/eng/cms/index.php/component/cp/country/TH) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Verify at DNP / DOA Thailand / CITES</p></div>'))
```
