---
title: "Restricted and Protected Plant Species in Sri Lanka"
subtitle: "Fauna and Flora Protection Ordinance (Cap. 469, as amended 2022), Forest Ordinance No. 16 of 1907 — Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,            ~taxonomic_group, ~iucn_status,            ~cites_appendix,  ~legal_status_lk,                              ~restriction_type,              ~governing_law,                    ~managing_institution,   ~notes,

  # --- FFPO Cap. 469 Schedule V — Protected Plants ---
  "Nelu",                    "Strobilanthes sessilis",     "Magnoliopsida",  "Endangered",            "Not listed",     "Protected (FFPO Schedule V)",                 "Collection prohibited",        "FFPO Cap. 469 Schedule V",        "DWC",                   "Mass flowering every 12 years; Knuckles Range; critical bird habitat; mass fruiting drives rodent cycles",
  "Palk lily",               "Crinum zeylanicum",          "Magnoliopsida",  "Endangered",            "Not listed",     "Protected (FFPO Schedule V)",                 "Collection prohibited",        "FFPO Cap. 469 Schedule V",        "DWC",                   "Sri Lanka endemic; riverine and wet grassland margins; declining due to drainage and agriculture",
  "Sri Lanka blue water lily","Nymphaea nouchali",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (FFPO Schedule V)",                 "Collection prohibited",        "FFPO Cap. 469 Schedule V",        "DWC",                   "National flower of Sri Lanka (Nil Mahanel); protected under FFPO; culturally significant",
  "Wild jasmine",            "Jasminum angustifolium",     "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (FFPO Schedule V)",                 "Collection prohibited",        "FFPO Cap. 469 Schedule V",        "DWC",                   "Native wild jasmine; protected in national parks and sanctuaries; fragrant white flowers",
  "Agarwood",                "Gyrinops walla",             "Magnoliopsida",  "Critically Endangered", "Appendix II",    "Protected (FFPO Schedule V / CITES App. II)", "Collection/trade prohibited",  "FFPO Cap. 469 / CITES",           "DWC",                   "Sri Lanka endemic; illegally harvested for resinous heartwood (walla patta); severe poaching pressure; wet zone forests",

  # --- Forest Ordinance No. 16/1907 — Protected Timber Species ---
  "Ebony",                   "Diospyros ebenum",           "Magnoliopsida",  "Near Threatened",       "Not listed",     "Protected timber (Forest Ordinance)",         "Felling regulated",            "Forest Ordinance No. 16/1907",    "Forest Department",     "Ceylon ebony; highly prized dense black timber; historically over-exploited for export",
  "Ceylon ironwood",         "Mesua ferrea",               "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected forest species",                    "Felling regulated",            "Forest Ordinance No. 16/1907",    "Forest Department",     "Nagassam; national tree of Sri Lanka; dense durable timber; also sacred/temple plantings",
  "Satinwood",               "Chloroxylon swietenia",      "Magnoliopsida",  "Vulnerable",            "Not listed",     "Protected timber (Forest Ordinance)",         "Felling regulated",            "Forest Ordinance No. 16/1907",    "Forest Department",     "Burutha; premier cabinet timber of Sri Lanka; illegal logging persists; dry and intermediate zone forests",
  "Velvet tamarind",         "Dialium ovoideum",           "Magnoliopsida",  "Vulnerable",            "Not listed",     "Protected timber (Forest Ordinance)",         "Felling regulated",            "Forest Ordinance No. 16/1907",    "Forest Department",     "Sri Lanka endemic; lowland rainforest; valuable timber; restricted felling permit required",
  "Nedun",                   "Pericopsis mooniana",        "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected timber (Forest Ordinance / CITES)", "Felling/trade regulated",      "Forest Ordinance No. 16/1907 / CITES", "Forest Department",  "Highly valued timber species; CITES Appendix II since 2019; wet zone lowland forests; slow-growing",
  "Hora",                    "Dipterocarpus zeylanicus",   "Magnoliopsida",  "Vulnerable",            "Not listed",     "Protected timber (Forest Ordinance)",         "Felling regulated",            "Forest Ordinance No. 16/1907",    "Forest Department",     "Sri Lanka endemic dipterocarp; dominant canopy tree in lowland wet zone; oleoresin also harvested",
  "Halmilla",                "Berrya cordifolia",          "Magnoliopsida",  "Near Threatened",       "Not listed",     "Protected timber (Forest Ordinance)",         "Felling regulated",            "Forest Ordinance No. 16/1907",    "Forest Department",     "Trincomalee wood; valued boat-building timber; dry zone forests; over-exploited historically",
  "Alexandrian laurel",      "Calophyllum inophyllum",     "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected coastal/forest species",            "Coastal buffer zone controls", "Forest Ordinance / CEA Acts",     "Forest Dept / CEA",     "Coastal and riparian tree; important for soil stability; traditional medicine uses",
  "Teak",                    "Tectona grandis",            "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected timber (Forest Ordinance)",         "Felling regulated",            "Forest Ordinance No. 16/1907",    "Forest Department",     "Introduced plantation timber; all state-owned teak felling requires FD permit; major revenue species",
  "Jak",                     "Artocarpus heterophyllus",   "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage tree designations",                  "Felling restricted (heritage)","Forest Ordinance / Local Auth.",   "Forest Dept / LAs",     "National fruit tree; large mature specimens protected under heritage tree by-laws in several districts",

  # --- CITES-Listed Sri Lankan Plants — Orchids, Cycads, Aquilaria ---
  "Sri Lanka orchid",        "Dendrobium macraei",         "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected (FFPO Schedule V / CITES App. II)", "Collection/trade regulated",   "FFPO Cap. 469 / CITES",           "DWC",                   "Endemic epiphytic orchid; montane cloud forests; prized in horticulture",
  "Ceylon pitcher plant",    "Nepenthes distillatoria",    "Magnoliopsida",  "Endangered",            "Appendix II",    "Strictly Protected (FFPO / CITES App. II)",   "Collection/trade prohibited",  "FFPO Cap. 469 / CITES",           "DWC",                   "Only Nepenthes species native to Sri Lanka; wet zone lowland bogs; severe habitat loss",
  "Lanka bulbophyllum",      "Bulbophyllum elegantulum",   "Magnoliopsida",  "Endangered",            "Appendix II",    "Protected (FFPO / CITES App. II)",            "Collection/trade regulated",   "FFPO Cap. 469 / CITES",           "DWC",                   "Miniature endemic orchid; restricted to Knuckles and Peak Wilderness; epiphyte on mossy branches",
  "Foxtail orchid",          "Rhynchostylis retusa",       "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected (CITES App. II)",                   "Trade regulated",              "CITES / FFPO Cap. 469",           "DWC",                   "Pendulous inflorescence orchid; collected for ornamental use; lowland and intermediate zone forests",
  "Blue vanda",              "Vanda tessellata",           "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected (CITES App. II)",                   "Trade regulated",              "CITES / FFPO Cap. 469",           "DWC",                   "Vandaceous orchid with tessellated flowers; traditional Ayurvedic use; dry zone scrublands",
  "Ceylon cycad",            "Cycas nathorstii",           "Cycadopsida",    "Endangered",            "Appendix II",    "Protected (FFPO / CITES App. II)",            "Collection/trade prohibited",  "FFPO Cap. 469 / CITES",           "DWC",                   "Sri Lanka endemic cycad; restricted to dry zone coastal scrub; fewer than 2,500 mature individuals",

  # --- Sri Lanka Red List 2012 — Threatened Endemic Species ---
  "Doona",                   "Shorea cordifolia",          "Magnoliopsida",  "Critically Endangered", "Not listed",     "Protected (Red List / FFPO)",                 "Collection prohibited",        "FFPO Cap. 469 / Forest Ordinance","DWC / Forest Dept",     "Sri Lanka endemic dipterocarp; confined to Sinharaja and adjacent fragments; canopy emergent",
  "Sri Lanka rose",          "Argyreia populifolia",       "Magnoliopsida",  "Critically Endangered", "Not listed",     "Protected (Red List 2012)",                   "Collection prohibited",        "FFPO Cap. 469",                   "DWC",                   "Endemic climber; restricted to a few dry zone localities; fewer than 50 mature plants known",
  "Mal-beraliya",            "Stemonoporus moonii",        "Magnoliopsida",  "Endangered",            "Not listed",     "Protected (Red List / Forest Ordinance)",     "Felling/collection prohibited","Forest Ordinance No. 16/1907",    "Forest Department",     "Sri Lanka endemic; Sinharaja canopy tree; genus Stemonoporus entirely endemic to Sri Lanka (15 spp.)",
  "Syzygium makul",          "Syzygium makul",             "Magnoliopsida",  "Critically Endangered", "Not listed",     "Protected (Red List 2012)",                   "Collection prohibited",        "FFPO Cap. 469",                   "DWC",                   "Sri Lanka endemic; known from fewer than 5 locations in wet zone; fruit tree of conservation concern",
  "Kekiriwara",              "Schumacheria castaneifolia", "Magnoliopsida",  "Vulnerable",            "Not listed",     "Protected (Red List 2012)",                   "Collection prohibited",        "FFPO Cap. 469",                   "DWC",                   "Sri Lanka endemic shrub; genus Schumacheria endemic to Sri Lanka; wet zone understorey; 3 species in genus",
  "Wallapatta",              "Gyrinops walla",             "Magnoliopsida",  "Critically Endangered", "Appendix II",    "Protected (Red List / FFPO / CITES)",         "Collection/trade prohibited",  "FFPO Cap. 469 / CITES",           "DWC",                   "See also Agarwood above; Red List CR due to illegal harvesting for agarwood resin; wet zone endemic",
  "Hora-mee",                "Dipterocarpus hispidus",     "Magnoliopsida",  "Endangered",            "Not listed",     "Protected (Red List / Forest Ordinance)",     "Felling prohibited",           "Forest Ordinance No. 16/1907",    "Forest Department",     "Sri Lanka endemic dipterocarp; lowland wet zone only; large buttressed tree; habitat severely fragmented",

  # --- Invasive Alien Species ---
  "Water hyacinth",          "Eichhornia crassipes",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — management required",              "Control mandated",             "CEA regulations",                 "CEA / Irrigation Dept", "Blocks irrigation channels and reservoirs; major problem in Mahaweli and north-central tanks",
  "Prickly pear",            "Opuntia dillenii",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive species",                            "Control measures active",      "CEA / Forest Dept regulations",   "Forest Dept / CEA",     "Invasive in dry zone scrublands and coastal dunes; displaces native vegetation",
  "Lantana",                 "Lantana camara",             "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive species — control required",         "Eradication programmes",       "CEA / Forest Dept regulations",   "Forest Dept / CEA",     "Aggressive invasive shrub; colonises degraded forests, grasslands, and plantation edges across all zones",
  "Mile-a-minute vine",      "Mikania micrantha",          "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive species — control required",         "Eradication programmes",       "CEA regulations",                 "CEA / Forest Dept",     "Fast-growing Neotropical vine; smothers native vegetation in wet zone; major threat to forest regeneration",
  "Mesquite",                "Prosopis juliflora",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive species — control required",         "Eradication/management",       "CEA / Forest Dept regulations",   "Forest Dept / CEA",     "Invasive in dry zone; Hambantota, Puttalam coasts; displaces native scrub; thorny thickets impede access",
  "Parthenium weed",         "Parthenium hysterophorus",   "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive species — control required",         "Control measures active",      "CEA regulations",                 "CEA / Dept Agriculture","Allergenic invasive; spreading in dry zone agricultural lands; reduces crop yields; health hazard",

  # --- Endemic Medicinal Plants under Traditional Medicine Regulations ---
  "Weniwelgeta",             "Coscinium fenestratum",      "Magnoliopsida",  "Vulnerable",            "Not listed",     "Protected medicinal plant",                   "Collection regulated",         "FFPO Cap. 469 / Ayurveda Act",    "DWC / Dept Ayurveda",   "Yellow vine; highly valued in traditional medicine (tree turmeric); over-harvested from wild; wet zone climber",
  "Bin-kohomba",             "Munronia pinnata",           "Magnoliopsida",  "Vulnerable",            "Not listed",     "Protected medicinal plant",                   "Collection regulated",         "FFPO Cap. 469 / Ayurveda Act",    "DWC / Dept Ayurveda",   "Small endemic herb; key ingredient in traditional fever remedies; over-collected; cultivation efforts underway",
  "Iramusu",                 "Hemidesmus indicus",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Regulated medicinal plant",                   "Collection regulated",         "Ayurveda Act / Forest Ordinance", "Dept Ayurveda / FD",    "Widely used in Ayurvedic medicine (blood purifier); roots over-harvested; collection permits required in reserves"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","sri_lanka_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>mutate(country="Sri Lanka",region="",jurisdiction="NATIONAL",jurisdiction_group="",
    classification=taxonomic_group,taxonLevel="species",Note=notes,genus_I=word(prefName,1),is_aquatic="no",aquatic_type_I="")|>
    select(GBIFusageKey,country,region,jurisdiction,jurisdiction_group,prefName,classification,taxonLevel,family,englishName,synonyms,Note,genus_I,is_aquatic,aquatic_type_I,legal_status_lk,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Sri Lanka is a global **biodiversity hotspot** with approximately **3,200 native plant species**, of which roughly **26% are endemic** — an extraordinary proportion for an island of its size (65,610 km²). The wet zone rainforests of the southwestern lowlands and the montane forests of the central highlands harbour the greatest concentration of endemic plants. Sri Lanka's primary instrument for plant protection is the **Fauna and Flora Protection Ordinance (FFPO, Cap. 469)**, which lists protected species in its schedules and establishes the framework for the country's network of national parks and sanctuaries.

<div class="alert alert-info" role="alert">
This report covers **`r n_species` legally restricted plant species** listed under FFPO Schedule V (protected plants). DWC ([dwc.gov.lk](https://www.dwc.gov.lk)) maintains the authoritative species schedules.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Fauna and Flora Protection Ordinance (FFPO, Cap. 469) | 1937 (amended 2009, 2022) | Schedule V — protected plants; prohibits collection, possession, trade without permit; strict penalties |
| Forest Ordinance No. 16 | 1907 (amended) | Forest reserves; protected timber species; felling permits |
| National Environmental Act No. 47 | 1980 (amended) | CEA establishment; environmental impact assessments; pollution control |
| Botanic Gardens Ordinance | 1928 | National botanic gardens as ex-situ conservation sites |
| CITES Implementation | 1979 (accession) | Wildlife Trade Regulations; DWC as Management Authority |

Full text: [DWC — dwc.gov.lk](https://www.dwc.gov.lk) | [FAOLEX Sri Lanka](https://www.fao.org/faolex/country-profiles/general-profile/en/?iso3=LKA)

---

# Regional / Subnational Restrictions {#regional}

Sri Lanka is a unitary state with 9 provinces and 25 districts, but environmental legislation is primarily national, administered by the **Department of Wildlife Conservation (DWC)**, **Forest Department (FD)**, and **Central Environmental Authority (CEA)**. Subnational variation occurs primarily through: (1) provincial implementation capacity differences; (2) heritage tree designations under local authority; and (3) the **Northern Province** post-conflict restoration context.

## Wet Zone vs. Dry Zone — Practical Regulatory Variation

Sri Lanka's biodiversity is concentrated in the **Wet Zone** (southwest, ~26% of land area, ~90% of plant endemics). While law applies uniformly, enforcement intensity differs significantly:

- **Wet Zone** (Western, Sabaragamuwa, Southern Provinces): FD has more ranger capacity; permits for forest produce more actively enforced; medicinal plant collection from Sinharaja UNESCO World Heritage Site is subject to FD Research Permit + UNESCO obligations.
- **Dry Zone** (Northern, Eastern, North Central, Uva Provinces): Larger areas per ranger; traditional community access to temple forest reserves (*Dematagolla*) is de facto recognised; REDD+ pilot areas in Uda Walawe buffer zone have specific NFA restrictions.

## Northern Province — Post-Conflict Conservation Context

Following the end of Sri Lanka's civil conflict (2009), the **Northern Province** (Jaffna Peninsula, Vanni) has undergone ecological recovery. The **FD Northern Province Range Office** (Vavuniya) has issued specific restrictions:

- Palmyra palm (*Borassus flabellifer*) — culturally significant to Tamil communities; tapping and felling restricted under FD orders to prevent further loss following conflict-period deforestation.
- Mangrove restoration areas in Jaffna Lagoon — CEA has designated specific mangrove restoration zones where any clearing or collection is prohibited pending recovery.

## Provincial Heritage Tree Programs

Several provincial councils and local authorities have designated **Heritage Trees** under local authority by-laws:

- **Colombo Municipal Council**: Heritage tree register with approximately 400 trees (species including *Terminalia arjuna*, *Ficus religiosa*, *Mangifera indica* heritage varieties) — CMC permit required for pruning; felling is prohibited.
- **Kandy Municipal Council**: Sacred Bo trees (*Ficus religiosa*) and rain trees in the Kandy World Heritage precinct — any interference requires both KMC and Buddhist Clergy consent.
- **Galle Fort Heritage Zone**: UNESCO World Heritage Site designation restricts any vegetation clearing within the Fort; Galle Municipal Council enforces this.

## Biodiversity Secretariat — Provincial Implementation

The **Biodiversity Secretariat (BS)** under the Ministry of Environment coordinates implementation across provinces. The **National Biodiversity Strategy and Action Plan (NBSAP 2016–2022)** identifies provincial-level implementation as a key gap — most provinces lack dedicated biodiversity officers, resulting in highly variable enforcement of the FFPO and FDO species protections.

<div class="alert alert-info" role="alert">
**Source note** — DWC Sri Lanka: [dwc.gov.lk](https://www.dwc.gov.lk) | FD Sri Lanka: [forestdept.gov.lk](https://www.forestdept.gov.lk) | CEA: [cea.lk](https://www.cea.lk) | Biodiversity Secretariat: [biodiversity.lk](https://www.biodiversity.lk).
</div>

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| DWC (Department of Wildlife Conservation) | FFPO administration; national parks; CITES management authority; Schedule V enforcement |
| Forest Department | Forest Ordinance enforcement; forest reserves; protected timber species permits |
| CEA (Central Environmental Authority) | Environmental permits; EIA oversight; invasive species coordination |
| National Botanic Gardens (Peradeniya) | Ex-situ plant conservation; seed banks; orchid collections |
| IUCN Sri Lanka | Red List assessments; conservation research |
| Ministry of Environment | Policy oversight; international convention coordination |

---

# Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Sri Lanka Status`=legal_status_lk,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Protected Plant Species in Sri Lanka",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Sri Lanka Status`=coalesce(legal_status_lk,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Sri Lanka — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Complete Species Table (Template Format) {#complete-template}

The table below presents all species in the **18-column `regulated_plants_v9.7` template format** — the canonical client deliverable schema. This is the same data as the downloadable CSV.

```{r complete-template-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    styled_kable(
      caption = "Complete Species Listing — 18-Column Template Format (GBIFusageKey, country, region, jurisdiction, jurisdiction_group, prefName, classification, taxonLevel, family, englishName, synonyms, Note, genus_I, is_aquatic, aquatic_type_I, legal_status, restriction_type, governing_law)",
      height = "650px"
    )
}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | FFPO Schedule V plant list is outdated — many IUCN-assessed threatened endemics not yet scheduled | High |
| 2 | Illegal timber trade (ebony, satinwood) through informal channels continues | High |
| 3 | Nepenthes distillatoria wet zone habitat destruction ongoing from land conversion | High |
| 4 | Ornamental plant collection from montane forests (orchids, ferns) poorly monitored | Medium |
| 5 | Water hyacinth management resources insufficient relative to scale of infestation | Medium |
| 6 | Coordination between DWC and Forest Department on overlapping jurisdictions | Medium |

---

# References {#references}

| Resource | Link |
|----------|------|
| DWC — Department of Wildlife Conservation | [dwc.gov.lk](https://www.dwc.gov.lk) |
| Forest Department Sri Lanka | [forestdept.gov.lk](https://www.forestdept.gov.lk) |
| CEA Sri Lanka | [cea.lk](https://www.cea.lk) |
| GBIF Sri Lanka | [gbif.org/country/LK](https://www.gbif.org/country/LK/summary) |
| CITES Sri Lanka | [cites.org — Sri Lanka](https://cites.org/eng/cms/index.php/component/cp/country/LK) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Verify at DWC / Forest Department / CITES</p></div>'))
```
