---
title: "Restricted and Protected Plant Species in Brazil"
subtitle: "Portaria MMA 443/2014 (National Threatened Plants), Lei 12.651/2012 (Native Vegetation Protection Law), Lei 13.123/2015 (Biodiversity Law) — Preliminary Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
.kqb_pkg <- "/Users/kwizera.jvk/Desktop/Kwiz Computing Comes to Life/Kwiz Quants/libs/kwizbranding"
suppressMessages(devtools::load_all(.kqb_pkg, quiet=TRUE))
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,            ~taxonomic_group, ~iucn_status,            ~cites_appendix,  ~legal_status_br,                              ~restriction_type,              ~governing_law,                    ~managing_institution,   ~notes,
  "Pau-brasil / Brazilwood",  "Paubrasilia echinata",      "Magnoliopsida",  "Endangered",            "Appendix II",    "Endangered (Port. 443/2014) — National Tree / CITES App. II","Collection/trade prohibited","Port. MMA 443/2014 / CITES",     "ICMBio / IBAMA",        "Brazil's national tree and namesake; bow-making wood; historically over-exploited; Atlantic Forest endemic; <10% habitat remains",
  "Paraná pine / Araucária",  "Araucaria angustifolia",    "Pinopsida",      "Critically Endangered", "Appendix I",     "CR (Port. 443/2014) — Strictly Protected (CITES App. I)","Felling/collection prohibited","Port. MMA 443/2014 / CITES",     "ICMBio / IBAMA",        "Atlantic Forest conifer; >97% population decline; seeds (pinhão) culturally important; Paraná and Santa Catarina",
  "Cattleya orchid",          "Cattleya labiata",           "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Vulnerable (Port. 443/2014 / CITES App. II)",  "Collection/trade regulated",   "Port. MMA 443/2014 / CITES",      "ICMBio",                "Brazil national orchid; flamboyant epiphyte; Pernambuco Atlantic Forest; wild collection for horticulture a major threat",
  "Giant Amazon water lily",  "Victoria amazonica",        "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected habitat — Amazon biome regulations", "Habitat protection",           "Lei 12.651/2012 / SNUC",          "ICMBio / IBAMA",        "Amazonian floodplain icon; floating leaves to 3 m; Amazonia habitat protected under Forest Code and SNUC",
  "Wild cacao",               "Theobroma cacao",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Wild relatives regulated — ABS under Lei 13.123","ABS permit required for research","Lei 13.123/2015",               "MMA / IBAMA",           "Wild cacao relatives in Amazon; biodiversity law requires access and benefit sharing for commercial genetic use",
  "Guaraná",                  "Paullinia cupana",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Commercially regulated — ABS and geographic indication","Commercial use regulated",  "Lei 13.123/2015 / INPI",          "MMA / IBAMA / INPI",    "Amazon stimulant; wild populations monitored; commercial cultivation extensive; geographic indication for Maués region",
  "Brazilian pepper tree",    "Schinus terebinthifolia",    "Magnoliopsida",  "Least Concern",         "Not listed",     "Native — invasive outside Brazil; monitored",  "Native — no restriction domestically","Lei 12.651/2012",              "IBAMA",                 "Aroeira-da-praia; Brazilian native; highly invasive in Florida/Caribbean/Hawaii; export/trade monitored under biosafety",
  "Water hyacinth",           "Eichhornia crassipes",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — management mandated",              "Control mandated",             "Lei 9.605/1998 / IBAMA regulations","IBAMA / ANA",          "Aguapé; covers >100,000 ha of Brazilian waterways; hydroelectric turbine and navigation impact; Pantanal concern",
  "Leucaena",                 "Leucaena leucocephala",      "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — control required in Atlantic Forest","Control mandated in reserves",  "IBAMA / ICMBio management plans", "ICMBio / IBAMA",        "Major invasive in Atlantic Forest fragments and Cerrado margins; nitrogen-fixer that alters soil chemistry",
  "Pinus",                    "Pinus elliottii",            "Pinopsida",      "Least Concern",         "Not listed",     "Invasive (South Brazil) — control in Atlantic Forest","Removal mandated in protected areas","CONAMA Resolution 369/2006",  "ICMBio / State enviro agencies","Plantation escapee; invades Araucaria mixed forest and grasslands in Paraná, Santa Catarina, Rio Grande do Sul"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","brazil_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_br,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Brazil is the world's most biodiverse nation, harbouring approximately **50,000 plant species** — roughly **20% of all plant species on Earth**. The country contains six major biomes: Amazon, Atlantic Forest, Cerrado, Caatinga, Pantanal, and Pampa. The Atlantic Forest is one of the world's most threatened and species-rich biomes, with less than 12% of original cover remaining. Brazil's national threatened plant list — **Portaria MMA 443/2014** — covers 2,113 species across IUCN threat categories. The **Native Vegetation Protection Law (Lei 12.651/2012)** — the Forest Code — governs vegetation protection on private land and defines Permanent Preservation Areas (APPs) and Legal Reserve requirements.

<div class="alert alert-info" role="alert">
**Preliminary Report** — Full species listing from Portaria MMA 443/2014 (National List of Threatened Plant Species, 2,113 species) to be completed. Current report covers `r n_species` representative species. Legislation is primarily in Portuguese; the REFLORA database ([reflora.jbrj.gov.br](http://reflora.jbrj.gov.br)) provides an English-language taxonomic interface.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Portaria MMA 443/2014 | 2014 | National List of Threatened Plant Species — 2,113 species in CR/EN/VU/NT/DD categories |
| Lei 12.651/2012 — Native Vegetation Protection Law (Forest Code) | 2012 | Permanent Preservation Areas (APPs); Legal Reserve (RL) requirements; vegetation protection on private land |
| Lei 13.123/2015 — Biodiversity Law | 2015 | Access and benefit sharing (ABS) for genetic resources; replaces MP 2.186-16/2001 |
| Lei 9.605/1998 — Environmental Crimes Law | 1998 | Criminal penalties for flora crimes; illegal deforestation; invasive species non-management |
| Sistema Nacional de Unidades de Conservação (SNUC) — Lei 9.985/2000 | 2000 | National protected areas system; 12 categories of conservation units |

Full text: [ICMBio](https://www.icmbio.gov.br) | [MMA](https://www.gov.br/mma) | [IBAMA](https://www.ibama.gov.br) | [FAOLEX Brazil](https://faolex.fao.org/brazil)

---

# Regional / Subnational Restrictions {#regional}

Brazil's federated system grants states (and the Federal District) significant environmental authority. States may enact their own threatened species lists, expand protected categories, and impose stricter harvest or trade controls than federal law — they may not be *weaker* than federal law. The most significant state instruments for plant species are summarised below.

| State | Key Instrument | Plants Scope | Enforcing Body |
|-------|---------------|-------------|----------------|
| **São Paulo** | Resolução SMA 57/2016 (*Lista de Espécies da Flora Ameaçadas*) | 314 threatened plant species listed; stricter than Portaria 443/2014 for many taxa; collection and commercialisation prohibited without CETESB/SMA permit | SEMA-SP (Secretaria de Meio Ambiente) / CETESB |
| **Minas Gerais** | Deliberação COPAM 367/2008 (*Lista da Flora Ameaçada de Extinção*) | 406 threatened plants listed; Cerrado species especially; major overlap with Atlantic Forest and Cerrado biomes | SUPRAM / SEMAD |
| **Paraná** | Decreto Estadual 3148/2004 (*Lista Vermelha da Flora Ameaçada — PR*) | ~236 species; araucaria-mixed forest species; forestry licensing conditioned on rare plant surveys | IAP (Instituto Ambiental do Paraná) |
| **Santa Catarina** | Resolução CONSEMA 51/2008 | 160+ threatened plant species; araucaria and Atlantic Forest taxa | FATMA (Fundação do Meio Ambiente) |
| **Rio Grande do Sul** | Decreto 52.109/2014 (*Lista de Espécies da Flora Nativa Ameaçadas*) | 300+ taxa listed; grassland (campo) species notably included; native grassland conversion restricted | SEMA-RS / FEPAM |
| **Amazonas** | Lei Estadual 3135/2007 (*Política Estadual sobre Mudanças Climáticas*) + Decreto 24.375/2004 (*SEUC-AM*) | State conservation units may restrict plant collection beyond federal rules; ABS provisions apply in state PAs | SDS Amazonas (now SEMA-AM) |
| **Pará** | SEMA-PA management plans for state conservation units | Timber and NTFP harvest restrictions within state PAs; enforcement of Forest Code APP and RL in the Amazon | SEMA-Pará / IDEFLOR-Bio |
| **Bahia** | IMA-BA (Instituto do Meio Ambiente e Recursos Hídricos) | Atlantic Forest remnants; Caatinga endemic plants; licensing conditioning | IMA Bahia |

## Key State-Level Mechanisms

1. **State threatened plant lists** — São Paulo, Minas Gerais, Paraná, Santa Catarina, and Rio Grande do Sul each maintain official state lists that add hundreds of species not covered by the federal Portaria MMA 443/2014. Commercial use, collection, and trade of listed species require state environmental permits.

2. **State Conservation Units (SEUC)** — Each state has its own system of conservation units (parks, ecological stations, APA areas) with individual management plans that may impose species-specific collection bans. Amazonian states (Amazônas, Pará, Mato Grosso) have extensive state PA networks.

3. **CONAMA Resolutions** — National-level (federal) but implemented regionally: CONAMA Resolution 369/2006 governs intervention in Permanent Preservation Areas (APPs); states add site-specific restrictions. CONAMA Resolution 423/2010 addresses Atlantic Forest species and native grassland.

4. **Forest Code compliance conditionality** — State environmental agencies (e.g., CETESB in SP, FEPAM in RS) enforce CAR (Cadastro Ambiental Rural) registration, a precondition for land use. Rural properties with irregular CAR status face restrictions on timber and NTFP harvest from native vegetation.

<div class="alert alert-info" role="alert">
**Source note** — State threatened plant lists are enacted by state secretariat resolutions or COPAM deliberations and may be updated independently of federal lists. Always verify the current state list with the relevant state environmental secretariat. Key portal: [Sistema Nacional de Informação sobre Meio Ambiente (SINIMA)](https://www.gov.br/mma/pt-br/acesso-a-informacao/sistemas-e-ferramentas/sinima) | [INMA — Instituto Nacional da Mata Atlântica](https://www.inma.gov.br).
</div>

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| ICMBio (Chico Mendes Institute for Biodiversity Conservation) | Protected areas management; threatened species action plans; CITES management authority |
| IBAMA (Brazilian Institute of Environment and Renewable Natural Resources) | Environmental licensing; enforcement; invasive species; timber trade monitoring |
| MMA (Ministry of Environment and Climate Change) | Biodiversity policy; Forest Code; threatened species listing; Portaria 443 |
| Jardim Botânico do Rio de Janeiro (JBRJ) | National herbarium; REFLORA database; plant taxonomy; Red List assessments |
| INPA (National Institute for Amazonian Research) | Amazonian flora research; botanical surveys |
| EMBRAPA | Agricultural biodiversity; crop wild relative conservation; genetic resources |

---

# Preliminary Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Brazil Status`=legal_status_br,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Preliminary — Protected Plant Species in Brazil",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Brazil Status`=coalesce(legal_status_br,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Brazil — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Amazon deforestation — 11,568 km² lost in 2022 (INPE); enforcement against illegal clearing inadequate | High |
| 2 | Forest Code (Lei 12.651/2012) 2012 amnesty provisions reduce legal liability for pre-2008 deforestation | High |
| 3 | Portaria 443/2014 covers 2,113 species but enforcement capacity insufficient relative to Atlantic Forest area | High |
| 4 | Araucaria angustifolia (CR) — illegal logging for timber continues in southern Brazil buffer zones | High |
| 5 | Lei 13.123/2015 ABS implementation for wild plant genetic resources incomplete — many sectors non-compliant | Medium |
| 6 | Invasive Pinus and Leucaena removal in protected areas severely underfunded | Medium |

---

# References {#references}

| Resource | Link |
|----------|------|
| ICMBio | [icmbio.gov.br](https://www.icmbio.gov.br) |
| IBAMA | [ibama.gov.br](https://www.ibama.gov.br) |
| JBRJ — REFLORA Database | [reflora.jbrj.gov.br](http://reflora.jbrj.gov.br) |
| Flora do Brasil 2020 | [floradobrasil.jbrj.gov.br](http://floradobrasil.jbrj.gov.br) |
| GBIF Brazil | [gbif.org/country/BR](https://www.gbif.org/country/BR/summary) |
| CITES Brazil | [cites.org — Brazil](https://cites.org/eng/cms/index.php/component/cp/country/BR) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Preliminary — Verify at ICMBio / IBAMA / JBRJ / CITES</p></div>'))
```
