---
title: "Restricted and Protected Plant Species in Singapore"
subtitle: "Parks and Trees Act 2005 (Cap. 216), Singapore Red Data Book (3rd Ed.) — Preliminary Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
.kqb_pkg <- "/Users/kwizera.jvk/Desktop/Kwiz Computing Comes to Life/Kwiz Quants/libs/kwizbranding"
suppressMessages(devtools::load_all(.kqb_pkg, quiet=TRUE))
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,            ~taxonomic_group, ~iucn_status,            ~cites_appendix,  ~legal_status_sg,                              ~restriction_type,              ~governing_law,                    ~managing_institution,   ~notes,
  "Tembusu",                 "Fagraea fragrans",           "Magnoliopsida",  "Least Concern",         "Not listed",     "Heritage Tree — Parks and Trees Act 2005",    "Removal/damage prohibited",    "Parks and Trees Act 2005 (Cap.216)","NParks",               "Featured on Singapore $5 note; one of Singapore's most iconic trees; 235 heritage trees designated nationally",
  "Sea apple / Jambu ayer laut","Syzygium grande",         "Magnoliopsida",  "Near Threatened",       "Not listed",     "Heritage Tree / Parks and Trees Act 2005",    "Protected from removal",       "Parks and Trees Act 2005 (Cap.216)","NParks",               "Coastal heritage tree; important in Singapore's streetscape and park canopies; heritage tree programme",
  "Raffles' pitcher plant",  "Nepenthes rafflesiana",      "Magnoliopsida",  "Vulnerable",            "Appendix II",    "Protected (Singapore Red Data Book VU / CITES App. II)","Collection prohibited",  "Parks and Trees Act / CITES",     "NParks",                "Named for Stamford Raffles; Singapore's last remaining lowland heath forest pitcher plant populations",
  "Singapore orchid",        "Spathoglottis plicata",      "Magnoliopsida",  "Least Concern",         "Appendix II",    "Protected (CITES App. II) — heritage significance","Collection regulated",      "CITES / Parks and Trees Act",     "NParks",                "Singapore Purple Orchid; common terrestrial orchid; CITES App. II; important in Singapore's horticultural heritage",
  "Jade vine",               "Strongylodon macrobotrys",   "Magnoliopsida",  "Vulnerable",            "Not listed",     "Vulnerable (Red Data Book) — cultivated/protected","Ex-situ protection",      "NParks Gardens Conservation",     "NParks / SBGC",         "Spectacular Philippine vine; VU; cultivated in Singapore Botanic Gardens; wild range threatened",
  "Nipah palm",              "Nypa fruticans",             "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected (mangrove habitat protection)",     "Mangrove habitat protected",   "Parks and Trees Act / MPA regulations","NParks",            "Mangrove associate; Sundarbans and Singapore coastal habitats; Singapore mangrove restoration included",
  "Jelutong",                "Dyera costulata",            "Magnoliopsida",  "Vulnerable",            "Not listed",     "Vulnerable (Red Data Book) — protected in reserves","Habitat protection",      "Parks and Trees Act 2005",        "NParks",                "Lowland rainforest tree; latex historically used for chewing gum; VU; Bukit Timah NR populations protected",
  "Mimosa vine",             "Mimosa pigra",               "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — active management",                "Control mandated",             "NParks invasive species programme","NParks",               "Giant sensitive plant; invades freshwater wetlands; Sungei Buloh and Pulau Ubin; outcompetes native mangrove fringes",
  "Mile-a-minute weed",      "Mikania micrantha",          "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — active management",                "Control mandated",             "NParks invasive species programme","NParks",               "Smothers native vegetation in Singapore's secondary forests and forest edges; widespread in southern reserves",
  "Singapore rhododendron",  "Melastoma malabathricum",    "Magnoliopsida",  "Least Concern",         "Not listed",     "Native species — protected in nature reserves","Habitat protection in reserves","Parks and Trees Act 2005",       "NParks",                "Senduduk; common pioneer; protected within nature reserves; important for native bee pollinators in Singapore"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","singapore_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_sg,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Singapore retains approximately **2,280 native plant species** on its main island and associated islets — a remarkable number for a city-state of only 733 km². The country has lost over **95% of its original forest cover** since the 19th century, making the remaining patches in Bukit Timah Nature Reserve, the Central Catchment, and the Southern Ridges exceptionally important. Singapore's plant protection regime is anchored in the **Parks and Trees Act 2005 (Cap. 216)**, which empowers NParks to designate **Heritage Trees** and manage nature reserves. Singapore also operates one of the world's most comprehensive **Heritage Tree programmes**, with over **235 designated Heritage Trees** across the island.

<div class="alert alert-info" role="alert">
**Preliminary Report** — Full species listing from the Singapore Red Data Book (3rd Edition, NParks 2021) to be completed. Current report covers `r n_species` representative species. NParks ([nparks.gov.sg](https://www.nparks.gov.sg)) publishes the authoritative species lists and heritage tree register.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Parks and Trees Act (Cap. 216) | 2005 (Rev. Ed.) | Heritage Trees designation; nature reserve management; plant protection; fines up to SGD 50,000 |
| National Parks Board Act (Cap. 198A) | 1996 (Rev. Ed.) | NParks powers; garden city management; ex-situ conservation |
| Endangered Species (Import and Export) Act (Cap. 92A) | 1989 (Rev. Ed.) | CITES implementation; import/export permits for listed species |
| Environmental Protection and Management Act | 1999 (Rev. Ed.) | Environmental quality; pollution control; integrated with NParks operations |
| Singapore Red Data Book (3rd Ed.) | 2021 | Scientific threat assessment for native flora and fauna (NParks publication) |

Full text: [NParks](https://www.nparks.gov.sg) | [Singapore Statutes Online](https://sso.agc.gov.sg) | [FAOLEX Singapore](https://faolex.fao.org/singapore)

---

# Subnational / Zonal Restrictions {#regional}

Singapore is a **city-state** with no states, provinces, or other subnational jurisdictions — national legislation applies uniformly across the entire 733 km² territory. There are no separate regional environmental agencies or subnational legislatures. However, functionally significant spatial variation exists through **NParks' designated conservation zones**, **Urban Redevelopment Authority (URA) conservation area plans**, and **park management regulations** that create a de facto zonal restriction system for plant species.

## NParks Conservation Areas — Management Zone System

**National Parks Board (NParks)** manages Singapore's parks, nature reserves, and green corridors under the **National Parks Board Act (Cap. 198A)**. The four **Nature Reserves** designated under the **Parks and Trees Act Cap. 216** have the strongest protections:

| Nature Reserve / Park | Location | Plant Species Restrictions |
|----------------------|----------|--------------------------|
| **Bukit Timah Nature Reserve** | Central Catchment | Primary rainforest remnant; absolute prohibition on plant collection, transplanting, or damage; *Shorea* dipterocarp species, *Hopea sangal*, and endemic *Livistona singapurensis* palm protected under NR designation; research access by NParks permit only |
| **Central Catchment Nature Reserve (CCNR)** | Central Singapore | Singapore's largest forest reserve (3,043 ha); *Livistona singapurensis*, *Hopea*, and *Shorea* species; no collection; buffer zones with additional permit requirements |
| **Labrador Nature Reserve** | Southern Ridge | Coastal cliff forest; rare coastal plants; collection prohibited; access controlled by NParks |
| **Sungei Buloh Wetland Reserve** | Northern coast | Mangroves and coastal wetland plants; Ramsar site; all plant collection prohibited; mangrove species research requires NParks permit |

## Parks and Trees Act — Street Tree and Heritage Tree Register

Beyond nature reserves, the **Parks and Trees Act** regulates:

1. **Streetside/roadside trees**: All trees along public roads are under NParks management; removal or pruning requires NParks approval regardless of the tree's protected status.

2. **Heritage Tree Register**: NParks maintains a Heritage Tree Register of trees with high cultural, historical, or ecological significance. As of 2024, **261 Heritage Trees** are listed, spanning 53 species including:
   - *Ficus benghalensis* (Banyan) — multiple specimens
   - *Terminalia catappa* (Sea almond) — beachfront heritage trees
   - *Tembusu* (*Fagraea fragrans*) — iconic Singapore tree on the $5 note
   - **Penalties for damaging Heritage Trees**: Up to SGD $50,000 fine under the Parks and Trees Act.

3. **Vacant Land (weed/invasive) regulations**: NEA and NParks jointly enforce clearance of invasive plants on vacant land; *Mimosa pigra*, *Mikania micrantha*, and *Chromolaena odorata* control is mandated.

## Urban Redevelopment Authority (URA) — Conservation Area Green Rules

URA's **Master Plan** designates conservation areas (e.g., Chinatown, Kampong Glam, Little India, Emerald Hill) where removal of existing mature trees on conserved buildings requires a **Conservation Technical Condition (CTC)** from URA — a layer above NParks requirements. This effectively creates additional tree protection in heritage districts.

<div class="alert alert-info" role="alert">
**Note on Subnational Absence** — Singapore has no subnational environmental legislation. All species and plant restrictions are implemented nationally through NParks, NEA, and AVA (Agri-Food and Veterinary Authority — now SFA). The zonal management system described above represents the functional equivalent of subnational variation in a city-state context.
</div>

<div class="alert alert-info" role="alert">
**Source note** — NParks: [nparks.gov.sg](https://www.nparks.gov.sg) | Heritage Trees: [nparks.gov.sg/gardens-parks-and-nature/heritage-trees](https://www.nparks.gov.sg) | Parks and Trees Act: [Singapore Statutes Online](https://sso.agc.gov.sg) | URA Master Plan: [ura.gov.sg](https://www.ura.gov.sg).
</div>

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| NParks (National Parks Board) | Parks and Trees Act administration; Heritage Trees; nature reserves; CITES management authority |
| AVS (Animal and Veterinary Service) | Animal and plant import/export controls; biosecurity |
| NEA (National Environment Agency) | Environmental quality; pollution controls; environmental health |
| Singapore Botanic Gardens (SBGC) | UNESCO World Heritage Site; orchid breeding; seed bank; ex-situ conservation |
| Lee Kong Chian Natural History Museum | Herbarium collections; plant taxonomy; species identification |
| NUS Dept of Biological Sciences | Botanical research; flora surveys; Red Data Book assessments |

---

# Preliminary Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Singapore Status`=legal_status_sg,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Preliminary — Protected Plant Species in Singapore",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Singapore Status`=coalesce(legal_status_sg,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Singapore — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Remaining native forest extremely fragmented — any development adjacent to reserves creates edge effects | High |
| 2 | Invasive species (Mikania, Mimosa) management in secondary forest areas resource-constrained | High |
| 3 | Climate change — Singapore temperatures rising; montane species have no altitude refuge to migrate | Medium |
| 4 | Heritage Tree programme focuses on individual specimens, not populations or habitats | Medium |
| 5 | Red Data Book (3rd Ed.) species listings not all legally formalised under Parks and Trees Act | Medium |
| 6 | Limited seed banking capacity relative to number of critically threatened native plants | Low |

---

# References {#references}

| Resource | Link |
|----------|------|
| NParks Singapore | [nparks.gov.sg](https://www.nparks.gov.sg) |
| Singapore Botanic Gardens | [sbg.org.sg](https://www.sbg.org.sg) |
| Singapore Red Data Book | [NParks RDB](https://www.nparks.gov.sg/biodiversity/wildlife-in-singapore/species-list) |
| GBIF Singapore | [gbif.org/country/SG](https://www.gbif.org/country/SG/summary) |
| CITES Singapore | [cites.org — Singapore](https://cites.org/eng/cms/index.php/component/cp/country/SG) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Preliminary — Verify at NParks / Singapore Red Data Book / CITES</p></div>'))
```
