---
title: "Restricted and Protected Plant Species in India"
subtitle: "Wildlife Protection Act Schedule VI, Biological Diversity Act, and CITES — A Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")

suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})

# ── kwizbranding ──────────────────────────────────────────────────────────────
if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(), "docs", "www"), file.path(getwd(), "www"),
                            file.path(getwd()), file.path(here::here()))
.kqb_load_logo <- function(fn, ck) {
  for (d in .kqb_logo_candidates) {
    p <- file.path(d, fn)
    if (file.exists(p)) {
      assign(ck, paste0("data:image/svg+xml;base64,", base64enc::base64encode(readBin(p, "raw", file.size(p)))),
             envir = .GlobalEnv); return(invisible(TRUE))
    }
  }
}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",    ".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg", ".kqb_display_cache")

styled_kable <- function(df, caption=NULL, scroll=TRUE, height="450px") {
  k <- df |> kbl(caption=caption, escape=FALSE, align="l") |>
    kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),
                  full_width=TRUE, font_size=12)
  if (scroll) k <- scroll_box(k, height=height)
  k
}

# ── Embedded species data ─────────────────────────────────────────────────────
# Source: Wild Life (Protection) Act 1972 Schedule VI (as amended by WPA Amendment Act 2022),
#         Biological Diversity Act 2002, CITES Appendices (CoP19, 2022).
# Schedule VI lists "Specified Plants" — collection, possession, trade prohibited without licence.
species <- tribble(
  ~common_name, ~scientific_name, ~taxonomic_group, ~iucn_status, ~cites_appendix, ~legal_status_india, ~restriction_type, ~governing_law, ~managing_institution, ~notes,
  # ── WPA 1972 Schedule VI — Specified Plants (all listed taxa) ─────────────────
  "Slipper orchid (India)", "Paphiopedilum druryi", "Magnoliopsida", "Critically Endangered", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Most threatened Indian slipper orchid; Agasthyamalai hills; <100 plants",
  "Paphiopedilum fairrieanum", "Paphiopedilum fairrieanum", "Magnoliopsida", "Endangered", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Himalayan slipper orchid; Sikkim/Arunachal; horticultural demand",
  "Paphiopedilum venustum", "Paphiopedilum venustum", "Magnoliopsida", "Vulnerable", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "NE India slipper orchid; lowland forests of Assam/Meghalaya",
  "Paphiopedilum villosum", "Paphiopedilum villosum", "Magnoliopsida", "Vulnerable", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "NE India/Myanmar border; sub-montane forests; heavy collection",
  "Paphiopedilum spicerianum", "Paphiopedilum spicerianum", "Magnoliopsida", "Endangered", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "NE India endemic; Assam/Nagaland/Manipur; smuggling for trade",
  "Paphiopedilum insigne", "Paphiopedilum insigne", "Magnoliopsida", "Vulnerable", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "NE India; Nagaland/Manipur; historically over-collected for European trade",
  "Lady slipper orchid", "Cypripedium cordigerum", "Magnoliopsida", "Least Concern", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "High-altitude Himalayan orchid; entire genus Schedule VI",
  "Cypripedium himalaicum", "Cypripedium himalaicum", "Magnoliopsida", "Least Concern", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Himalayan alpine orchid; J&K to Arunachal; high-altitude meadows",
  "Cypripedium elegans", "Cypripedium elegans", "Magnoliopsida", "Least Concern", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Small Himalayan slipper orchid; Sikkim/Bhutan border forests",
  "Blue vanda orchid", "Vanda coerulea", "Magnoliopsida", "Endangered", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "NE India; heavily collected for horticulture; blue orchid icon",
  "Renanthera imschootiana", "Renanthera imschootiana", "Magnoliopsida", "Vulnerable", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Red vanda; NE India/Myanmar; montane forest epiphyte; rare",
  "Indian marsh orchid", "Dactylorhiza hatagirea", "Magnoliopsida", "Least Concern", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Alpine orchid; used in Shilajit preparation; heavy collection pressure",
  "Kuth / Costus root", "Saussurea costus", "Magnoliopsida", "Critically Endangered", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Critically threatened; traditional medicine and perfumery; Himalayan",
  "Indian podophyllum", "Podophyllum hexandrum", "Magnoliopsida", "Vulnerable", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Source of podophyllotoxin (anti-cancer drug); alpine Himalaya; over-harvested",
  "Cycas beddomei", "Cycas beddomei", "Cycadopsida", "Critically Endangered", "Appendix I", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Endemic to Andhra Pradesh; all Cycas spp. Schedule VI",
  "Cycas circinalis", "Cycas circinalis", "Cycadopsida", "Endangered", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Queen sago; Western Ghats; traditional food plant; habitat loss",
  "Cycas pectinata", "Cycas pectinata", "Cycadopsida", "Vulnerable", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "NE India hill cycad; Assam/Meghalaya/Arunachal; forest loss",
  "Himalayan yew", "Taxus wallichiana", "Pinopsida", "Endangered", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Source of taxol (cancer treatment); heavily exploited across Himalaya",
  "Indian serpentine / Rauvolfia", "Rauvolfia serpentina", "Magnoliopsida", "Least Concern", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Sedative alkaloid reserpine; over-collected from wild",
  "Shilajeet plant / Frerea", "Frerea indica", "Magnoliopsida", "Critically Endangered", "Not listed", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Extremely rare succulent; Junnar, Pune dist.; <50 plants known",
  "Dioscorea (medicinal yam)", "Dioscorea deltoidea", "Magnoliopsida", "Least Concern", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Source of diosgenin (contraceptive); over-harvested in Himalaya",
  "Spikenard / Jatamansi", "Nardostachys jatamansi", "Magnoliopsida", "Critically Endangered", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Himalayan aromatic herb; heavily traded for ayurvedic use; CITES II",
  "Himalayan poppy", "Meconopsis regia", "Magnoliopsida", "Least Concern", "Not listed", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Alpine poppy; high-altitude endemic; ornamental collection threat",
  "Red Sanders", "Pterocarpus santalinus", "Magnoliopsida", "Endangered", "Appendix II", "Schedule VI — Specified Plant", "Harvesting / export restricted", "WPA 1972 Schedule VI", "MoEFCC / WCCB / APFD", "Most trafficked plant; smuggling to China/SE Asia; Seshachalam hills endemic",
  "Pitcher plant (Indian)", "Nepenthes khasiana", "Magnoliopsida", "Endangered", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Only Nepenthes in India; Meghalaya endemic; carnivorous plant; habitat loss",
  "Euphorbia (succulent)", "Euphorbia nivulia", "Magnoliopsida", "Least Concern", "Appendix II", "Schedule VI — Specified Plant", "Regulated collection/trade", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Succulent euphorbia; entire genus regulated under CITES",
  "Kutki / Picrorhiza", "Picrorhiza kurrooa", "Magnoliopsida", "Endangered", "Not listed", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Alpine Himalayan herb; hepatoprotective; massively over-collected",
  "Ativisha / Aconitum", "Aconitum heterophyllum", "Magnoliopsida", "Endangered", "Not listed", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Himalayan monkshood; ayurvedic Ativisha; anti-fever; extremely rare",
  "Ceropegia (India)", "Ceropegia spp.", "Magnoliopsida", "Various", "Appendix II", "Schedule VI — Specified Plant", "Collection/trade prohibited", "WPA 1972 Schedule VI", "MoEFCC / WCCB", "Entire genus listed; several Indian endemics CR/EN; tuberous plants",
  # ── CITES-listed Indian plants (not in Schedule VI) ───────────────────────────
  "Indian sandalwood", "Santalum album", "Magnoliopsida", "Vulnerable", "Appendix II", "State Protected Tree / CITES II", "Govt owned in Karnataka/AP", "Karnataka Forest Act / AP Forest Act / CITES", "State Forest Dept / WCCB", "Govt ownership in Karnataka; major illicit trade; CITES II since CoP19",
  "Agarwood (Indian)", "Aquilaria malaccensis", "Magnoliopsida", "Critically Endangered", "Appendix II", "CITES App. II — Export regulated", "Export requires CITES permit", "CITES / WPA / EXIM Policy", "MoEFCC / WCCB", "Resinous heartwood for oud perfume; NE India; massively over-harvested",
  "Orchid tree / Paphiopedilum hirsutissimum", "Paphiopedilum hirsutissimum", "Magnoliopsida", "Vulnerable", "Appendix I", "CITES App. I", "All trade prohibited", "CITES / WPA 1972", "MoEFCC / WCCB", "NE India/SE Asia; large-flowered slipper orchid; smuggling for collectors",
  "Dendrobium nobile", "Dendrobium nobile", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "Noble dendrobium; NE India; TCM ingredient; heavily traded from Himalayan forests",
  "Aerides multiflora", "Aerides multiflora", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "Fox-brush orchid; W Ghats and NE India; epiphytic; local trade",
  "Bulbophyllum leopardinum", "Bulbophyllum leopardinum", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "Himalayan mini-orchid; Sikkim/Arunachal; all Orchidaceae CITES II",
  "Coelogyne cristata", "Coelogyne cristata", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "Angel orchid; Himalayan epiphyte; popular in horticulture; mass-collected",
  "Cymbidium devonianum", "Cymbidium devonianum", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "NE India pendant cymbidium; popular orchid; wild collection pressure",
  "Phalaenopsis taenialis", "Phalaenopsis taenialis", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "Himalayan Phalaenopsis; deciduous; winter-flowering; collection pressure",
  "Pleione praecox", "Pleione praecox", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "Peacock orchid; NE India; autumn-flowering alpine; horticultural demand",
  "Thunia alba", "Thunia alba", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "White thunia; Himalayan terrestrial orchid; traded for gardens",
  "Rhynchostylis retusa", "Rhynchostylis retusa", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II / State flower Assam", "Export regulated", "CITES / Assam Forest rules", "MoEFCC / WCCB / Assam FD", "Foxtail orchid; state flower of Assam/Arunachal; festival use (Bihu)",
  # ── Biological Diversity Act regulated species ────────────────────────────────
  "Ashwagandha (wild)", "Withania somnifera", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "ABS permit for commercial use", "Biological Diversity Act 2002", "NBA / State SBBs", "Key ayurvedic adaptogen; wild populations declining; now mostly cultivated",
  "Guduchi / Giloy", "Tinospora cordifolia", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "ABS permit for commercial use", "Biological Diversity Act 2002", "NBA / State SBBs", "Ayurvedic immunomodulator; demand surged during COVID-19; wild harvest concerns",
  "Brahmi", "Bacopa monnieri", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "ABS permit for commercial use", "Biological Diversity Act 2002", "NBA / State SBBs", "Memory-enhancing herb; international supplement trade; wetland species",
  "Shatavari", "Asparagus racemosus", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "ABS permit for commercial use", "Biological Diversity Act 2002", "NBA / State SBBs", "Ayurvedic tonic; root over-harvest; declining wild populations across India",
  "Shankhpushpi", "Convolvulus prostratus", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "ABS permit for commercial use", "Biological Diversity Act 2002", "NBA / State SBBs", "Brain tonic herb; grassland species; commercial ayurvedic demand",
  "Guggul", "Commiphora wightii", "Magnoliopsida", "Critically Endangered", "Not listed", "BD Act — ABS regulated / State protected", "Harvest regulated by State FD", "Biological Diversity Act 2002 / State Forest Acts", "NBA / Rajasthan FD", "Oleo-gum resin for cholesterol; CR; Rajasthan/Gujarat arid zone; over-tapped",
  "Isabgol / Psyllium", "Plantago ovata", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS for wild genetic resources", "ABS for commercial cultivar development", "Biological Diversity Act 2002", "NBA / ICAR", "Wild relatives in Rajasthan/Gujarat; cultivated commercially; ABS for genetics",
  "Safed Musli", "Chlorophytum borivilianum", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "ABS permit for commercial use", "Biological Diversity Act 2002", "NBA / State SBBs", "Tuberous aphrodisiac herb; wild populations severely depleted; now cultivated",
  "Amla / Indian gooseberry", "Phyllanthus emblica", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated / Forest produce", "Collection regulated under Forest Acts", "Biological Diversity Act 2002 / State Forest Acts", "NBA / State FDs", "Key ayurvedic fruit; NTFP under forest law; commercial ABS applies",
  "Arjun tree", "Terminalia arjuna", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "Bark harvest regulated", "Biological Diversity Act 2002 / State Forest Acts", "NBA / State FDs", "Cardiotonic bark; heavily harvested for herbal medicine; riverbank species",
  "Bael / Wood apple", "Aegle marmelos", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "Collection regulated", "Biological Diversity Act 2002", "NBA / State SBBs", "Sacred tree; Shiva worship; medicinal fruit; wild harvest monitored",
  "Tulsi / Holy basil", "Ocimum tenuiflorum", "Magnoliopsida", "Least Concern", "Not listed", "BD Act — ABS regulated", "ABS for commercial products", "Biological Diversity Act 2002", "NBA / State SBBs", "Sacred medicinal herb; commercial supplement industry; ABS required for export",
  # ── State-protected trees ─────────────────────────────────────────────────────
  "Teak (natural forests)", "Tectona grandis", "Magnoliopsida", "Least Concern", "Not listed", "State Protected — felling restricted", "Felling requires Forest Dept permit", "State Forest Acts (multiple states)", "State Forest Depts", "Logging from natural forests restricted in most states; plantations regulated",
  "Sal tree", "Shorea robusta", "Magnoliopsida", "Least Concern", "Not listed", "State Protected — revenue tree", "Felling restricted under Forest Acts", "State Forest Acts (UP/MP/Bihar/Jharkhand/WB)", "State Forest Depts", "Major timber tree; government-controlled; NTFP (sal leaves/seeds) regulated",
  "Khejri / Prosopis cineraria", "Prosopis cineraria", "Magnoliopsida", "Least Concern", "Not listed", "State tree Rajasthan — Protected", "Felling prohibited", "Rajasthan Forest Act / Bishnoi customary law", "Rajasthan Forest Dept", "Desert keystone; Bishnoi community protection; state tree; felling banned",
  "Shisham / Indian rosewood", "Dalbergia sissoo", "Magnoliopsida", "Least Concern", "Appendix II", "CITES App. II (all Dalbergia)", "Export regulated", "CITES / State Forest Acts", "State FDs / WCCB", "All Dalbergia CITES II since 2017; major timber; wilt disease devastating plantations",
  "Indian rosewood / Malabar", "Dalbergia latifolia", "Magnoliopsida", "Vulnerable", "Appendix II", "CITES App. II / State Protected", "Export prohibited / felling regulated", "CITES / Karnataka/Kerala Forest Acts", "State FDs / WCCB", "Premium furniture timber; South Indian species; CITES II",
  "Sissoo / Dalbergia melanoxylon (India range)", "Dalbergia melanoxylon", "Magnoliopsida", "Near Threatened", "Appendix II", "CITES App. II", "Export regulated", "CITES", "MoEFCC / WCCB", "African blackwood also in Indian trade; all Dalbergia CITES II",
  # ── Invasive alien species ────────────────────────────────────────────────────
  "Water hyacinth", "Eichhornia crassipes", "Magnoliopsida", "Least Concern", "Not listed", "Notified Invasive Species", "Control mandated by states", "EP Act 1986 / State rules", "MoEFCC / State PCBs", "Major freshwater invasive; covers millions of ha; hydropower threat",
  "Parthenium weed", "Parthenium hysterophorus", "Magnoliopsida", "Least Concern", "Not listed", "Invasive / Noxious Weed", "State-level control orders", "State Agricultural Acts / EP Act", "ICAR / State Depts", "Widespread invasive; severe human allergy; crop competitor; all states",
  "Lantana", "Lantana camara", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — National Park threat", "Eradication in protected areas", "WPA 1972 / PA management plans", "Forest Dept / Protected Areas", "Major forest understory invasive; covers 13M ha; managed in tiger reserves",
  "Prosopis (mesquite)", "Prosopis juliflora", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — semi-arid degradation", "State control programmes", "EP Act 1986 / State rules", "MoEFCC / CAZRI", "Semi-arid invasive; massive spread in Rajasthan, Gujarat, Tamil Nadu; >5M ha",
  "Mikania (mile-a-minute)", "Mikania micrantha", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — NE India / Western Ghats", "Eradication in protected areas", "WPA 1972 / PA management plans", "Forest Dept / Tea estates", "Smothering climber; invades tea gardens and NE India forests; rapid growth",
  "Chromolaena / Siam weed", "Chromolaena odorata", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — Western Ghats/NE India", "Control in protected areas", "PA management plans", "Forest Dept", "Fire-promoting invasive shrub; transforms forest understorey; W Ghats",
  "Ipomoea carnea", "Ipomoea carnea", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — wetland degrader", "Control mandated in some states", "State wetland rules", "State PCBs / Forest Dept", "Toxic shrub; invades wetlands and wastelands; cattle poisoning",
  "Salvinia (giant)", "Salvinia molesta", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — aquatic weed", "Control mandated", "EP Act 1986 / State rules", "State PCBs / ICAR", "Giant floating fern; smothers Kerala backwaters; biological control deployed",
  "Cabomba", "Cabomba caroliniana", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — aquarium trade escape", "Import restricted", "Quarantine rules / EXIM Policy", "MoEFCC / Plant quarantine", "Aquatic weed spreading from aquarium trade; Kerala/Tamil Nadu waterways",
  "Mesquite / Vilayati babool", "Prosopis glandulosa", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — arid zone", "State control programmes", "EP Act 1986 / State rules", "CAZRI / State FDs", "Dryland invasive hybridising with P. juliflora; Northern India spread",
  "Mimosa (giant sensitive)", "Mimosa diplotricha", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — NE India plantation threat", "Control in tea/rubber estates", "State Agricultural Acts", "ICAR / State Depts", "Thorny scrambler; invades tea gardens and roadsides in NE India",
  "Ageratum / Goat weed", "Ageratum conyzoides", "Magnoliopsida", "Least Concern", "Not listed", "Invasive — crop fields", "Weed management mandated", "State Agricultural Acts", "ICAR / State Depts", "Pantropical weed; major rice and tea field invader; allelopathic"
)

n_species    <- nrow(species)
n_sch6       <- sum(str_detect(species$governing_law, "WPA"), na.rm = TRUE)
n_invasive   <- sum(str_detect(species$restriction_type, regex("invasive|control|eradication", ignore_case = TRUE)), na.rm = TRUE)
n_cites_i    <- sum(species$cites_appendix == "Appendix I",  na.rm = TRUE)
n_cites_ii   <- sum(species$cites_appendix == "Appendix II", na.rm = TRUE)
n_cr         <- sum(species$iucn_status == "Critically Endangered", na.rm = TRUE)
n_en         <- sum(species$iucn_status == "Endangered", na.rm = TRUE)

# ── GBIF resolution ──────────────────────────────────────────────────────────
proj_root  <- here::here()
gbif_cache <- file.path(proj_root, "output", "tables", "india_regulated_plants_gbif.csv")

if (!file.exists(gbif_cache)) {
  clean_name <- function(nm) nm |> str_remove_all("\\s+spp\\.?$") |>
    str_remove_all("\\(.*?\\)") |> str_squish()
  names_df <- tibble(name = clean_name(species$scientific_name), kingdom = "Plantae")
  backbone_res <- tryCatch(name_backbone_checklist(names_df), error = function(e) {
    map_dfr(names_df$name, function(nm) tryCatch(
      as_tibble(name_backbone(name=nm, kingdom="Plantae")),
      error=function(e2) tibble(usageKey=NA_integer_, canonicalName=nm, matchType="NONE")))
  })
  get_vernacular <- function(key) {
    if (is.na(key)) return(NA_character_)
    tryCatch({
      vn <- name_usage(key=as.numeric(key), data="vernacularNames", limit=100)
      if (is.null(vn$data)||nrow(vn$data)==0) return(NA_character_)
      en <- vn$data |> filter(language %in% c("eng","en")) |> pull(vernacularName) |> unique()
      if (length(en)==0) NA_character_ else paste(en, collapse=", ")
    }, error=function(e) NA_character_)
  }
  get_synonyms <- function(key) {
    if (is.na(key)) return(NA_character_)
    tryCatch({
      syn <- name_usage(key=as.numeric(key), data="synonyms", limit=100)
      if (is.null(syn$data)||nrow(syn$data)==0) return(NA_character_)
      s <- syn$data |> pull(canonicalName) |> unique() |> na.omit()
      if (length(s)==0) NA_character_ else paste(s, collapse=", ")
    }, error=function(e) NA_character_)
  }
  backbone_clean <- backbone_res |> as_tibble() |>
    mutate(GBIFusageKey=as.character(usageKey),
           prefName=coalesce(species, canonicalName, verbatim_name),
           gbif_family=family, gbif_genus=genus)
  species_matched <- species |> mutate(clean_sci=clean_name(scientific_name)) |>
    bind_cols(backbone_clean |> select(GBIFusageKey, prefName, gbif_family, gbif_genus,
                                        gbif_matchType=matchType, gbif_confidence=confidence))
  keys <- species_matched |> filter(!is.na(GBIFusageKey)) |> distinct(GBIFusageKey) |> pull(GBIFusageKey)
  vl <- list(); sl <- list()
  for (i in seq_along(keys)) {
    k <- keys[i]; vl[[k]] <- get_vernacular(k); sl[[k]] <- get_synonyms(k)
    if (i %% 5==0) Sys.sleep(0.2)
  }
  vd <- tibble(GBIFusageKey=names(vl), gbif_english=unlist(vl), gbif_synonyms=unlist(sl[names(vl)]))
  enriched <- species_matched |> left_join(vd, by="GBIFusageKey") |>
    mutate(englishName=coalesce(gbif_english, common_name), synonyms=gbif_synonyms, family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"), recursive=TRUE, showWarnings=FALSE)
  enriched |> mutate(country="India",region="",jurisdiction="NATIONAL",jurisdiction_group="",
    classification=taxonomic_group,taxonLevel="species",Note=notes,genus_I=word(prefName,1),is_aquatic="no",aquatic_type_I="") |>
    select(GBIFusageKey,country,region,jurisdiction,jurisdiction_group,prefName,classification,taxonLevel,family,englishName,synonyms,Note,genus_I,is_aquatic,aquatic_type_I,
                     legal_status_india, restriction_type, governing_law) |> write_csv(gbif_cache)
}
if (file.exists(gbif_cache)) {
  plants_gbif     <- read_csv(gbif_cache, col_types=cols(.default=col_character()), show_col_types=FALSE)
  n_gbif_resolved <- sum(!is.na(plants_gbif$GBIFusageKey) & plants_gbif$GBIFusageKey!="")
  n_gbif_families <- n_distinct(plants_gbif$family, na.rm=TRUE)
} else {
  plants_gbif <- NULL; n_gbif_resolved <- 0L; n_gbif_families <- 0L
}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04, size="280px auto"))
uri <- kqb_watermark_uri()
if (nzchar(uri)) cat(paste0('<style>\nbody::after {\n  content: "";\n  position: fixed;\n',
  '  top:0; left:0; width:100%; height:100%;\n  z-index:9000;\n  pointer-events:none;\n',
  '  background-image: url("', uri, '");\n  background-repeat:repeat;\n  background-size:280px auto;\n',
  '  opacity:0.04;\n}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

India is one of the world's **17 megadiverse** countries, hosting approximately **47,513 plant species** — representing about 11% of global flora — across 4 recognised biodiversity hotspots (Himalaya, Western Ghats, Indo-Burma, Sundaland). The country's flora includes numerous species of global conservation significance that are simultaneously under severe collection and trade pressure.

This report documents **`r n_species` plant species** restricted under Indian law. **`r n_sch6`** are listed on Schedule VI of the Wild Life (Protection) Act 1972 (as amended 2022), making collection, possession, and trade unlawful without a permit from the Chief Wildlife Warden. **`r n_invasive`** are notified invasive species. **`r n_cites_i`** carry CITES Appendix I status and **`r n_cites_ii`** are on Appendix II. **`r n_cr`** are IUCN Critically Endangered and **`r n_en`** are Endangered.

---

# Legal Framework {#legal-framework}

## Wild Life (Protection) Act, 1972 — Schedule VI {#wpa}

The **Wild Life (Protection) Act, 1972** (WLPA) — as comprehensively amended by the Wild Life (Protection) Amendment Act 2022 — is India's primary wildlife legislation. **Schedule VI** lists **"Specified Plants"** for which collection, uprooting, damage, trade, and possession are prohibited without a licence from the Chief Wildlife Warden of the relevant state.

Key provisions:
- **Section 17A:** No person shall uproot, damage, collect, acquire or possess any Scheduled plant (other than for personal possession/cultivation without trade intent).
- **Section 17C:** No person shall deal in, sell, or offer for sale any Scheduled plant or its derivative without a licence.
- **Section 51 (penalties):** For Schedule VI violations: imprisonment up to **3 years** and/or fine up to **₹25,000** (enhanced by 2022 Amendment).
- **Section 49B:** Import of Scheduled species requires CITES documentation and permission of Director General of Foreign Trade.

The **2022 Amendment** reorganised the Schedule system (from VI Schedules to IV), consolidating some provisions. Schedule VI equivalent species are now in the new Schedule structure.

Full text: [IndiaCode — WPA 1972](https://www.indiacode.nic.in/bitstream/123456789/1726/2/A1972-53.pdf)

<div class="alert alert-warning" role="alert">
**⚠ Implementation Gap** — Schedule VI lists only approximately **29 species/genera** explicitly. Given India's 47,000+ plant species, this coverage is extremely narrow. The Botanical Survey of India (BSI) has assessed hundreds of additional species as threatened, but these lack Schedule VI protection. The Ministry of Environment, Forest and Climate Change (MoEFCC) has not comprehensively updated the Schedule VI list since the original enactment despite documented declines in many medicinal and horticultural plant species.
</div>

## Biological Diversity Act, 2002 {#bda}

The **Biological Diversity Act, 2002** (Act No. 18 of 2003) implements the Convention on Biological Diversity (CBD) and the Nagoya Protocol. It establishes the National Biodiversity Authority (NBA) to regulate:
- **Access to Biological Resources:** Any person seeking to obtain biological resources for commercial use must obtain prior approval from the NBA (foreigners/foreign entities) or file intimation with State Biodiversity Boards (Indians).
- **Benefit Sharing:** Commercial use of biological resources must be accompanied by benefit sharing with local communities and knowledge holders.
- **Biodiversity Management Committees (BMCs):** Gram Panchayats and urban local bodies must establish BMCs and maintain People's Biodiversity Registers (PBRs).

The **Biological Diversity (Amendment) Act, 2023** made significant changes, exempting traditional practitioners and cultivators from ABS requirements and simplifying the research pathway.

Full text: [WIPO Lex — Biological Diversity Act 2002](https://www.wipo.int/wipolex/en/legislation/details/6058)

## Environment (Protection) Act, 1986 {#epa}

The **Environment (Protection) Act, 1986** (EPA) provides a framework for Central Government notifications covering invasive species and specific environmental restrictions. Several State Pollution Control Boards have issued orders under EPA for control of major invasive plants such as *Parthenium hysterophorus* and *Eichhornia crassipes*.

## CITES Implementation in India {#cites}

India ratified CITES in 1976. The **WCCB (Wildlife Crime Control Bureau)** under MoEFCC is the primary CITES Management Authority. CITES implementation is through Schedule VI (plants) and Schedules I–III (animals) of the WPA 1972.

Key CITES-listed Indian plant groups:
- All Orchidaceae (Appendix II; *Paphiopedilum* spp. Appendix I)
- All Cycadaceae (Appendix I/II)
- *Taxus* spp. (Himalayan yew) — Appendix II
- *Dioscorea deltoidea* — Appendix II
- *Pterocarpus santalinus* (Red Sanders) — Appendix II
- *Saussurea costus* (Kuth) — Appendix I

---

# Regional / Subnational Restrictions {#regional}

India's constitutional framework gives states concurrent legislative authority over forests and wildlife. Each state has a **Chief Wildlife Warden** (CWW) who administers the Wild Life (Protection) Act, a **State Biodiversity Board (SBB)** under the Biological Diversity Act, and a **State Forest Department** with its own rules for forest produce collection. Many states impose restrictions *beyond* the WPA Schedule VI list, particularly for commercially valuable medicinal plants and timber trees.

## State Biodiversity Boards (SBBs) — Biological Diversity Act, 2002

Each of India's 28 states and 8 Union Territories has a **State Biodiversity Board** constituted under the BDA. SBBs regulate commercial access to biological resources within the state. Specific state-level restrictions include:

| State / Region | Key State Instrument | Notable Plant Species Restrictions |
|---------------|---------------------|----------------------------------|
| **Karnataka** | Karnataka Forest Act 1963 (as amended); KFD sandalwood order | **Indian sandalwood** (*Santalum album*): Government owns all naturally grown sandalwood trees on private and public land (Karnataka Sandalwood Act 1970-era amendments); harvesting requires KFD license; sale through KDTDC only |
| **Andhra Pradesh / Telangana** | AP Forest Act 1967; APFD Red Sanders enforcement regime | **Red Sanders** (*Pterocarpus santalinus*): Both states have dedicated task forces; all wild trees are government property; export prohibited without APFD NOC |
| **Himachal Pradesh** | HP Forest Produce Transit Rules 2019; HP Forest Act 1927 | *Taxus wallichiana* (Himalayan yew): Transit of yew bark and needles banned; HP Forest Dept has district-level collection permits for medicinal plants; over 70 plant species on HP state restricted list |
| **Uttarakhand** | UK Forest Policy 2016; Medicinal Plant Board rules | *Saussurea costus*, *Nardostachys jatamansi*, *Podophyllum hexandrum*: State-level harvest quotas; Uttarakhand Forest Development Corporation controls commercial collection |
| **Sikkim** | Sikkim Forest Act 2023; Sikkim SBB rules | State declared an organic state — all synthetic pesticide use banned; strict collection restrictions on high-altitude medicinal plants; *Fritillaria cirrhosa* and orchids specifically listed |
| **Meghalaya / Nagaland / Mizoram** | NE States Forest Acts; Customary law | Tribal customary laws (*Ri* / *Ilam* systems) restrict access to community forests; state SBBs work with village biodiversity registers; orchid collection heavily regulated |
| **Kerala** | Kerala Forest Act 1961; KFRI guidelines | Specific extraction quotas for 168 medicinal plant species under Kerala Medicinal Plants Board (KMPB); Cardamom Hills Reserve species specially protected |
| **Tamil Nadu** | TN Forest Act 1882 (revised); TNPCB rules | *Pterocarpus santalinus*, red sanders endemic range in TN border districts; state SBB ABS framework adds 15 state-specific species to national list |
| **Rajasthan / Gujarat** | State Forest Acts; CAZRI guidelines for arid zones | *Prosopis cineraria* (khejri — state tree of Rajasthan) protected; cutting without permit under Forest Act; state-level Bishnoi community protection has customary legal recognition |
| **Maharashtra** | MoEFCC-MPCB convergence; Mumbai Mangrove rules | Mangrove tree species: Maharashtra enacted specific mangrove protection orders; Mumbai HC rulings have given mangroves virtual PA status; state forest dept maintains mangrove cell |

## Chief Wildlife Warden Permits — State-Level Licensing

For WPA Schedule VI species, **state CWW permits** are required for:
- Collection for bona fide scientific research
- Cultivation under CITES documentation
- Export/import for approved purposes

CWW permit authority means enforcement is **state-specific**: a permit from Karnataka does not cover collection in Himachal Pradesh for the same species. This creates significant inter-state trade regulatory complexity.

## People's Biodiversity Registers (PBRs) — Village-Level Restrictions

Under BDA Section 41, Gram Panchayats and urban local bodies must establish **Biodiversity Management Committees (BMCs)** and maintain **People's Biodiversity Registers (PBRs)** documenting local biological resources. BMCs may levy a **benefit-sharing charge** on commercial users of locally documented resources. Several states (Kerala, Karnataka, Sikkim) have advanced PBR frameworks with >1,000 registers documented that effectively create village-level species access controls.

<div class="alert alert-info" role="alert">
**Source note** — State SBB portals: [NBA India — State Biodiversity Boards](https://www.nbaindia.org/state-bdb.html) | Karnataka KFD: [kfd.gov.in](https://www.kfd.gov.in) | Uttarakhand Medicinal Plant Board: [umb.uk.gov.in](http://umb.uk.gov.in) | Kerala KMPB: [kmpb.kerala.gov.in](https://kmpb.kerala.gov.in).
</div>

---

# Regulatory Institutions {#institutions}

```{r institutions-table}
tribble(
  ~`Abbrev.`, ~`Institution`,                                    ~`Ministry`,    ~`Mandate`,
  "MoEFCC",   "Ministry of Environment, Forest and Climate Change","GoI",        "Central regulator; WPA administration; CITES authority; NBA oversight",
  "WCCB",     "Wildlife Crime Control Bureau",                   "MoEFCC",       "Central intelligence and enforcement; coordinates interstate WPA/CITES cases",
  "BSI",      "Botanical Survey of India",                       "MoEFCC",       "Botanical inventory; Red List assessments; flora of India; Schedule VI science",
  "NBA",      "National Biodiversity Authority",                  "MoEFCC",       "ABS regulation; bioprospecting approvals; Nagoya Protocol implementation",
  "Forest Dept","State Forest Departments (28 states + 8 UTs)",  "State Govts",  "WPA enforcement; Chief Wildlife Wardens; TOPS permits; reserved forest management",
  "NBWL",     "National Board for Wildlife",                      "MoEFCC",       "Policy oversight; protected area management; species conservation plans",
  "FSI",      "Forest Survey of India",                           "MoEFCC",       "National forest inventory; remote sensing; forest cover mapping",
  "ICAR",     "Indian Council of Agricultural Research",          "DARE",         "Invasive species research; plant protection; crop pest and weed management"
) |>
  styled_kable(caption = "Table 1. Institutions with Plant Regulatory Mandates in India", height = "380px")
```

---

# Comprehensive Plant Species Listing {#species}

```{r species-overview-fig, fig.cap="India restricted plant species by legal category."}
species |>
  mutate(category = case_when(
    str_detect(restriction_type, regex("invasive|eradication|control", ignore_case=TRUE)) ~ "Invasive Species",
    str_detect(governing_law, "WPA") ~ "WPA Schedule VI",
    TRUE ~ "Other Protected"
  )) |>
  count(category) |>
  mutate(category = fct_reorder(category, n)) |>
  ggplot(aes(x=n, y=category, fill=category)) +
  geom_col(width=0.6, show.legend=FALSE) +
  geom_text(aes(label=n), hjust=-0.2, size=4) +
  scale_x_continuous(expand=expansion(mult=c(0,0.2))) +
  scale_fill_brewer(palette="Set2") +
  labs(title="India Restricted Plant Species by Legal Category",
       subtitle=glue("Total: {n_species} species"),
       x="Number of species", y=NULL) +
  theme_minimal(base_size=12) +
  theme(panel.grid.major.y=element_blank())
```

## Master Species Table

```{r master-table}
species |>
  arrange(governing_law, iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`, `Group`=taxonomic_group,
         `IUCN`=iucn_status, `CITES`=cites_appendix,
         `India Legal Status`=legal_status_india,
         `Managing Inst.`=managing_institution, `Notes`=notes) |>
  styled_kable(caption="Table 2. Restricted/Protected Plant Species in India",
               height="600px")
```

## Schedule VI — Specified Plants

The Wild Life (Protection) Act Schedule VI lists approximately 29 plant taxa for which collection, trade, and possession without a licence from the Chief Wildlife Warden is prohibited. The list heavily emphasises medicinally and horticulturally significant species facing exploitation pressure.

```{r sch6-table}
species |>
  filter(str_detect(governing_law, "WPA")) |>
  arrange(iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`, `IUCN`=iucn_status, `CITES`=cites_appendix,
         `Notes`=notes) |>
  styled_kable(caption=glue("Table 3. WPA 1972 Schedule VI Plant Species ({n_sch6} listed)"), height="500px")
```

## Invasive Species

India's invasive plant crisis is driven by rapid spread in disturbed habitats, agricultural areas, and protected areas. The Working Group on Invasive Alien Species (under the National Board for Wildlife) has documented major threats from *Parthenium*, *Lantana*, *Prosopis*, and *Eichhornia*.

```{r invasive-table}
species |>
  filter(str_detect(restriction_type, regex("invasive|control|eradication", ignore_case=TRUE))) |>
  arrange(common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`, `India Status`=legal_status_india,
         `Notes`=notes) |>
  styled_kable(caption=glue("Table 4. Invasive Plant Species in India ({n_invasive} listed)"), height="350px")
```

## GBIF Backbone Resolution {#gbif}

```{r gbif-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    mutate(
      gbif_url = if_else(!is.na(GBIFusageKey) & GBIFusageKey!="",
                         glue("https://www.gbif.org/species/{GBIFusageKey}"), NA_character_),
      iucn_url = paste0("https://www.iucnredlist.org/search?query=",
                        URLencode(str_squish(coalesce(prefName,"")), reserved=TRUE)),
      `Scientific Name` = if_else(!is.na(gbif_url),
                                   glue("<a href='{gbif_url}' target='_blank'><em>{prefName}</em></a>"),
                                   glue("<em>{coalesce(prefName,'')}</em>")),
      `IUCN` = glue("<a href='{iucn_url}' target='_blank'>IUCN &#x2197;</a>")
    ) |>
    transmute(`#`=row_number(), `Scientific Name`, `Common Name`=coalesce(englishName,""),
              `Family`=coalesce(family,""), `GBIF Key`=coalesce(GBIFusageKey,""),
              `India Status`=coalesce(legal_status_india,""),
              `Synonyms`=str_trunc(coalesce(synonyms,""),80), `IUCN`) |>
    styled_kable(caption=glue("Table 5. India Regulated Plants — GBIF Backbone Resolved ({n_gbif_resolved} of {n_species} matched | {n_gbif_families} families)"),
                 height="600px")
} else { cat("*GBIF data will be generated on first render with internet access.*") }
```

---

# Regulatory Gaps {#gaps}

```{r gaps-table}
tribble(
  ~`#`, ~`Gap Area`,                                           ~`Severity`, ~`Status (2025)`,
  "1",  "Schedule VI covers ~29 taxa; thousands unthreatened", "High",      "Unresolved",
  "2",  "BSI Red List assessments not triggering WPA updates",  "High",      "Unresolved",
  "3",  "Red Sanders smuggling persists despite Schedule VI",   "High",      "Partial enforcement",
  "4",  "Medicinal plant over-collection from wild (NTFPs)",    "High",      "Unresolved",
  "5",  "BDA ABS implementation — benefit sharing rarely occurs","High",     "BDA Amendment 2023",
  "6",  "Invasive species legislation fragmented across states", "Medium",   "Unresolved",
  "7",  "Orchid trade enforcement: nursery vs wild unclear",    "Medium",    "Unresolved",
  "8",  "Traditional knowledge protections weak",               "Medium",    "BDA Amendment 2023"
) |>
  kbl(caption="Table 6. Regulatory Gaps for Plant Species in India", escape=FALSE, align="l") |>
  kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"), full_width=TRUE, font_size=12) |>
  column_spec(3, bold=TRUE, color=ifelse(c(TRUE,TRUE,TRUE,TRUE,TRUE,FALSE,FALSE,FALSE),"white","black"),
              background=ifelse(c(TRUE,TRUE,TRUE,TRUE,TRUE,FALSE,FALSE,FALSE),"#c0392b","#e67e22"))
```

## Schedule VI Coverage Gap

India's Schedule VI is one of the most narrowly scoped protected plant lists among megadiverse countries. With ~47,500 species of vascular plants recorded in India (BSI National Database, 2024), and the Botanical Survey of India assessing over 1,200 species as Threatened or Near Threatened, the WPA list of approximately 29 taxa offers protection to fewer than 0.1% of India's plant species. The legal mechanism to add species exists but requires a Central Government amendment — a slow process without any statutory review timetable.

## Red Sanders Trafficking

Red sanders (*Pterocarpus santalinus*), endemic to the Seshachalam hills of Andhra Pradesh, remains among the world's most heavily trafficked plant commodities. Despite Schedule VI listing, CITES Appendix II listing, and active seizures by the Andhra Pradesh Forest Department, large-scale smuggling (primarily to China, Hong Kong, and Japan) persists. Between 2015 and 2022, Indian courts processed hundreds of red sanders cases, but organised crime networks continue to operate, often exploiting porous port oversight at Chennai and Visakhapatnam.

---

# Key References {#references}

| Resource | Link |
|----------|------|
| Wild Life (Protection) Act 1972 — IndiaCode | [indiacode.nic.in](https://www.indiacode.nic.in/bitstream/123456789/1726/2/A1972-53.pdf) |
| WPA Amendment Act 2022 | [MoEFCC](https://www.moef.gov.in) |
| Biological Diversity Act 2002 — WIPO Lex | [wipo.int](https://www.wipo.int/wipolex/en/legislation/details/6058) |
| National Biodiversity Authority | [nbaindia.org](https://www.nbaindia.org) |
| Botanical Survey of India | [bsi.gov.in](https://www.bsi.gov.in) |
| Wildlife Crime Control Bureau | [wccb.gov.in](https://www.wccb.gov.in) |
| GBIF India | [gbif.org/country/IN](https://www.gbif.org/country/IN/summary) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm <- kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',
  if(nzchar(logo_sm)) paste0(logo_sm,'<br/>') else "",
  '<p style="font-weight:600;color:#0d2240;">', kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; For informational purposes only &nbsp;&bull;&nbsp; Verify at MoEFCC / IndiaCode / BSI</p>',
  '</div>'))
```
