---
title: "Restricted and Protected Plant Species in Chile"
subtitle: "Ley 19.473 (Hunting Law 1996), DS No. 13/2013 (Species Classification), Ley 18.362 (SNASPE) — Preliminary Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})
.kqb_pkg <- "/Users/kwizera.jvk/Desktop/Kwiz Computing Comes to Life/Kwiz Quants/libs/kwizbranding"
suppressMessages(devtools::load_all(.kqb_pkg, quiet=TRUE))
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"docs","www"),file.path(getwd(),"www"),file.path(getwd()),file.path(here::here()))
.kqb_load_logo <- function(fn,ck){for(d in .kqb_logo_candidates){p<-file.path(d,fn);if(file.exists(p)){assign(ck,paste0("data:image/svg+xml;base64,",base64enc::base64encode(readBin(p,"raw",file.size(p)))),envir=.GlobalEnv);return(invisible(TRUE))}}}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")
styled_kable <- function(df,caption=NULL,scroll=TRUE,height="450px"){k<-df|>kbl(caption=caption,escape=FALSE,align="l")|>kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"),full_width=TRUE,font_size=12);if(scroll)k<-scroll_box(k,height=height);k}

species <- tribble(
  ~common_name,              ~scientific_name,            ~taxonomic_group, ~iucn_status,            ~cites_appendix,  ~legal_status_cl,                              ~restriction_type,              ~governing_law,                    ~managing_institution,   ~notes,
  "Monkey puzzle tree",      "Araucaria araucana",         "Pinopsida",      "Endangered",            "Appendix I",     "National Monument — Strictly Protected (CITES App. I)","Felling/collection prohibited","Ley 19.473 / DS 43/1990 / CITES", "CONAF / SAG",           "Pehuén; Chile-Argentina endemic; sacred to Mapuche/Pehuenche peoples; 1,000+ yr lifespan; illegal logging persists",
  "Alerce",                  "Fitzroya cupressoides",      "Pinopsida",      "Endangered",            "Appendix I",     "National Monument — Strictly Protected (CITES App. I)","Felling/trade prohibited",     "DS 490/1976 / CITES",             "CONAF",                 "Alerce / Patagonian cypress; up to 3,600 yr lifespan; one of world's oldest trees; protected since 1976; past logging severe",
  "Juan Fernandez tree fern", "Thyrsopteris elegans",      "Polypodiopsida", "Endangered",            "Not listed",     "Endangered (DS 13/2013) — Juan Fernandez Endemic","Collection prohibited",        "DS 13/2013 / Ley 18.362",         "CONAF",                 "Sole member of family Thyrsopteridaceae; Juan Fernandez archipelago; world's rarest tree fern; <1,000 plants",
  "Chilean wine palm",       "Jubaea chilensis",           "Magnoliopsida",  "Vulnerable",            "Not listed",     "Vulnerable (DS 13/2013) — protected",         "Felling prohibited; harvest regulated","DS 13/2013 / Ley 19.473",    "SAG / CONAF",           "Coquito palm; world's southernmost palm; slow-growing; sap harvest traditionally killed trees; populations fragmented",
  "Peumo",                   "Cryptocarya alba",           "Magnoliopsida",  "Near Threatened",       "Not listed",     "Near Threatened (DS 13/2013) — protected",    "Regulated harvest",            "DS 13/2013",                      "SAG / CONAF",           "Sclerophyllous matorral endemic; fruit and timber uses; protected in national parks; private land exploitation concern",
  "Garra de León",           "Leontochir ovallei",         "Magnoliopsida",  "Endangered",            "Not listed",     "Endangered (DS 13/2013)",                     "Collection prohibited",        "DS 13/2013",                      "SAG / CONAF",           "Chile endemic; Atacama coastal fog zone; only species in genus; bulb collection for horticulture a threat",
  "Palma de Pascua",         "Howea forsteriana",          "Magnoliopsida",  "Least Concern",         "Not listed",     "Monitored — Easter Island flora protected",   "Collection monitored",         "Ley 18.362 (SNASPE)",             "CONAF",                 "Note: Paschalococos disperta (Easter Island palm) is extinct; Howea represents current protected palm species on island",
  "Endangered quillay",      "Quillaja saponaria",         "Magnoliopsida",  "Least Concern",         "Not listed",     "Protected — bark harvest regulated",          "Harvest regulated (bark)",     "DS 13/2013 / SAG regulations",    "SAG",                   "Soapbark tree; bark saponins commercially valuable (vaccines, food industry); harvest permits required",
  "Port Jackson wattle",     "Acacia saligna",             "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — control required",                 "Control mandated",             "SAG / MMA regulations",           "SAG / MMA",             "Major invasive in central Chile matorral; displaces native sclerophyllous shrubs; fire risk; biological control explored",
  "Water hyacinth",          "Eichhornia crassipes",       "Magnoliopsida",  "Least Concern",         "Not listed",     "Invasive — management required",              "Control mandated",             "SAG / MMA regulations",           "SAG / MMA",             "Invasive in waterways of south-central Chile; blocks irrigation channels and disrupts aquatic ecosystems"
)

n_species <- nrow(species)
proj_root <- here::here()
gbif_cache <- file.path(proj_root,"output","tables","chile_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  cn <- function(nm) nm|>str_remove_all("\\s+spp\\.?$")|>str_remove_all("\\(.*?\\)")|>str_squish()
  nd <- tibble(name=cn(species$scientific_name),kingdom="Plantae")
  br <- tryCatch(name_backbone_checklist(nd),error=function(e) map_dfr(nd$name,function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  gv <- function(k){if(is.na(k))return(NA_character_);tryCatch({vn<-name_usage(key=as.numeric(k),data="vernacularNames",limit=100);if(is.null(vn$data)||nrow(vn$data)==0)return(NA_character_);en<-vn$data|>filter(language%in%c("eng","en"))|>pull(vernacularName)|>unique();if(length(en)==0)NA_character_ else paste(en,collapse=", ")},error=function(e)NA_character_)}
  gs <- function(k){if(is.na(k))return(NA_character_);tryCatch({syn<-name_usage(key=as.numeric(k),data="synonyms",limit=100);if(is.null(syn$data)||nrow(syn$data)==0)return(NA_character_);s<-syn$data|>pull(canonicalName)|>unique()|>na.omit();if(length(s)==0)NA_character_ else paste(s,collapse=", ")},error=function(e)NA_character_)}
  bc <- br|>as_tibble()|>mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species|>mutate(cs=cn(scientific_name))|>bind_cols(bc|>select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm|>filter(!is.na(GBIFusageKey))|>distinct(GBIFusageKey)|>pull(GBIFusageKey)
  vl<-list();sl<-list()
  for(i in seq_along(keys)){k<-keys[i];vl[[k]]<-gv(k);sl[[k]]<-gs(k);if(i%%5==0)Sys.sleep(0.2)}
  vd<-tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  en<-sm|>left_join(vd,by="GBIFusageKey")|>mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  en|>select(GBIFusageKey,prefName,family,englishName,synonyms,legal_status_cl,restriction_type,governing_law)|>write_csv(gbif_cache)
}
if(file.exists(gbif_cache)){plants_gbif<-read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE);n_gbif_resolved<-sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="");n_gbif_families<-n_distinct(plants_gbif$family,na.rm=TRUE)}else{plants_gbif<-NULL;n_gbif_resolved<-0L;n_gbif_families<-0L}
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04,size="280px auto"))
uri<-kqb_watermark_uri();if(nzchar(uri))cat(paste0('<style>\nbody::after{\ncontent:"";position:fixed;top:0;left:0;width:100%;height:100%;z-index:9000;pointer-events:none;background-image:url("',uri,'");background-repeat:repeat;background-size:280px auto;opacity:0.04;}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Chile contains approximately **5,100 native plant species**, with an exceptionally high endemism rate of approximately **50%** — largely driven by the country's geographic isolation between the Andes, the Pacific Ocean, the Atacama Desert, and Patagonian ice fields. The Chilean Mediterranean sclerophyllous zone is recognised as a global **biodiversity hotspot**. The legal framework for plant protection centres on the **DS No. 13/2013 species classification system**, which applies IUCN criteria to produce national threatened species lists, administered by CONAF (forestry) and SAG (agriculture and hunting).

<div class="alert alert-info" role="alert">
**Preliminary Report** — Full species listing from DS 13/2013 (species classification system) to be completed. Current report covers `r n_species` representative species. Note: *Araucaria araucana* and *Fitzroya cupressoides* are designated National Monuments and carry the strictest protection under Chilean law.
</div>

---

# Legal Framework (Summary) {#legal-framework}

| Law / Instrument | Year | Key Provision |
|-----------------|------|---------------|
| Ley 19.473 — Hunting Law (Caza) | 1996 | Prohibits hunting/collection of protected species; penalty framework |
| DS No. 13/2013 — Species Classification | 2013 | IUCN-based national threatened species classification system; Reglamento de Clasificación |
| Ley 18.362 — SNASPE / CONAF | 1984 | National System of Protected Wild Areas of the State; national parks and reserves |
| DS 490/1976 | 1976 | Alerce declared National Monument; all felling/trade prohibited |
| DS 43/1990 | 1990 | Araucaria araucana declared National Monument |
| Ley 20.283 — Native Forest Recovery | 2008 | Regulates native forest management and recovery; felling permits |

Full text: [CONAF](https://www.conaf.cl) | [SAG](https://www.sag.gob.cl) | [FAOLEX Chile](https://faolex.fao.org/chile)

---

# Regulatory Institutions {#institutions}

| Institution | Role |
|-------------|------|
| CONAF (Corporación Nacional Forestal) | National parks; native forest law; Araucaria and Alerce protection; permits |
| SAG (Servicio Agrícola y Ganadero) | Hunting law enforcement; agricultural plant imports; invasive species |
| MMA (Ministry of Environment) | Biodiversity policy; species classification system (DS 13/2013); CITES management authority |
| IUCN Chile / Botanical experts | Technical committees for species classification review |
| CONAMA (former — now MMA) | Environmental assessment and policy framework |
| Universidad de Chile / Universidad de Concepción | Botanical taxonomy and herbarium collections |

---

# Preliminary Species Listing {#species}

```{r master-table}
species|>arrange(iucn_status,common_name)|>mutate(`Scientific Name`=paste0("*",scientific_name,"*"))|>
  select(`Common Name`=common_name,`Scientific Name`,`IUCN`=iucn_status,`CITES`=cites_appendix,
         `Chile Status`=legal_status_cl,`Managing Inst.`=managing_institution,`Notes`=notes)|>
  styled_kable(caption="Table 1. Preliminary — Protected Plant Species in Chile",height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if(!is.null(plants_gbif)&&nrow(plants_gbif)>0){plants_gbif|>
  mutate(gu=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
         iu=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
         `Scientific Name`=if_else(!is.na(gu),glue("<a href='{gu}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
         `IUCN`=glue("<a href='{iu}' target='_blank'>IUCN &#x2197;</a>"))|>
  transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
            `GBIF Key`=coalesce(GBIFusageKey,""),`Chile Status`=coalesce(legal_status_cl,""),
            `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`)|>
  styled_kable(caption=glue("Table 2. Chile — GBIF Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} fam.)"),height="450px")
}else{cat("*GBIF data will be generated on first render.*")}
```

---

# Regulatory Gaps (Summary) {#gaps}

| # | Gap | Severity |
|---|-----|----------|
| 1 | Illegal Araucaria and Alerce timber trade continues despite National Monument status | High |
| 2 | Wildfire risk to Araucaria and native matorral — climate-driven fire frequency increasing | High |
| 3 | DS 13/2013 classification reviews slow — many endemic species awaiting formal listing | High |
| 4 | Juan Fernandez Islands — invasive species (rabbits, Acacia) destroying endemic plant habitat | High |
| 5 | Chilean wine palm private land populations have limited legal protection outside parks | Medium |
| 6 | Acacia saligna invasion control underfunded relative to scale of invasion in central Chile | Medium |

---

# References {#references}

| Resource | Link |
|----------|------|
| CONAF Chile | [conaf.cl](https://www.conaf.cl) |
| SAG — Servicio Agrícola y Ganadero | [sag.gob.cl](https://www.sag.gob.cl) |
| MMA — Ministry of Environment | [mma.gob.cl](https://www.mma.gob.cl) |
| GBIF Chile | [gbif.org/country/CL](https://www.gbif.org/country/CL/summary) |
| CITES Chile | [cites.org — Chile](https://cites.org/eng/cms/index.php/component/cp/country/CL) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm<-kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm))paste0(logo_sm,'<br/>')else"",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Preliminary — Verify at CONAF / SAG / MMA / CITES</p></div>'))
```
