---
title: "Restricted and Protected Plant Species in Indonesia"
subtitle: "Law No. 5/1990, Government Regulation 7/1999, and CITES — Comprehensive Reference"
author: "Kwiz Research"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: {collapsed: false, smooth_scroll: true}
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: hide
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", fig.width=9, fig.height=5, out.width="100%")
suppressPackageStartupMessages({
  library(tidyverse); library(kableExtra); library(glue); library(janitor); library(rgbif)
})

if (!requireNamespace("kwizbranding", quietly = TRUE)) {
  remotes::install_github("Kwiz-Computing-Technologies-Limited/kwiz-R-packages",
                          subdir = "kwizbranding", upgrade = "never")
}
library(kwizbranding)
kqb_co <- kqb_company("research"); kqb_col <- kqb_colours()
.kqb_logo_candidates <- c(file.path(here::here(),"src","www"), file.path(getwd(),"www"),
                            file.path(getwd()), file.path(here::here()))
.kqb_load_logo <- function(fn, ck) {
  for (d in .kqb_logo_candidates) { p <- file.path(d,fn)
    if (file.exists(p)) { assign(ck, paste0("data:image/svg+xml;base64,",
      base64enc::base64encode(readBin(p,"raw",file.size(p)))), envir=.GlobalEnv); return(invisible(TRUE)) } }
}
.kqb_load_logo("Kwiz Black Logo Transparent bg.svg",".kqb_wm_cache")
.kqb_load_logo("Kwiz Inverted Color Transparent bg.svg",".kqb_display_cache")

styled_kable <- function(df, caption=NULL, scroll=TRUE, height="450px") {
  k <- df |> kbl(caption=caption, escape=FALSE, align="l") |>
    kable_styling(bootstrap_options=c("striped","hover","condensed","responsive"), full_width=TRUE, font_size=12)
  if (scroll) k <- scroll_box(k, height=height); k
}

# ── Species data ──────────────────────────────────────────────────────────────
# Sources: Government Regulation No. 7/1999 (Preservation of Plant and Animal Species);
#          Government Regulation No. 106/2018 (updated list); CITES CoP19 2022.
species <- tribble(
  ~common_name,               ~scientific_name,             ~taxonomic_group, ~iucn_status,           ~cites_appendix, ~legal_status_id,                  ~restriction_type,          ~governing_law,          ~managing_institution,  ~notes,
  # ── PP 7/1999 Protected Flora ──────────────────────────────────────────────────
  "Rafflesia",                "Rafflesia arnoldii",         "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Collection/trade prohibited","GR No. 7/1999",         "KLHK / BKSDA",         "World's largest flower; Sumatra endemic; obligate parasite",
  "Titan arum / corpse flower","Amorphophallus titanum",    "Magnoliopsida",  "Endangered",           "Appendix II",   "Protected Species (GR 7/1999)",    "Collection/trade regulated", "GR No. 7/1999 / CITES", "KLHK / BKSDA",         "Largest unbranched inflorescence; Sumatra endemic; extreme rarity",
  "Padma raksasa (Rafflesia patma)","Rafflesia patma",      "Magnoliopsida",  "Critically Endangered","Not listed",    "Protected Species (GR 7/1999)",    "Collection/trade prohibited","GR No. 7/1999",         "KLHK / BKSDA",         "Java-endemic Rafflesia; Pangandaran and Ujung Kulon; fewer than 50 known plants",
  "Bunga bangkai Sumatra",    "Amorphophallus gigas",       "Magnoliopsida",  "Endangered",           "Appendix II",   "Protected Species (GR 7/1999)",    "Collection/trade regulated", "GR No. 7/1999 / CITES", "KLHK / BKSDA",         "Giant Sumatran corpse flower; taller than A. titanum; North Sumatra highlands",
  "Keruing",                  "Dipterocarpus retusus",      "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Harvest regulated",         "GR No. 7/1999 / Forestry","KLHK / BKSDA",        "PP 7/1999 Appendix; major dipterocarp timber tree; Java and Sumatra",
  "Damar (cat's-eye resin)",  "Hopea odorata",              "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Harvest regulated",         "GR No. 7/1999",         "KLHK / BKSDA",         "PP 7/1999 listed; valued resin-producing dipterocarp; lowland forests",
  "Kepuh / sterculia",        "Sterculia foetida",          "Magnoliopsida",  "Least Concern",        "Not listed",    "Protected Species (GR 7/1999)",    "Collection regulated",      "GR No. 7/1999",         "KLHK / BKSDA",         "PP 7/1999 protected; culturally important shade tree; Java and Nusa Tenggara",
  "Sawo kecik",               "Manilkara kauki",            "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Collection/trade prohibited","GR No. 7/1999",         "KLHK / BKSDA",         "PP 7/1999; sacred tree in Javanese culture; population severely depleted",
  "Rasamala",                 "Altingia excelsa",           "Magnoliopsida",  "Near Threatened",      "Not listed",    "Protected Species (GR 7/1999)",    "Harvest regulated",         "GR No. 7/1999",         "KLHK / BKSDA",         "PP 7/1999; flagship montane tree of West Java; resin tree",
  "Bayur",                    "Pterospermum javanicum",     "Magnoliopsida",  "Least Concern",        "Not listed",    "Protected Species (GR 7/1999)",    "Collection regulated",      "GR No. 7/1999",         "KLHK / BKSDA",         "PP 7/1999 listed; large canopy tree; Java and Bali",
  # ── CITES-listed Indonesian plants ─────────────────────────────────────────────
  "Slipper orchid (Borneo)",  "Paphiopedilum rothschildianum","Magnoliopsida","Critically Endangered","Appendix I",    "Protected Species (GR 7/1999)",    "Collection/trade prohibited","GR No. 7/1999 / CITES", "KLHK / BKSDA",         "Rarest slipper orchid; Kinabalu-adjacent Kalimantan populations",
  "Slipper orchid (Sumatra)", "Paphiopedilum supardii",     "Magnoliopsida",  "Critically Endangered","Appendix I",    "Protected Species (GR 7/1999)",    "Collection/trade prohibited","GR No. 7/1999 / CITES", "KLHK / BKSDA",         "Kalimantan endemic; presumed extinct in wild; CITES Appendix I",
  "Slipper orchid (Java)",    "Paphiopedilum javanicum",    "Magnoliopsida",  "Endangered",           "Appendix I",    "Protected Species (GR 7/1999)",    "Collection/trade prohibited","GR No. 7/1999 / CITES", "KLHK / BKSDA",         "Javanese endemic slipper orchid; montane forests; all Paphiopedilum spp. CITES I",
  "Borneo orchid (black)",    "Coelogyne pandurata",        "Magnoliopsida",  "Least Concern",        "Appendix II",   "Protected (CITES Appendix II)",     "Collection permit required", "CITES Regs / GR 7/1999","KLHK / BKSDA",         "Black orchid of Borneo; highly prized horticulturally",
  "Moon orchid",              "Phalaenopsis amabilis",      "Magnoliopsida",  "Least Concern",        "Appendix II",   "Protected (CITES Appendix II)",     "Export permit required",    "CITES Regs / GR 7/1999","KLHK / BKSDA",         "National flower of Indonesia (Puspa Pesona); all Orchidaceae CITES II",
  "Vanda tricolour",          "Vanda tricolor",             "Magnoliopsida",  "Least Concern",        "Appendix II",   "Protected Species (GR 7/1999)",    "Collection permit required", "GR No. 7/1999 / CITES", "KLHK / BKSDA",         "PP 7/1999 protected orchid; Java and Bali; highly sought by collectors",
  "Pitcher plant (Sumatra)",  "Nepenthes sumatrana",        "Magnoliopsida",  "Vulnerable",           "Appendix II",   "Protected (CITES Appendix II)",     "Export permit required",    "CITES Regs / GR 7/1999","KLHK / BKSDA",         "Sumatran endemic pitcher plant; all Nepenthes spp. CITES II",
  "Pitcher plant (Rajah-type)","Nepenthes rajah",           "Magnoliopsida",  "Endangered",           "Appendix II",   "Protected (CITES Appendix II)",     "Collection/trade regulated", "CITES Regs",            "KLHK / BKSDA",         "Borneo highland; largest pitchers of any Nepenthes; overcollected",
  "Pitcher plant (Kalimantan)","Nepenthes lowii",           "Magnoliopsida",  "Least Concern",        "Appendix II",   "Protected (CITES Appendix II)",     "Export permit required",    "CITES Regs",            "KLHK / BKSDA",         "Borneo endemic; mutualism with tree shrews; all Nepenthes CITES II",
  "Agarwood (gaharu)",        "Aquilaria malaccensis",      "Magnoliopsida",  "Critically Endangered","Appendix II",   "Protected Species (GR 7/1999)",    "Harvest/trade regulated",   "GR No. 7/1999 / CITES", "KLHK / BKSDA",         "Major Indonesian export; overharvested for oud/incense; Sumatra and Kalimantan",
  "Agarwood (Kalimantan)",    "Aquilaria microcarpa",       "Magnoliopsida",  "Vulnerable",           "Appendix II",   "Protected (CITES Appendix II)",     "Harvest/trade regulated",   "CITES Regs / Forestry",  "KLHK / BKSDA",         "Borneo endemic agarwood species; all Aquilaria spp. CITES II since 2004",
  "Ramin",                    "Gonystylus bancanus",        "Magnoliopsida",  "Vulnerable",           "Appendix II",   "Protected Timber (CITES App II)",   "Export quota / permit",     "CITES Regs / Forestry",  "KLHK / BKSDA",         "Peat swamp timber; Indonesia zero export quota since 2001; major smuggling target",
  "Sumatra cycad",            "Cycas rumphii",              "Cycadopsida",    "Least Concern",        "Appendix II",   "Protected (CITES Appendix II)",     "Collection/trade regulated", "CITES Regs / GR 7/1999","KLHK",                 "All Cycas spp. CITES II; Indonesian populations heavily collected",
  "Coelacanth palm",          "Livistona benthamii",        "Liliopsida",     "Vulnerable",           "Appendix II",   "Protected (CITES Appendix II)",     "Export permit required",    "CITES Regs",            "KLHK",                 "All palms (Palmae) listed under CITES Appendix II",
  # ── Indonesian endemic threatened species ──────────────────────────────────────
  "Sulawesi ironwood (ebony)","Diospyros celebica",         "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Timber Species",          "Harvest/export regulated",  "Forestry Law / GR 7/1999","KLHK / BKSDA",        "Sulawesi endemic; Makassar ebony; extremely slow growth; heavy illegal logging",
  "Borneo ironwood (ulin)",   "Eusideroxylon zwageri",      "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Harvest/trade regulated",   "GR No. 7/1999 / Forestry","KLHK / BKSDA",        "PP 7/1999 protected; Borneo and Sumatra; prized construction timber; slow-growing",
  "Papua kauri / agathis",    "Agathis labillardieri",      "Pinopsida",      "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Harvest regulated",         "GR No. 7/1999 / Forestry","KLHK / BKSDA",        "Papua endemic conifer; resin (copal) tree; heavily logged",
  "Cendana / Timor sandalwood","Santalum album",            "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Timber Species",          "Harvest/trade regulated",   "NTT Prov. Reg. / Forestry","KLHK / Prov. NTT",   "East Nusa Tenggara provincial monopoly timber; severe overexploitation",
  "Meranti (white)",          "Shorea bracteolata",         "Magnoliopsida",  "Endangered",           "Not listed",    "Protected Timber Species",          "Harvest/export regulated",  "Forestry Laws",          "KLHK / KPHP",          "Sumatra and Kalimantan lowland dipterocarp; major timber species; habitat loss",
  "Meranti (red)",            "Shorea leprosula",           "Magnoliopsida",  "Endangered",           "Not listed",    "Protected Timber Species",          "Harvest/export regulated",  "Forestry Laws",          "KLHK / KPHP",          "Dominant lowland dipterocarp; Sumatra and Kalimantan; plantation candidate",
  "Jelutung",                 "Dyera costulata",            "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Harvest regulated",         "GR No. 7/1999 / Forestry","KLHK / BKSDA",        "PP 7/1999 protected; peat swamp forests; latex tree; Sumatra and Kalimantan",
  "Cempaka / magnolia (Sulawesi)","Magnolia liliifera",     "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Species (GR 7/1999)",    "Collection regulated",      "GR No. 7/1999",         "KLHK / BKSDA",         "PP 7/1999 protected; culturally significant; Sulawesi and Maluku",
  "Kayu hitam Sulawesi",      "Diospyros pilosanthera",     "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Timber Species",          "Harvest/export regulated",  "Forestry Laws",          "KLHK / BKSDA",         "Sulawesi endemic ebony; limited range; illegally traded for musical instruments",
  # ── Timber species under Ministerial Decree ────────────────────────────────────
  "Teak (jati)",              "Tectona grandis",            "Magnoliopsida",  "Least Concern",        "Not listed",    "Regulated Timber Species",          "Harvest/transport permit",  "Forestry Laws / Perhutani","Perum Perhutani / KLHK","Java plantation monopoly by Perhutani; wild teak overexploited in Sulawesi and NTT",
  "Borneo rosewood",          "Dalbergia latifolia",        "Magnoliopsida",  "Vulnerable",           "Appendix II",   "Protected Timber (CITES App II)",   "Export permit required",    "CITES Regs / Forestry",  "KLHK / BKSDA",         "All Dalbergia Appendix II since 2017; Java and Sulawesi; major export commodity",
  "Merbau",                   "Intsia bijuga",              "Magnoliopsida",  "Vulnerable",           "Not listed",    "Protected Timber Species",          "Harvest/export regulated",  "Forestry Laws",          "KLHK / KPHP",          "Highly valued Pacific timber; widespread illegal logging in Maluku and Papua",
  "Sono keling (Indian rosewood)","Dalbergia sissooides",   "Magnoliopsida",  "Endangered",           "Appendix II",   "Protected Timber (CITES App II)",   "Export permit required",    "CITES Regs / Forestry",  "KLHK / BKSDA",         "All Dalbergia CITES II; Java furniture trade; heavily depleted wild populations",
  "Kayu manis / cinnamon (wild)","Cinnamomum burmanni",     "Magnoliopsida",  "Least Concern",        "Not listed",    "Regulated Harvest Species",         "Harvest permit required",   "Forestry Laws",          "KLHK / KPHP",          "Wild populations regulated; major plantation crop in West Sumatra (Kerinci)",
  "Tengkawang / illipe nut",  "Shorea macrophylla",         "Magnoliopsida",  "Endangered",           "Not listed",    "Protected Species (GR 7/1999)",    "Harvest regulated",         "GR No. 7/1999 / Forestry","KLHK / BKSDA",        "PP 7/1999 protected; Kalimantan NTFP; fat-producing dipterocarp",
  # ── Invasive Alien Species ─────────────────────────────────────────────────────
  "Acacia nilotica (prickly acacia)","Acacia nilotica",     "Magnoliopsida",  "Least Concern",        "Not listed",    "Priority Invasive Species",         "Eradication mandated",      "KLHK Decree / NP regulations","KLHK / Baluran NP", "Invaded >50% of Baluran NP savanna (East Java); major removal programmes since 1990s",
  "Water hyacinth",           "Eichhornia crassipes",       "Liliopsida",     "Least Concern",        "Not listed",    "Priority Invasive Species",         "Control mandated",          "Law 32/2024 / GR 60/2024","KLHK / BNPB",         "Major invasive in Sulawesi and Java freshwaters; clogs Lake Tempe, Rawapening",
  "Mikania vine",             "Mikania micrantha",          "Magnoliopsida",  "Least Concern",        "Not listed",    "Invasive Species",                  "Control in protected areas","KLHK regulations",      "KLHK / BKSDA",         "Mile-a-minute vine; major threat in disturbed forests Kalimantan and Sulawesi",
  "Giant sensitive plant",    "Mimosa pigra",               "Magnoliopsida",  "Least Concern",        "Not listed",    "Invasive Species",                  "Control in protected areas","KLHK regulations",      "KLHK / BKSDA",         "Aggressive wetland invader; Way Kambas NP (Lampung); displaces native vegetation",
  "Lantana",                  "Lantana camara",             "Magnoliopsida",  "Least Concern",        "Not listed",    "Invasive Species",                  "Control in protected areas","KLHK regulations",      "KLHK / BKSDA",         "Widespread invasive shrub; degraded forests Java, Bali, NTT; toxic to livestock"
)

n_species <- nrow(species); n_cr <- sum(species$iucn_status=="Critically Endangered",na.rm=TRUE)

proj_root  <- here::here()
gbif_cache <- file.path(proj_root, "output", "tables", "indonesia_regulated_plants_gbif.csv")
if (!file.exists(gbif_cache)) {
  clean_name <- function(nm) nm |> str_remove_all("\\s+spp\\.?$") |> str_remove_all("\\(.*?\\)") |> str_squish()
  names_df <- tibble(name=clean_name(species$scientific_name), kingdom="Plantae")
  backbone_res <- tryCatch(name_backbone_checklist(names_df), error=function(e)
    map_dfr(names_df$name, function(nm) tryCatch(as_tibble(name_backbone(name=nm,kingdom="Plantae")),
      error=function(e2) tibble(usageKey=NA_integer_,canonicalName=nm,matchType="NONE"))))
  get_vern <- function(key) { if(is.na(key)) return(NA_character_)
    tryCatch({ vn <- name_usage(key=as.numeric(key),data="vernacularNames",limit=100); if(is.null(vn$data)||nrow(vn$data)==0) return(NA_character_)
      en <- vn$data |> filter(language %in% c("eng","en")) |> pull(vernacularName) |> unique(); if(length(en)==0) NA_character_ else paste(en,collapse=", ") }, error=function(e) NA_character_) }
  get_syn <- function(key) { if(is.na(key)) return(NA_character_)
    tryCatch({ syn <- name_usage(key=as.numeric(key),data="synonyms",limit=100); if(is.null(syn$data)||nrow(syn$data)==0) return(NA_character_)
      s <- syn$data |> pull(canonicalName) |> unique() |> na.omit(); if(length(s)==0) NA_character_ else paste(s,collapse=", ") }, error=function(e) NA_character_) }
  bc <- backbone_res |> as_tibble() |> mutate(GBIFusageKey=as.character(usageKey),prefName=coalesce(species,canonicalName,verbatim_name),gbif_family=family)
  sm <- species |> mutate(cs=clean_name(scientific_name)) |> bind_cols(bc |> select(GBIFusageKey,prefName,gbif_family,gbif_matchType=matchType,gbif_confidence=confidence))
  keys <- sm |> filter(!is.na(GBIFusageKey)) |> distinct(GBIFusageKey) |> pull(GBIFusageKey)
  vl <- list(); sl <- list()
  for (i in seq_along(keys)) { k <- keys[i]; vl[[k]] <- get_vern(k); sl[[k]] <- get_syn(k); if(i%%5==0) Sys.sleep(0.2) }
  vd <- tibble(GBIFusageKey=names(vl),gbif_english=unlist(vl),gbif_synonyms=unlist(sl[names(vl)]))
  enriched <- sm |> left_join(vd,by="GBIFusageKey") |> mutate(englishName=coalesce(gbif_english,common_name),synonyms=gbif_synonyms,family=gbif_family)
  dir.create(file.path(proj_root,"output","tables"),recursive=TRUE,showWarnings=FALSE)
  enriched |> mutate(country="Indonesia",region="",jurisdiction="NATIONAL",jurisdiction_group="",
    classification=taxonomic_group,taxonLevel="species",Note=notes,genus_I=word(prefName,1),is_aquatic="no",aquatic_type_I="") |>
    select(GBIFusageKey,country,region,jurisdiction,jurisdiction_group,prefName,classification,taxonLevel,family,englishName,synonyms,Note,genus_I,is_aquatic,aquatic_type_I,legal_status_id,restriction_type,governing_law) |> write_csv(gbif_cache)
}
if (file.exists(gbif_cache)) {
  plants_gbif <- read_csv(gbif_cache,col_types=cols(.default=col_character()),show_col_types=FALSE)
  n_gbif_resolved <- sum(!is.na(plants_gbif$GBIFusageKey)&plants_gbif$GBIFusageKey!="")
  n_gbif_families <- n_distinct(plants_gbif$family,na.rm=TRUE)
} else { plants_gbif <- NULL; n_gbif_resolved <- 0L; n_gbif_families <- 0L }
```

```{r kqb-watermark, results='asis', echo=FALSE}
cat(kqb_watermark_css(opacity=0.04, size="280px auto"))
uri <- kqb_watermark_uri()
if (nzchar(uri)) cat(paste0('<style>\nbody::after {\n  content: "";\n  position: fixed;\n',
  '  top:0; left:0; width:100%; height:100%;\n  z-index:9000;\n  pointer-events:none;\n',
  '  background-image: url("', uri, '");\n  background-repeat:repeat;\n  background-size:280px auto;\n',
  '  opacity:0.04;\n}\n</style>\n'))
```

---

# Executive Summary {.unnumbered}

Indonesia is the world's largest archipelagic nation and a **megadiverse** country hosting approximately **28,000 plant species** — roughly 11% of global flora — across Sumatra, Java, Kalimantan, Sulawesi, and Papua. It is the top ASEAN country on GBIF, with Herbarium Bogoriense (nearly 1 million specimens) as a major global botanical dataset.

<div class="alert alert-info" role="alert">
This report covers **`r n_species` legally restricted plant species** listed under KLHK (Ministry of Environment and Forestry) Government Regulation 7/1999 and updated GR 106/2018.
</div>

This report covers **`r n_species` legally restricted plant species** (`r n_cr` Critically Endangered). Indonesia's primary legislation for protected species is **Government Regulation No. 7/1999** on Preservation of Plant and Animal Species, updated by **GR No. 106/2018**, under the authority of **Law No. 5/1990** on Conservation of Biological Resources and Ecosystems. The updated **Law No. 32/2024** further strengthened conservation provisions.

---

# Legal Framework (Summary) {#legal-framework}

| Law | Year | Key Provision |
|-----|------|---------------|
| Law No. 5/1990 — Conservation of Biological Resources | 1990 | Framework law; defines protected species categories; penalties for violations |
| Government Regulation No. 7/1999 — Protected Species | 1999 | Lists ~75 protected plant species; collection/trade prohibited |
| Government Regulation No. 106/2018 — Updated List | 2018 | Updated and expanded protected species list |
| Law No. 41/1999 — Forestry | 1999 | Forest conservation; timber regulation; protected forest areas |
| Law No. 32/2024 — Amendment (Conservation) | 2024 | Strengthened penalties; modern enforcement mechanisms |
| CITES (ratified 1978) | — | Appendix I/II/III species; KLHK as Management Authority |

**Legal database:** [FLEVIN — Indonesian Laws in English](https://www.flevin.com/id/lgso/translations/)

---

# Regional / Subnational Restrictions {#regional}

Indonesia's decentralisation framework (**UU No. 23/2014** on Regional Governance) grants provinces and districts significant authority over natural resource management, including forest use and plant collection. However, national protected species lists under **Government Regulation 7/1999** (updated by PP 106/2018) apply uniformly across Indonesia — provincial regulations may only add further restrictions, not remove them. In practice, the most significant subnational variation arises from provincial forestry permits, customary law (*adat*), and Special Region status.

## Province-Level Key Variations

| Province | Key Instrument | Plant Species Scope |
|---------|---------------|-------------------|
| **Aceh** | Qanun (Sharia-derived provincial regulations); RTRW Aceh | Leuser Ecosystem — additional protection for lowland forest species including *Rafflesia* and pitcher plants (*Nepenthes*); Leuser Development Agency (LDA) has independent conservation authority over the 2.6 million ha Leuser Ecosystem; illegal plant collection prosecuted under Qanun 23/2002 |
| **Papua** | UU Otonomi Khusus 21/2001 (Papua Special Autonomy); UU 2/2021 | Customary lands (*tanah adat*) managed by indigenous *Ondoafi*; commercial plant extraction requires local customary board (*Dewan Adat Papua*) consent in addition to KLHK permits; *Agathis labillardieri* and *Pterocarpus indicus* within adat territories subject to community veto |
| **Papua Barat (West Papua)** | As Papua above; mangrove conservation area regulations | Mangrove species (*Rhizophora*, *Avicennia*) protected through provincial coastal zone plans; New Province: South Papua (Provinsi Papua Selatan, 2022), Central Papua (Papua Tengah) now add additional administrative layers |
| **Kalimantan (all 5 provinces)** | RTRW (Spatial Planning) regulations; Forestry Ministry concession rules | *Shorea* and *Dipterocarpus* timber species — provincial logging moratoria in certain protection forests; Heart of Borneo initiative overlaps with provincial spatial plans; *Dyera costulata* and illipe nut (*Shorea macrophylla*) harvest controls by district forestry offices |
| **Sumatra** | Sumatra Biodiversity Strategy; Harimau Sumatra conservation zone | *Aquilaria malaccensis* (agarwood): provincial anti-trafficking task forces in Aceh, North Sumatra, Riau; *Rafflesia arnoldii* — Bengkulu provincial protection ordinance; Kerinci Seblat landscape regulations restrict collection |
| **Sulawesi** | Wallacea Biodiversity Hotspot framework; provincial RTRW | Endemic species specific to Sulawesi (e.g., *Palaquium* spp., rattan *Calamus* spp.) — North Sulawesi and Central Sulawesi have rattan collection quotas; Togean Islands protected area has collection ban |
| **Maluku** | Sasi customary law | *Sasi* is a Malukan customary resource management system with legal recognition; sasi-closed areas prohibit harvest of designated plant species (sea grasses, mangroves, sago) during seasonal closures; provincial government codifies sasi in regional regulations |

## Customary Law (*Hukum Adat*) and *Masyarakat Hukum Adat*

Indonesia's Constitutional Court rulings (Putusan MK 35/2012) and **UU 41/1999 on Forestry** (as amended) recognise customary forests (*Hutan Adat*) as distinct from state forests. Communities with recognised *masyarakat hukum adat* status control plant resource access within their customary territories. Over 5.4 million hectares of customary forests have been formally recognised as of 2024 — in these areas, national species permit requirements coexist with customary authority.

<div class="alert alert-info" role="alert">
**Source note** — [KLHK KSDAE (Directorate of Biodiversity Conservation)](https://ksdae.menlhk.go.id) | Leuser Ecosystem Authority: [leuserdevelopment.com](https://leuserdevelopment.com) | Papua Special Autonomy: [BIG Papua](https://big.go.id) | Sasi: [IUCN — Customary Resource Management in Maluku](https://www.iucn.org).
</div>

---

# Regulatory Institutions {#institutions}

| Abbrev. | Institution | Mandate |
|---------|-------------|---------|
| KLHK | Ministry of Environment and Forestry | Primary environmental regulator; protected species; CITES Management Authority |
| BKSDA | Natural Resources Conservation Agency (regional) | Protected area management; species permit enforcement |
| LIPI / BRIN | National Research and Innovation Agency | Biodiversity science; species assessments; Herbarium Bogoriense |
| BKSDA | Conservation agencies (per province) | Regional protected species enforcement; wildlife crime |

---

# Plant Species Listing {#species}

```{r master-table}
species |>
  arrange(iucn_status, common_name) |>
  mutate(`Scientific Name` = paste0("*", scientific_name, "*")) |>
  select(`Common Name`=common_name, `Scientific Name`, `IUCN`=iucn_status, `CITES`=cites_appendix,
         `Indonesia Status`=legal_status_id, `Managing Inst.`=managing_institution, `Notes`=notes) |>
  styled_kable(caption="Table 1. Protected Plant Species in Indonesia", height="450px")
```

## GBIF Backbone Resolution

```{r gbif-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    mutate(gbif_url=if_else(!is.na(GBIFusageKey)&GBIFusageKey!="",glue("https://www.gbif.org/species/{GBIFusageKey}"),NA_character_),
           iucn_url=paste0("https://www.iucnredlist.org/search?query=",URLencode(str_squish(coalesce(prefName,"")),reserved=TRUE)),
           `Scientific Name`=if_else(!is.na(gbif_url),glue("<a href='{gbif_url}' target='_blank'><em>{prefName}</em></a>"),glue("<em>{coalesce(prefName,'')}</em>")),
           `IUCN`=glue("<a href='{iucn_url}' target='_blank'>IUCN &#x2197;</a>")) |>
    transmute(`#`=row_number(),`Scientific Name`,`Common Name`=coalesce(englishName,""),`Family`=coalesce(family,""),
              `GBIF Key`=coalesce(GBIFusageKey,""),`Indonesia Status`=coalesce(legal_status_id,""),
              `Synonyms`=str_trunc(coalesce(synonyms,""),80),`IUCN`) |>
    styled_kable(caption=glue("Table 2. Indonesia — GBIF Backbone Resolved ({n_gbif_resolved} of {n_species} | {n_gbif_families} families)"), height="450px")
} else { cat("*GBIF data will be generated on first render with internet access.*") }
```

---

# Complete Species Table (Template Format) {#complete-template}

The table below presents all species in the **18-column `regulated_plants_v9.7` template format** — the canonical client deliverable schema. This is the same data as the downloadable CSV.

```{r complete-template-table}
if (!is.null(plants_gbif) && nrow(plants_gbif) > 0) {
  plants_gbif |>
    styled_kable(
      caption = "Complete Species Listing — 18-Column Template Format (GBIFusageKey, country, region, jurisdiction, jurisdiction_group, prefName, classification, taxonLevel, family, englishName, synonyms, Note, genus_I, is_aquatic, aquatic_type_I, legal_status, restriction_type, governing_law)",
      height = "650px"
    )
}
```

---

# Regulatory Gaps {#gaps}

| # | Gap Area | Severity | Status |
|---|----------|----------|--------|
| 1 | GR 7/1999 species list outdated; GR 106/2018 update partial | High | GR 106/2018 issued |
| 2 | Deforestation rate: one of highest globally; habitat loss | High | Moratorium active |
| 3 | Online plant trade via social media (orchids, Rafflesia) | High | Partial enforcement |
| 4 | Herbarium Bogoriense — digitisation lags field diversity | Medium | GBIF publishing active |
| 5 | IAS framework weak; no standalone invasive species law | Medium | Draft regulations |

---

# References {#references}

| Resource | Link |
|----------|------|
| Law No. 5/1990 — English translation | [FLEVIN](https://www.flevin.com/id/lgso/translations/JICA%20Mirror/english/4.05.1990.eng.qc.html) |
| GR No. 7/1999 — English translation | [UNODC](https://www.unodc.org/cld/uploads/res/document/regulation-number-7-year-1999-on-preserving-flora-and-fauna-species-english_html/Regulation_1999_EN.pdf) |
| KLHK | [menlhk.go.id](https://www.menlhk.go.id) |
| Herbarium Bogoriense | [herbarium.lipi.go.id](http://herbarium.lipi.go.id) |
| GBIF Indonesia | [gbif.org/country/ID](https://www.gbif.org/country/ID/summary) |

---

```{r session-info}
sessionInfo()
```

```{r kqb-footer, results='asis', echo=FALSE}
logo_sm <- kqb_logo_img(height="28px")
cat(paste0('<div class="kqb-report-footer">',if(nzchar(logo_sm)) paste0(logo_sm,'<br/>') else "",
  '<p style="font-weight:600;color:#0d2240;">',kqb_co$name,' &mdash; ',kqb_co$tagline,'</p>',
  '<p>',kqb_co$legal_name,' | ',kqb_co$country,'</p>',
  '<p style="margin-top:6px;font-size:10px;color:#aaa;">Generated ',format(Sys.Date(),"%d %B %Y"),
  ' &nbsp;&bull;&nbsp; Verify at KLHK / CITES</p></div>'))
```
